### 2020-04-29 README.Rmd
render README.md using github actions

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```

# gratia

<!-- badges: start -->
[![R build status](https://github.com/gavinsimpson/gratia/workflows/R-CMD-check/badge.svg)](https://github.com/gavinsimpson/gratia/actions)
<!-- badges: end -->
[![Build Status](https://travis-ci.org/gavinsimpson/gratia.svg?branch=master)](https://travis-ci.org/gavinsimpson/gratia)
[![Build status](https://ci.appveyor.com/api/projects/status/w7pj8773t5b8fxkb/branch/master?svg=true)](https://ci.appveyor.com/project/gavinsimpson/gratia/branch/master)
[![codecov.io](https://codecov.io/gh/gavinsimpson/gratia/branch/master/graph/badge.svg)](https://codecov.io/gh/gavinsimpson/gratia)
[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/gratia)](https://cran.r-project.org/package=gratia)
<!-- badges: end -->

## Overview

Graceful 'ggplot'-based graphics and utility functions for working with generalized additive models (GAMs) fitted using the 'mgcv' package. Provides a reimplementation of the `plot()` method for GAMs that 'mgcv' provides, as well as 'tidyverse' compatible representations of estimated smooths.

## Features

```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 7}
library('mgcv')
library('gratia')
df <- gamSim(eg = 1, n = 400, verbose = FALSE)
m <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, model = "REML")
draw(m)
```

```{r apprise-figure, dev = "png", fig.height = 6, fig.width = 7}
appraise(m)
```

The main features of *gratia* are currently

* A *ggplot2*-based replacement for `mgcv:::plot.gam()`: `draw(gam_model)`.

    ![Estimated smooths from a GAM](man/figures/README-draw-gam-figure.png)

    Note specialist smoothers (`bs %in% c("sos","mrf","so")`) are not currently supported, but univariate, *factor* and *continuous* `by`-variable smooths, simple random effect smooths (`bs = 're'`), factor-smooth interaction smooths (`bs = "fs"`), and bivariate tensor product smooths are supported,

* Estimatation of derivatives of fitted smoothers: `derivatives()` (older functionality in `fderiv(gam_model)`),

* Estimation of point-wise across-the-function confidence intervals and simultaneous intervals for smooths: `confint(gam_model)`.

* Model diagnostics via `appraise()`

    ![Model diagnostics figure](man/figures/README-appraise-figure.png)

## Installing *gratia*

*gratia* is now available on CRAN, and can be installed with

```{r install, eval = FALSE}
install.packages('gratia')
```

however *gratia* is under active development and you may wish to install the development version from github. The easiest way to do this is via the `install_github()` function from package *remotes*. Make sure you have *remotes* installed, then run

```{r install-github, eval = FALSE}
remotes::install_github("gavinsimpson/gratia")
```

to install the package.

## History

*gratia* grew out of an earlier package, *schoenberg*, itself a development of the earlier package *tsgam*, which was originally intended to be used with GAMs fitted to time series. As I was developing *tsgam* however it became clear that the package could be used more generally and that the name "tsgam" was no longer appropriate. To avoid breaking blog posts I had written using *tsgam* I decided to copy the git repo and all the history to a new repo for the package under the name *schoenberg*. At a later date someone released another package called *schoenberg* to CRAN, so that scuppered that idea. Now I'm calling the package *gratia*. Hopefully I won't have to change it again…

## Why *gratia*?

In naming his [*greta*](https://github.com/greta-dev/greta) package, Nick Golding observed the recent phenomena of naming statistical modelling software, such as Stan or Edward, after individuals that played a prominent role in the development of the field. This lead Nick to name his Tensor Flow-based package *greta* after [*Grete Hermann*](https://greta-stats.org/articles/webpages/why_greta.html).

In the same spirit, *gratia* is named in recognition of the contributions of [Grace Wahba](https://en.wikipedia.org/wiki/Grace_Wahba), who did pioneering work on the penalised spline models that are at the foundation of the way GAMs are estimated in *mgcv*. I wanted to name the package *grace*, to explicitly recognise Grace's contributions, but unfortunately there was already a package named *Grace* on CRAN. So I looked elsewhere for inspiration.

The English word "grace" derives from the Latin *gratia*, meaning "favor, charm, thanks" ([according to Merriam Webster](https://www.merriam-webster.com/dictionary/grace)).

The chair that Grace Wabha currently holds is named after [Isaac J Schoenberg](https://en.wikipedia.org/wiki/Isaac_Jacob_Schoenberg), a former University Madison-Wisconsin Professor of Mathematics, who in a 1946 paper provided the first mathematical reference to "splines". (Hence the previous name for the package.)

### 2020-04-30 README.Rmd
tweaking; fix a typo in referenceing a figure; render only to markdown

@@
 
-```{r apprise-figure, dev = "png", fig.height = 6, fig.width = 7}
+```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 7}
 appraise(m)
@@
 
-    ![Estimated smooths from a GAM](man/figures/README-draw-gam-figure.png)
+    ![Estimated smooths from a GAM](man/figures/README-draw-gam-figure-1.png)
 
@@
 
-    ![Model diagnostics figure](man/figures/README-appraise-figure.png)
+    ![Model diagnostics figure](man/figures/README-appraise-figure-1.png)
 

### 2020-04-30 README.Rmd
don't render the initial plot chunks

@@
 
-```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 7}
+```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 7, echo = FALSE, fig.show = "hide"}
 library('mgcv')
@@
 
-```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 7}
+```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 7, echo = FALSE, fig.show = "hide"}
 appraise(m)

### 2020-04-30 README.Rmd
don't render messages in chunks

@@
 
-```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 7, echo = FALSE, fig.show = "hide"}
+```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 7, echo = FALSE, fig.show = "hide", message = FALSE}
 library('mgcv')

### 2020-04-30 README.Rmd
remove duplicate badges end comment

@@
 [![R build status](https://github.com/gavinsimpson/gratia/workflows/R-CMD-check/badge.svg)](https://github.com/gavinsimpson/gratia/actions)
-<!-- badges: end -->
 [![Build Status](https://travis-ci.org/gavinsimpson/gratia.svg?branch=master)](https://travis-ci.org/gavinsimpson/gratia)

### 2020-04-30 README.Rmd
move package loading to set-up chunk

@@
 options(tibble.print_min = 5, tibble.print_max = 5)
+library('mgcv')
+library('gratia')
 ```
@@
 ```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 7, echo = FALSE, fig.show = "hide", message = FALSE}
-library('mgcv')
-library('gratia')
 df <- gamSim(eg = 1, n = 400, verbose = FALSE)

### 2020-04-30 README.Rmd
make the figures a bit wider

@@
 
-```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 7, echo = FALSE, fig.show = "hide", message = FALSE}
+```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 8, echo = FALSE, fig.show = "hide", message = FALSE}
 df <- gamSim(eg = 1, n = 400, verbose = FALSE)
@@
 
-```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 7, echo = FALSE, fig.show = "hide"}
+```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 8, echo = FALSE, fig.show = "hide"}
 appraise(m)

### 2020-04-30 README.Rmd
a bit wider still

@@
 
-```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 8, echo = FALSE, fig.show = "hide", message = FALSE}
+```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 9, echo = FALSE, fig.show = "hide", message = FALSE}
 df <- gamSim(eg = 1, n = 400, verbose = FALSE)
@@
 
-```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 8, echo = FALSE, fig.show = "hide"}
+```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 9, echo = FALSE, fig.show = "hide"}
 appraise(m)

### 2020-04-30 README.Rmd
tweaking figures again

@@
 
-```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 9, echo = FALSE, fig.show = "hide", message = FALSE}
+```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide", message = FALSE}
 df <- gamSim(eg = 1, n = 400, verbose = FALSE)
@@
 
-```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 9, echo = FALSE, fig.show = "hide"}
+```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide"}
 appraise(m)

### 2020-04-30 README.Rmd
access r-pkg.org badge with https

@@
 [![codecov.io](https://codecov.io/gh/gavinsimpson/gratia/branch/master/graph/badge.svg)](https://codecov.io/gh/gavinsimpson/gratia)
-[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/gratia)](https://cran.r-project.org/package=gratia)
+[![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/gratia)](https://cran.r-project.org/package=gratia)
 <!-- badges: end -->

### 2020-04-30 README.Rmd
set a seed for the gam simulated data

@@
 ```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide", message = FALSE}
+set.seed(42)
 df <- gamSim(eg = 1, n = 400, verbose = FALSE)
@@
 
-```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide"}
+```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide", dependson = -1}
 appraise(m)

### 2020-04-30 README.Rmd
use method argument not model FFS

@@
 df <- gamSim(eg = 1, n = 400, verbose = FALSE)
-m <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, model = "REML")
+m <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, method = "REML")
 draw(m)

### 2020-05-18 README.Rmd
add an extra example to the readme

--- 2020-04-30 README.Rmd
@@
 ```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide", message = FALSE}
-set.seed(42)
-df <- gamSim(eg = 1, n = 400, verbose = FALSE)
-m <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, model = "REML")
-draw(m)
+df1 <- data_sim("eg1", n = 400, seed = 42)
+m <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df1, method = "REML")
+draw(m, partial_residuals = TRUE)
+```
+
+```{r draw-gam-figure-2d, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide", message = FALSE}
+df2 <- data_sim("eg2", n = 1000, seed = 42)
+m2 <- gam(y ~ s(x, z),  data = df2, method = "REML")
+draw(m2)
 ```
@@
 
+    For example, the classic four term additive example from Gu & Wabha:
+
 	![Estimated smooths from a GAM](man/figures/README-draw-gam-figure-1.png)
+
+    Or for a bivariate smooth:
+
+    ![Estimated smooths from a GAM](man/figures/README-draw-gam-figure-2d-1.png)
 

### 2020-05-18 README.Rmd
update readme typo/bug

@@
 df1 <- data_sim("eg1", n = 400, seed = 42)
-m <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df1, method = "REML")
-draw(m, partial_residuals = TRUE)
+m1 <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df1, method = "REML")
+draw(m1, residuals = TRUE)
 ```

### 2020-05-18 README.Rmd
fix another typo
fix typo: Wabha -> Wahba

@@
 ```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide", dependson = -1}
-appraise(m)
+appraise(m1)
 ```
@@
 
-    For example, the classic four term additive example from Gu & Wabha:
+    For example, the classic four term additive example from Gu & Wahba:
 

### 2021-01-23 README.Rmd
remove the travis badge from readme

@@
 [![R build status](https://github.com/gavinsimpson/gratia/workflows/R-CMD-check/badge.svg)](https://github.com/gavinsimpson/gratia/actions)
-[![Build Status](https://travis-ci.org/gavinsimpson/gratia.svg?branch=master)](https://travis-ci.org/gavinsimpson/gratia)
 [![Build status](https://ci.appveyor.com/api/projects/status/w7pj8773t5b8fxkb/branch/master?svg=true)](https://ci.appveyor.com/project/gavinsimpson/gratia/branch/master)

### 2022-01-28 README.Rmd
add a cranlogs download badge

@@
 [![R build status](https://github.com/gavinsimpson/gratia/workflows/R-CMD-check/badge.svg)](https://github.com/gavinsimpson/gratia/actions)
-[![Build status](https://ci.appveyor.com/api/projects/status/w7pj8773t5b8fxkb/branch/master?svg=true)](https://ci.appveyor.com/project/gavinsimpson/gratia/branch/master)
 [![codecov.io](https://codecov.io/gh/gavinsimpson/gratia/branch/master/graph/badge.svg)](https://codecov.io/gh/gavinsimpson/gratia)
 [![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/gratia)](https://cran.r-project.org/package=gratia)
+[![](http://cranlogs.r-pkg.org/badges/grand-total/gratia)](http://cran.rstudio.com/web/packages/gratia/index.html)
 <!-- badges: end -->

### 2022-02-06 README.Rmd
tweak for readme

@@
 [![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/gratia)](https://cran.r-project.org/package=gratia)
-[![](http://cranlogs.r-pkg.org/badges/grand-total/gratia)](http://cran.rstudio.com/web/packages/gratia/index.html)
+[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/grand-total/gratia)](http://cran.rstudio.com/web/packages/gratia/index.html)
 <!-- badges: end -->
@@
 
-```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide", dependson = -1}
+```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide", dependson = -2}
 appraise(m1)

### 2022-02-06 README.Rmd
Deploying to gh-pages from @ gavinsimpson/gratia@4adf7ccf520acdb6071d6e4d546d11a6c7050e2f 🚀

@@
 
-Graceful 'ggplot'-based graphics and utility functions for working with generalized additive models (GAMs) fitted using the 'mgcv' package.
-Provides a reimplementation of the `plot()` method for GAMs that 'mgcv' provides, as well as 'tidyverse' compatible representations of estimated smooths.
+Graceful 'ggplot'-based graphics and utility functions for working with
+generalized additive models (GAMs) fitted using the 'mgcv' package. Provides a
+reimplementation of the `plot()` method for GAMs that 'mgcv' provides, as well
+as 'tidyverse' compatible representations of estimated smooths.
 
@@
 
-* A *ggplot2*-based replacement for `mgcv:::plot.gam()`: `draw(gam_model)`.
+* A *ggplot2*-based replacement for `mgcv:::plot.gam()`: `draw.gam()`.
 
@@
 
-    Note specialist smoothers (`bs %in% c("sos","mrf","so")`) are not currently supported, but univariate, *factor* and *continuous* `by`-variable smooths, simple random effect smooths (`bs = 're'`), factor-smooth interaction smooths (`bs = "fs"`), and bivariate tensor product smooths are supported,
+    Note specialist smoothers (`bs %in% c("sos","mrf","so")`) are not currently
+    supported, but univariate, *factor* and *continuous* `by`-variable smooths,
+    simple random effect smooths (`bs = 're'`), factor-smooth interaction
+    smooths (`bs = "fs"`), and bivariate tensor product smooths are supported,
 
-* Estimatation of derivatives of fitted smoothers: `derivatives()` (older functionality in `fderiv(gam_model)`),
+* Estimatation of derivatives of fitted smoothers: `derivatives()`,
 
-* Estimation of point-wise across-the-function confidence intervals and simultaneous intervals for smooths: `confint(gam_model)`.
+* Estimation of point-wise across-the-function confidence intervals and
+  simultaneous intervals for smooths: `confint.gam()`.
 
@@
 
-however *gratia* is under active development and you may wish to install the development version from github.
-The easiest way to do this is via the `install_github()` function from package *remotes*.
-Make sure you have *remotes* installed, then run
+however *gratia* is under active development and you may wish to install the
+development version from github. The easiest way to do this is via the
+`install_github()` function from package *remotes*. Make sure you have
+*remotes* installed, then run
 
@@
 
-*gratia* grew out of an earlier package, *schoenberg*, itself a development of the earlier package *tsgam*, which was originally intended to be used with GAMs fitted to time series.
-As I was developing *tsgam* however it became clear that the package could be used more generally and that the name "tsgam" was no longer appropriate.
-To avoid breaking blog posts I had written using *tsgam* I decided to copy the git repo and all the history to a new repo for the package under the name *schoenberg*.
-At a later date someone released another package called *schoenberg* to CRAN, so that scuppered that idea.
-Now I'm calling the package *gratia*.
-Hopefully I won't have to change it again…
+*gratia* grew out of an earlier package, *schoenberg*, itself a development of
+the earlier package *tsgam*, which was originally intended to be used with GAMs
+fitted to time series. As I was developing *tsgam* however it became clear that
+the package could be used more generally and that the name "tsgam" was no longer
+appropriate. To avoid breaking blog posts I had written using *tsgam* I decided
+to copy the git repo and all the history to a new repo for the package under the
+name *schoenberg*. At a later date someone released another package called
+*schoenberg* to CRAN, so that scuppered that idea. Now I'm calling the package
+*gratia*. Hopefully I won't have to change it again&hellip;
 
@@
 
-In naming his [*greta*](https://github.com/greta-dev/greta) package, Nick Golding observed the recent phenomena of naming statistical modelling software, such as Stan or Edward, after individuals that played a prominent role in the development of the field.
-This lead Nick to name his Tensor Flow-based package *greta* after [*Grete Hermann*](https://greta-stats.org/articles/webpages/why_greta.html).
+In naming his [*greta*](https://github.com/greta-dev/greta) package, Nick
+Golding observed the recent phenomena of naming statistical modelling software,
+such as Stan or Edward, after individuals that played a prominent role in the
+development of the field. This lead Nick to name his Tensor Flow-based package
+*greta* after [*Grete Hermann*](https://greta-stats.org/articles/webpages/why_greta.html).
 
-In the same spirit, *gratia* is named in recognition of the contributions of [Grace Wahba](https://en.wikipedia.org/wiki/Grace_Wahba), who did pioneering work on the penalised spline models that are at the foundation of the way GAMs are estimated in *mgcv*.
-I wanted to name the package *grace*, to explicitly recognise Grace's contributions, but unfortunately there was already a package named *Grace* on CRAN.
-So I looked elsewhere for inspiration.
+In the same spirit, *gratia* is named in recognition of the contributions of
+[Grace Wahba](https://en.wikipedia.org/wiki/Grace_Wahba), who did pioneering
+work on the penalised spline models that are at the foundation of the way GAMs
+are estimated in *mgcv*. I wanted to name the package *grace*, to explicitly
+recognise Grace's contributions, but unfortunately there was already a package
+named *Grace* on CRAN. So I looked elsewhere for inspiration.
 
-The English word "grace" derives from the Latin *gratia*, meaning "favor, charm, thanks" ([according to Merriam Webster](https://www.merriam-webster.com/dictionary/grace)).
+The English word "grace" derives from the Latin *gratia*, meaning "favor, charm,
+thanks" ([according to Merriam Webster](https://www.merriam-webster.com/dictionary/grace)).
 
-The chair that Grace Wabha currently holds is named after [Isaac J Schoenberg](https://en.wikipedia.org/wiki/Isaac_Jacob_Schoenberg), a former University Madison-Wisconsin Professor of Mathematics, who in a 1946 paper provided the first mathematical reference to "splines".
-(Hence the previous name for the package.)
+The chair that Grace Wabha currently holds is named after 
+[Isaac J Schoenberg](https://en.wikipedia.org/wiki/Isaac_Jacob_Schoenberg), a
+former University Madison-Wisconsin Professor of Mathematics, who in a 1946
+paper provided the first mathematical reference to "splines". (Hence the
+previous name for the package.)

### 2022-02-07 README.Rmd
Deploying to gh-pages from @ gavinsimpson/gratia@d1446126f1bdb4afc6a55ce3f506382e1be54587 🚀

@@
 [![R build status](https://github.com/gavinsimpson/gratia/workflows/R-CMD-check/badge.svg)](https://github.com/gavinsimpson/gratia/actions)
-[![codecov.io](https://codecov.io/gh/gavinsimpson/gratia/branch/master/graph/badge.svg)](https://codecov.io/gh/gavinsimpson/gratia)
+[![codecov.io](https://codecov.io/gh/gavinsimpson/gratia/branch/master/graph/badge.svg)](https://app.codecov.io/gh/gavinsimpson/gratia)
 [![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/gratia)](https://cran.r-project.org/package=gratia)
-[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/grand-total/gratia)](http://cran.rstudio.com/web/packages/gratia/index.html)
+[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/grand-total/gratia)](https://cran.r-project.org/package=gratia)
 <!-- badges: end -->

### 2022-03-18 README.Rmd
Deploying to gh-pages from @ gavinsimpson/gratia@85e85e71ebd43d4d43d639a836f611144a3153fc 🚀

@@
 
-    Note specialist smoothers (`bs %in% c("sos","mrf","so")`) are not currently
+    Note specialist smoothers (`bs %in% c("mrf","so")`) are not currently
     supported, but univariate, *factor* and *continuous* `by`-variable smooths,

### 2022-12-08 README.Rmd
add instructions for installing from the r universe

@@
 
-to install the package.
+to install the package. Alternatively, binary packages of the development
+version are available from rOpenSci's R Universe service:
+
+```{r install-r-universe, eval = FALSE}
+# Enable repository from gavinsimpson
+options(repos = c(
+  gavinsimpson = "https://gavinsimpson.r-universe.dev",
+  CRAN = "https://cloud.r-project.org"))
+
+# Download and install gratia in R
+install.packages("gratia")
+```
 

### 2023-02-16 README.Rmd
Deploying to gh-pages from @ gavinsimpson/gratia@7a0663c0f295a03399c016200f2c59078042ad66 🚀

@@
 [![R build status](https://github.com/gavinsimpson/gratia/workflows/R-CMD-check/badge.svg)](https://github.com/gavinsimpson/gratia/actions)
-[![codecov.io](https://codecov.io/gh/gavinsimpson/gratia/branch/master/graph/badge.svg)](https://app.codecov.io/gh/gavinsimpson/gratia)
+[![codecov](https://codecov.io/gh/gavinsimpson/gratia/branch/main/graph/badge.svg?token=GG5NQfgRFu)](https://codecov.io/gh/gavinsimpson/gratia)
 [![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/gratia)](https://cran.r-project.org/package=gratia)

### 2023-02-18 R/basis-funs.R
Deploying to gh-pages from @ gavinsimpson/gratia@a132e0ead3ad3d49a87f59552efbc65fc6b61e7a 🚀

@@ -191,8 +191,8 @@
     }
 
     if (is.null(data)) {
-        data <- smooth_data(model, ids[i],
-            n = n, n_2d = n_2d, n_3d = n_3d, n_4d = n_4d, offset = offset)
+        data <- smooth_data(model, ids[i], n = n, n_3d = n_3d, n_4d = n_4d,
+            offset = offset)
     }
 
     tbl <- tidy_basis(smooths[[i]], at = data, coefs = betas,
@@ -216,7 +216,7 @@
     }
     ## call smoothCon to create the basis as specified in `x`
     sm <- smoothCon(object, data = data, knots = knots,
-        absorb.cons = constraints)
+        absorb.cons = constraints, ...)
 
     ## sm will be a list, even if a single smooth, bc we could have multiple
     ## smoothers in case of factor `by` smooths.

### 2023-02-18 R/confint-methods.R

@@ -138,8 +138,7 @@
     ## bayesian covar matrix, possibly accounting for estimating smooth pars
     Vb <- vcov(x[["model"]], unconditional = x$unconditional)
     ## simulate un-biased deviations given bayesian covar matrix
-    buDiff <- mvnfast::rmvn(n = nsim, mu = rep(0, nrow(Vb)), sigma = Vb,
-        ncores = ncores)
+    buDiff <- rmvn(n = nsim, mu = rep(0, nrow(Vb)), sigma = Vb, ncores = ncores)
     ## apply wrapper to compute simultaneous interval critical value and
     ## corresponding simultaneous interval for each term
     res <- lapply(x[["derivatives"]][terms], FUN = simInt,

### 2023-02-18 R/data-sim.R

@@ -13,29 +13,21 @@
 #' @param theta numeric; the dispersion parameter \eqn{\theta} to use. The
 #'   default is entirely arbitrary, chosen only to provide simulated data that
 #'   exhibits extra dispersion beyond that assumed by under a Poisson.
-#' @param n_cat integer; the number of categories for categorical response.
-#'   Currently only used for `distr %in% c("ocat", "ordered categorical")`.
-#' @param cuts numeric; vector of cut points on the latent variable, excluding
-#'   the end points `-Inf` and `Inf`. Must be one fewer than the number of
-#'   categories: `length(cuts) == n_cat - 1`.
 #' @param seed numeric; the seed for the random number generator. Passed to
 #'   [base::set.seed()].
 #'
 #' @export
 #'
 #' @examples
-#' \dontshow{op <- options(pillar.sigfig = 5, cli.unicode = FALSE)
+#' \dontshow{
+#' set.seed(1)
+#' op <- options(pillar.sigfig = 5, cli.unicode = FALSE)
 #' }
-#' data_sim("eg1", n = 100, seed = 1)
-#'
-#' # an ordered categorical response
-#' data_sim("eg1", n = 100, dist = "ocat", n_cat = 4, cuts = c(-1, 0, 5))
+#' data_sim("eg1")
 #' \dontshow{options(op)}
 `data_sim` <- function(model = "eg1", n = 400, scale = 2, theta = 3,
                        dist = c("normal", "poisson", "binary",
-                                "negbin", "tweedie",
-                                "ocat", "ordered categorical"),
-                       n_cat = 4, cuts = c(-1, 0, 5),
+                                "negbin", "tweedie"),
                        seed = NULL) {
     ## sort out the seed
     if (!exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)) {
@@ -53,18 +45,13 @@
 
     ## check dist is OK
     dist <- match.arg(dist)
-    special_dist <- c("ordered categorical" = "ocat")
-    if (dist %in% names(special_dist)) {
-        dist <- unname(special_dist[dist])
-    }
 
     sim_fun <- switch(dist,
-        normal  = sim_normal,
-        poisson = sim_poisson,
-        binary  = sim_binary,
-        negbin  = sim_nb,
-        tweedie = sim_tweedie,
-        ocat = sim_normal)
+                      normal  = sim_normal,
[…]

### 2023-02-18 R/data-slice.R

@@ -16,39 +16,11 @@
          ">", call. = FALSE)
 }
 
-#' @export
-#' @rdname data_slice
-#' @importFrom tidyr expand_grid
-#' @importFrom rlang enquos eval_tidy
-`data_slice.data.frame` <- function(object, ...) {
-    # deal with ...
-    exprs <- rlang::enquos(...)
-    slice_vars <- purrr::map(exprs, rlang::eval_tidy, data = object)
-
-    # check now if there are elements of slice_vars that aren't in the object
-    vars <- names(object)
-    nms <- names(slice_vars)
-    if (any(i <- ! nms %in% vars)) {
-        message("Some specified variable(s) not used in `object``:\n",
-                paste(" * ", nms[i], collapse = "\n", sep = ""),
-                "\n")
-    }
-
-    # typical values, only needed ones that aren't
-    need_tv <- setdiff(vars, names(slice_vars))
-    if (length(need_tv) > 0L) {
-        tv <- typical_values(object)
-        slice_vars <- append(slice_vars, tv[need_tv])
-    }
-
-    expand_grid(!!!{slice_vars})
-}
-
 #' @param data an alternative data frame of values containing all the variables
 #'   needed to fit the model. If `NULL`, the default, the data used to fit the
 #'   model will be recovered using `model.frame`. User-supplied expressions
 #'   passed in `...` will be evaluated in `data`.
-#'
+#' 
 #' @export
 #' @rdname data_slice
 #' @importFrom tidyr expand_grid
@@ -305,24 +277,6 @@
 
     # return
     as_tibble(summ)
-}
-
-#' @export
-#' @rdname typical_values
-#' @importFrom tidyselect everything
-#' @importFrom dplyr summarise across
-#' @importFrom tibble as_tibble
-`typical_values.data.frame` <- function(object, vars = everything(), ...) {
-    # include/exclude any terms?
-    expr <- rlang::enquo(vars)
-    pos <- eval_select(expr, data = object)
-    object <- object[pos]
-
-    df <- object |>
-        summarise(across(everything(), .fns = value_closest_to_median))
-
-    # return
-    as_tibble(df)
 }
 
 #' All combinations of factor levels

### 2023-02-18 R/datasets.R

@@ -90,7 +90,7 @@ NULL
 #'   * `nativeBorn`: factor; Was the respondent born in the US? A factor with
 #'     levels `no` and `yes`.
 #'   * `ageGroup`: factor; grouped age of the respondent with levels `18-29`
-#'       `30-39`, `40-49`, `50-59`, and `60+`.
+#'       `30-39`, `40-49`, `50-59`, and `60+`.##' 
 #'
 #' @keywords data
 #' @name gss_vocab
@@ -107,3 +107,4 @@ NULL
 #' @name ref_sims
 #' @docType data
 NULL
+

### 2023-02-18 R/draw-gam.R

@@ -354,8 +354,7 @@
                 upper = .data$partial + (crit * .data$se))
             # need to alter the ylim if scales are fixed
             if (isTRUE(identical(scales, "fixed"))) {
-                ylims <- range(ylims, object$partial, object$upper,
-                    object$lower)
+                ylims <- range(ylims, object$partial, object$upper, object$lower)
             }
 
             para_plts <- para %>%

### 2023-02-18 R/draw-methods.R

@@ -430,7 +430,6 @@
     plt
 }
 
-#' @importFrom ggplot2 scale_fill_distiller guides guide_axis facet_wrap vars labs geom_contour coord_equal
 `plot_bivariate_basis` <- function(object,
                                    legend = FALSE,
                                    labeller = NULL,
@@ -455,12 +454,12 @@
         fill = .data[["value"]],
         group = .data[["bf"]])) +
         geom_raster() +
-        guides(x = guide_axis(angle = angle, check.overlap = TRUE)) +
+        guides(x = guide_axis(angle = angle)) +
         scale_fill_distiller(palette = "RdBu", type = "div")
 
     ## default labels if none supplied
     if (is.null(xlab)) {
-        xlab <- smooth_var[1]
+        xlab <- smooth_var[2]
     }
     if (is.null(ylab)) {
         ylab <- smooth_var[2]
@@ -495,8 +494,7 @@
     }
 
     if (isTRUE(contour)) {
-        plt <- plt + geom_contour(mapping = aes(z = .data[["value"]],
-        group = .data[["bf"]], fill = NULL),
+        plt <- plt + geom_contour(mapping = aes(z = .data[["value"]]),
             colour = contour_col,
             bins = n_contour,
             na.rm = TRUE)

### 2023-02-18 R/family-utils.R

@@ -296,25 +296,37 @@ family_type.family <- function(object, ...) {
 ## Other internal functions ---------------------------------------------------
 
 ## Workhorse link extractor
-#' @importFrom dplyr case_when
 `get_link_function` <- function(object, parameter = "location",
                                 inverse = FALSE, which_eta = NULL) {
     inverse <- as.logical(inverse)
     linfo <- object[["linfo"]]
     distr <- object[["family"]] # name of the the family
 
-    ## process distr for some familiee
-    distr <- case_when(
-        grepl("^Negative Binomial", distr, ignore.case = TRUE) ~ "nb",
-        grepl("^Tweedie", distr, ignore.case = TRUE) ~ "tweedie",
-        grepl("^Beta regression", distr, ignore.case = TRUE) ~ "beta",
-        grepl("^scaled t", distr, ignore.case = TRUE) ~ "scaled_t",
-        grepl("^Ordered Categorical", distr, ignore.case = TRUE) ~ "ocat",
-        grepl("^zero inflated Poisson", distr, ignore.case = TRUE) ~ "zip",
-        grepl("^Cox PH", distr, ignore.case = TRUE) ~ "cox_ph",
-        grepl("^censored normal", distr, ignore.case = TRUE) ~ "cnorm",
-        .default = as.character(distr)
-    )
+    ## process distr for some families
+    if (grepl("^Negative Binomial", distr)) {
+        distr <- "nb"
+    }
+    if (grepl("negative binomial", distr)) {
+        distr <- "nb"
+    }
+    if (grepl("^Tweedie", distr)) {
+        distr <- "tweedie"
+    }
+    if (identical(distr, "Beta regression")) {
+        distr <- "beta"
+    }
+    if (grepl("^scaled t", distr, ignore.case = TRUE)) {
+        distr <- "scaled_t"
+    }
+    if (identical(distr, "Ordered Categorical")) {
+        distr <- "ocat"
+    }
+    if (identical(distr, "zero inflated Poisson")) {
+        distr <- "zip"
+    }
+    if (identical(distr, "Cox PH")) {
+        distr <- "cox_ph"
+    }
 
     ## which link function
     lfun <-
@@ -347,8 +359,7 @@ family_type.family <- function(object, ...) {
                             which_eta = which_eta),
              multinom = multinom_link(object, parameter, inverse = inverse,
                                       which_eta = which_eta),
-             shash = shash_link(object, parameter, inverse = inverse),
-             cnorm = cnorm_link(object, parameter, inverse = inverse)
[…]

### 2023-02-18 R/fitted_values.R

@@ -13,11 +13,9 @@
 #'   `newdata`, and `se.fit` are already used and passed on to
 #'   [mgcv::predict.gam()].
 #'
-#' @note For most families, regardless of the scale on which the fitted values
-#'   are returned, the `se` component of the returned object is on the *link*
-#'   (*linear predictor*) scale, not the response scale. An exception is the
-#'   `mgcv::ocat()` family, for which the `se` is on the response scale if
-#'   `scale = "response"`.
+#' @note Regardless of the scale on which the fitted values are returned, the
+#'   `se` component of the returned object is on the *link* (*linear predictor*)
+#'   scale, not the response scale.
 #'
 #' @return A tibble (data frame) whose first *m* columns contain either the data
 #'   used to fit the model (if `data` was `NULL`), or the variables supplied to
@@ -28,19 +26,13 @@
 #' * `lower`, `upper`: the limits of the credible interval on the fitted values,
 #'     on the specified scale.
 #'
-#' Models fitted with certain families will include additional variables
-#'
-#' * `mgcv::ocat()` models: when `scale = "repsonse"`, the returned object will
-#'   contain a `row` column and a `category` column, which indicate to which row
-#'   of the `data` each row of the returned object belongs. Additionally, there
-#'   will be `nrow(data) * n_categories` rows in the returned object; each row
-#'   is the predicted probability for a single category of the response.
-#'
 #' @export
 #'
 #' @examples
 #' load_mgcv()
-#' \dontshow{op <- options(cli.unicode = FALSE, pillar.sigfig = 6)}
+#' \dontshow{
+#' op <- options(cli.unicode = FALSE, pillar.sigfig = 6)
+#' }
 #' sim_df <- data_sim("eg1", n = 400, dist = "normal", scale = 2, seed = 2)
 #' m <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = sim_df, method = "REML")
 #' fv <- fitted_values(m)
@@ -52,6 +44,9 @@
 
 #' @export
 #' @rdname fitted_values
+#' @importFrom rlang set_names .data
+#' @importFrom dplyr bind_cols mutate across
+#' @importFrom tibble as_tibble is_tibble
 `fitted_values.gam` <- function(object,
                                 data = NULL,
                                 scale = c("response",
@@ -66,37 +61,22 @@
 
[…]

### 2023-02-18 R/lp-matrix.R

@@ -71,7 +71,7 @@
 #' @importFrom pillar tbl_sum tbl_format_header
 `print.lp_matrix` <- function(x, ..., n = 5, max_footer_lines = 1) {
     class(x) <- class(x)[-1]
-    tbl <- as_tibble(as.data.frame(x))
+    tbl <- as_tibble(x)
     class(tbl) <- append(class(tbl), "tbl_lp_matrix", after = 0)
     print(tbl, ..., n = n, max_footer_lines = max_footer_lines)
 }

### 2023-02-18 R/posterior-samples.R

@@ -6,6 +6,7 @@
 #'   the model will be used for `data`, if available in `model`.
 #' @param n numeric; the number of posterior samples to return.
 #' @param seed numeric; a random seed for the simulations.
+#' @param scale character;
 #' @param freq logical; `TRUE` to use the frequentist covariance matrix of
 #'   the parameter estimators, `FALSE` to use the Bayesian posterior
 #'   covariance matrix of the parameters.
@@ -14,37 +15,13 @@
 #'   used, if available.
 #' @param weights numeric; a vector of prior weights. If `data` is null
 #'   then defaults to `object[["prior.weights"]]`, otherwise a vector of ones.
-#' @param n_cores number of cores for generating random variables from a
+#' @param ncores number of cores for generating random variables from a
 #'   multivariate normal distribution. Passed to [mvnfast::rmvn()].
 #'   Parallelization will take place only if OpenMP is supported (but appears
 #'   to work on Windows with current `R`).
-#' @param method character; which method should be used to draw samples from
-#'   the posterior distribution. `"gaussian"` uses a Gaussian (Laplace)
-#'   approximation to the posterior. `"mh"` uses a Metropolis Hastings sample
-#'   that alternates t proposals with proposals based on a shrunken version of
-#'   the posterior covariance matrix. `"inla"` uses a variant of Integrated
-#'   Nested Laplace Approximation due to Wood (2019), (currently not
-#'   implemented). `"user"` allows for user-supplied posterior draws
-#'   (currently not implemented).
-#' @param burnin numeric; number of samples to discard as the burnin draws.
-#'   Only used with `method = "mh"`.
-#' @param thin numeric; the number of samples to skip when taking `n` draws.
-#'   Results in `thin * n` draws from the posterior being taken. Only used with
-#'   `method = "mh"`.
-#' @param t_df numeric; degrees of freedome for t distribution proposals. Only
-#'   used with `method = "mh"`.
-#' @param rw_scale numeric; Factor by which to scale posterior covariance
-#'   matrix when generating random walk proposals. Negative or non finite to
-#'   skip the random walk step. Only used with `method = "mh"`.
 #' @param ... arguments passed to other methods. For `fitted_samples()`, these
[…]

### 2023-02-18 R/smooth-data.R

@@ -3,8 +3,6 @@
 #' @param model a fitted model
 #' @param id the number ID of the smooth within `model` to process.
 #' @param n numeric; the number of new observations to generate.
-#' @param n_2d numeric; the number of new observations to generate for the
-#'   second dimension of a 2D smooth. *Currently ignored*.
 #' @param n_3d numeric; the number of new observations to generate for the third
 #'   dimension of a 3D smooth.
 #' @param n_4d numeric; the number of new observations to generate for the
@@ -41,8 +39,8 @@
 #' smooth_data(m, id = 2, include_all = TRUE)
 #'
 #' \dontshow{options(op)}
-`smooth_data` <- function(model, id, n = 100, n_2d = NULL, n_3d = NULL,
-                          n_4d = NULL, offset = NULL, include_all = FALSE) {
+`smooth_data` <- function(model, id, n = 100, n_3d = NULL, n_4d = NULL,
+                          offset = NULL, include_all = FALSE) {
     mf <- model.frame(model)           # model.frame used to fit model
 
     ## remove response

### 2023-02-18 R/smooth-estimates.R

@@ -1675,14 +1675,6 @@
                               ...) {
     # handle splines on the sphere
 
-    # this currently needs the mapproj pkg for coord_map()
-    if (!requireNamespace("mapproj", quietly = TRUE)) {
-        message("\nPlotting SOS smooths uses `ggplot2::coord_map()`.\n",
-            "This requires that the {mapproj} package be installed.\n",
-            "Run: `install.packages(\"mapproj\")`\n")
-        stop("Package {mapproj} is not available.")
-    }
-
     if (is.null(variables)) {
         variables <- vars_from_label(unique(object[["smooth"]]))
     }

### 2023-02-18 tests/testthat/setup.R

@@ -25,7 +25,7 @@ su_m_quick_eg1 <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3),
                       data = quick_eg1,
                       method = "REML")
 
-su_m_quick_eg1_shrink <- gam(y ~ s(x0, bs = "ts") + s(x1, bs = "cs") +
+su_m_quick_eg1_shrink <- gam(y ~ s(x0, bs = "ts") + s(x1, bs = "ts") +
                                s(x2, bs = "ts") + s(x3, bs = "ts"),
                              data = quick_eg1,
                              method = "REML")
@@ -39,19 +39,15 @@ su_m_penalty <- gam(y ~ s(x0, bs = 'cr') + s(x1, bs = 'bs') +
                     data = su_eg1,
                     method = "REML")
 
-su_m_bivar <- gam(y ~ s(x, z, k = 40), data = su_eg2, method = "REML")
-
-su_m_bivar_ds <- gam(y ~ s(x, z, k = 20, bs = "ds"), data = su_eg2,
-  method = "REML")
+su_m_bivar <- gam(y ~ s(x, z, k = 40),
+                  data = su_eg2,
+                  method = "REML")
 
 su_m_trivar <- gam(y ~ s(x0, x1, x2), data = su_eg1, method = "REML")
 
 su_m_quadvar <- gam(y ~ s(x0, x1, x2, x3), data = su_eg1, method = "REML")
 
 su_m_bivar_te <- gam(y ~ te(x, z, k = c(5, 5)), data = su_eg2, method = "REML")
-
-su_m_bivar_ti <- gam(y ~ s(x, k = 5) + s(z, k = 5) + ti(x, z, k = c(5, 5)),
-  data = su_eg2, method = "REML")
 
 su_m_bivar_t2 <- gam(y ~ t2(x, z, k = c(5, 5)), data = su_eg2, method = "REML")
 
@@ -70,7 +66,8 @@ su_m_quadvar_t2 <- gam(y ~ t2(x0, x1, x2, x3, k = c(3, 3, 3, 3)),
 su_m_cont_by <- gam(y ~ s(x2, by = x1), data = su_eg3, method = "REML")
 
 su_m_factor_by <- gam(y ~ fac + s(x2, by = fac) + s(x0),
-                      data = su_eg4, method = "REML")
+                      data = su_eg4,
+                      method = "REML")
 
[…]

### 2023-02-18 tests/testthat/test-basis.R

@@ -37,35 +37,3 @@ test_that("basis() works with a scam", {
     plt <- draw(bs)
     expect_doppelganger("draw basis works with a scam single smooth", plt)
 })
-
-test_that("basis() works with bivariate tprs smooths", {
-    expect_silent(ds <- data_slice(su_eg2, x = evenly(x, n = 20),
-        z = evenly(z, n = 20)))
-    expect_silent(bs <- basis(s(x, z, k = 25), data = su_eg2, at = ds,
-        constraints = TRUE))
-    expect_s3_class(bs, "basis")
-
-    plt <- draw(bs)
-    expect_doppelganger("draw basis works with a bivariate tprs", plt)
-
-    skip_on_cran()
-    skip_on_os("mac")
-    plt <- draw(bs, contour = TRUE)
-    expect_doppelganger("draw basis works with a bivariate tprs contour", plt)
-})
-
-test_that("basis() works with bivariate te smooths", {
-    expect_silent(ds <- data_slice(su_eg2, x = evenly(x, n = 20),
-        z = evenly(z, n = 20)))
-    expect_silent(bs <- basis(te(x, z, k = c(5, 5)), data = su_eg2, at = ds,
-        constraints = TRUE))
-    expect_s3_class(bs, "basis")
-
-    plt <- draw(bs)
-    expect_doppelganger("draw basis works with a bivariate te", plt)
-
-    skip_on_cran()
-    skip_on_os("mac")
-    plt <- draw(bs, contour = TRUE)
-    expect_doppelganger("draw basis works with a bivariate te contour", plt)
-})

### 2023-02-18 tests/testthat/test-data-slice.R

@@ -81,8 +81,8 @@ test_that("data_slice works for a GAM with factor by", {
 })
 
 test_that("default data_slice method fails gracefully", {
-    expect_error(data_slice(1:10),
-                 "Don't know how to create a data slice from <integer>",
+    expect_error(data_slice(su_eg4),
+                 "Don't know how to create a data slice from <tbl_df>",
                  fixed = TRUE)
 })
 
@@ -214,20 +214,4 @@ test_that("data_slice with works with models with an offset", {
     expect_identical(nrow(ds), 50L)
     expect_identical(ncol(ds), 2L)
     expect_identical(ds$off, rep(1, length.out = 50))
-})
-
-test_that("typical_values works for a data frame", {
-    expect_silent(tv <- typical_values(su_eg2, vars = any_of(c("x", "z"))))
-    expect_snapshot(tv, cran = FALSE)
-})
-
-test_that("data_slice works for a data frame", {
-    expect_silent(ds <- data_slice(su_eg2, x = evenly(x, n = 25)))
-    expect_snapshot(ds, cran = FALSE)
-    expect_identical(nrow(ds), 25L)
-
-    expect_silent(ds <- data_slice(su_eg2, x = evenly(x, n = 25),
-        z = evenly(z, n = 25)))
-    expect_snapshot(ds, cran = FALSE)
-    expect_identical(nrow(ds), 625L)
 })

### 2023-02-18 tests/testthat/test-draw-methods.R

@@ -6,6 +6,12 @@ library("gratia")
 library("mgcv")
 library("ggplot2")
 
+## Need a local wrapper to allow conditional use of vdiffr
+# `expect_doppelganger` <- function(title, fig, ...) {
+#   testthat::skip_if_not_installed("vdiffr")
+#   vdiffr::expect_doppelganger(title, fig, ...)
+# }
+
 test_that("draw.evaluated_1d_smooth() plots the smooth", {
     withr::local_options(lifecycle_verbosity = "quiet")
     sm <- evaluate_smooth(su_m_univar_4, "s(x2)")
@@ -341,31 +347,6 @@ test_that("draw.derivates() plots derivatives for a GAM rotated labels", {
         plt)
 })
 
-test_that("draw plots partial derivatives for a GAM", {
-    d1 <- partial_derivatives(su_m_bivar_te, term = "te(x,z)", focal = "z",
-        type = "central")
-    plt <- draw(d1)
-    expect_doppelganger("draw partial derivatives for a GAM", plt)
-
-    plt <- draw(d1, scales = "fixed")
-    ## skip_on_ci()
-    expect_doppelganger("draw partial derivatives for a GAM with fixed scales",
-        plt)
-})
-
-test_that("draw plots partial derivs for GAM rotated labels", {
-    skip_on_cran()
-    d1 <- partial_derivatives(su_m_bivar_te, term = "te(x,z)", focal = "z",
-        type = "central")
-    plt <- draw(d1, angle = 45)
-    expect_doppelganger("draw partial derivatives for GAM rotated labels",
-        plt)
-
-    plt <- draw(d1, scales = "fixed", angle = 45)
-    expect_doppelganger("draw partial derivatives for GAM fixed scales rotated",
-        plt)
-})
-
 ## test that issue 39 stays fixed
 test_that("draw.gam doesn't create empty plots with multiple parametric terms", {
     plt <- draw(m_2_fac, rug = FALSE)
@@ -572,12 +553,4 @@ test_that("draw.penalty_df works with normalization", {
     plt <- draw(penalty(su_m_univar_4, "s(x1)"), normalize = TRUE)
     expect_doppelganger("draw penalty_df with single smooths normalized",
                         plt)
-})
-
-test_that("plotting sos smooths works", {
-    skip_on_cran()
-    skip_if_not_installed("mapproj")
-    skip_on_os("mac")
-    expect_silent(plt <- draw(m_sos, n = 25))
-    expect_doppelganger("draw works for sos smooths", plt)
 })

### 2023-02-18 tests/testthat/test-evenly.R

@@ -9,27 +9,3 @@ test_that("evenly with by works", {
     expect_identical(length(x), 7L)
     expect_identical(seq(1987, 1999, by = 2), x)
 })
-
-test_that("seq min max with by works", {
-    expect_silent(x <- seq_min_max(1987:1999, by = 2))
-    expect_identical(length(x), 7L)
-    expect_identical(seq(1987, 1999, by = 2), x)
-})
-
-test_that("ref_level works", {
-    f <- factor(sample(letters[1:5], 100, replace = TRUE),
-        levels = letters[1:5])
-    expect_silent(rl <- ref_level(f))
-    expect_identical(factor("a", levels = letters[1:5]), rl)
-
-    expect_error(ref_level(1:10), "'fct' must be a factor")
-})
-
-test_that("ref_level works", {
-    f <- factor(sample(letters[1:5], 100, replace = TRUE),
-        levels = letters[1:5])
-    expect_silent(rl <- level(f, "b"))
-    expect_identical(factor("b", levels = letters[1:5]), rl)
-
-    expect_error(level(1:10), "'fct' must be a factor")
-})

### 2023-02-18 tests/testthat/test-family-utils.R

@@ -223,13 +223,6 @@ test_that("link() works for cox.ph() family objects", {
     expect_identical(f, cox.ph()$linkfun)
 })
 
-test_that("link() works for cnorm() family objects", {
-    f <- link(cnorm())
-    expect_type(f, "closure")
-    expect_identical(f(val), val)
-    expect_identical(f, cnorm()$linkfun)
-})
-
 ## inv_link
 test_that("inv_link() works for gaussian() family objects", {
     f <- inv_link(gaussian())
@@ -351,13 +344,6 @@ test_that("inv_link() works for cox.ph() family objects", {
     expect_type(f, "closure")
     expect_identical(f(val), cox.ph()$linkinv(val))
     expect_identical(f, cox.ph()$linkinv)
-})
-
-test_that("inv_link() works for cnorm() family objects", {
-    f <- inv_link(cnorm())
-    expect_type(f, "closure")
-    expect_identical(f(val), val)
-    expect_identical(f, cnorm()$linkinv)
 })
 
 test_that("extract_link() works on gaussian() family objects", {
@@ -1011,20 +997,6 @@ test_that("extract_link() works on shash() family objects", {
     expect_identical(f(val), fam$linfo[[4L]]$linkinv(val))
     expect_identical(f, fam$linfo[[4L]]$linkinv)
 })
-
-test_that("extract_link() works on cnorm() family objects", {
-    ## link
-    f <- extract_link(cnorm())
-    expect_type(f, "closure")
-    expect_identical(f(val), val)
-    expect_identical(f, cnorm()$linkfun)
-    ## inverse
-    f <- extract_link(cnorm(), inverse = TRUE)
-    expect_type(f, "closure")
-    expect_identical(f(val), val)
-    expect_identical(f, cnorm()$linkinv)
-})
-
 
 ## tests some specific extract functions
 test_that("twlss_link() can extract a link function", {

### 2023-02-18 R/family-utils.R
Deploying to gh-pages from @ gavinsimpson/gratia@74cb4be29dcad53b04ee48911436d29c7088eea3 🚀

@@ -296,37 +296,25 @@ family_type.family <- function(object, ...) {
 ## Other internal functions ---------------------------------------------------
 
 ## Workhorse link extractor
+#' @importFrom dplyr case_when
 `get_link_function` <- function(object, parameter = "location",
                                 inverse = FALSE, which_eta = NULL) {
     inverse <- as.logical(inverse)
     linfo <- object[["linfo"]]
     distr <- object[["family"]] # name of the the family
 
-    ## process distr for some families
-    if (grepl("^Negative Binomial", distr)) {
-        distr <- "nb"
-    }
-    if (grepl("negative binomial", distr)) {
-        distr <- "nb"
-    }
-    if (grepl("^Tweedie", distr)) {
-        distr <- "tweedie"
-    }
-    if (identical(distr, "Beta regression")) {
-        distr <- "beta"
-    }
-    if (grepl("^scaled t", distr, ignore.case = TRUE)) {
-        distr <- "scaled_t"
-    }
-    if (identical(distr, "Ordered Categorical")) {
-        distr <- "ocat"
-    }
-    if (identical(distr, "zero inflated Poisson")) {
-        distr <- "zip"
-    }
-    if (identical(distr, "Cox PH")) {
-        distr <- "cox_ph"
-    }
+    ## process distr for some familiee
+    distr <- case_when(
+        grepl("^Negative Binomial", distr, ignore.case = TRUE) ~ "nb",
+        grepl("^Tweedie", distr, ignore.case = TRUE) ~ "tweedie",
+        grepl("^Beta regression", distr, ignore.case = TRUE) ~ "beta",
+        grepl("^scaled t", distr, ignore.case = TRUE) ~ "scaled_t",
+        grepl("^Ordered Categorical", distr, ignore.case = TRUE) ~ "ocat",
+        grepl("^zero inflated Poisson", distr, ignore.case = TRUE) ~ "zip",
+        grepl("^Cox PH", distr, ignore.case = TRUE) ~ "cox_ph",
+        grepl("^censored normal", distr, ignore.case = TRUE) ~ "cnorm",
+        .default = as.character(distr)
+    )
 
     ## which link function
     lfun <-
@@ -359,7 +347,8 @@ family_type.family <- function(object, ...) {
                             which_eta = which_eta),
              multinom = multinom_link(object, parameter, inverse = inverse,
                                       which_eta = which_eta),
-             shash = shash_link(object, parameter, inverse = inverse)
+             shash = shash_link(object, parameter, inverse = inverse),
[…]

### 2023-02-18 R/posterior-samples.R

@@ -6,7 +6,6 @@
 #'   the model will be used for `data`, if available in `model`.
 #' @param n numeric; the number of posterior samples to return.
 #' @param seed numeric; a random seed for the simulations.
-#' @param scale character;
 #' @param freq logical; `TRUE` to use the frequentist covariance matrix of
 #'   the parameter estimators, `FALSE` to use the Bayesian posterior
 #'   covariance matrix of the parameters.
@@ -15,13 +14,37 @@
 #'   used, if available.
 #' @param weights numeric; a vector of prior weights. If `data` is null
 #'   then defaults to `object[["prior.weights"]]`, otherwise a vector of ones.
-#' @param ncores number of cores for generating random variables from a
+#' @param n_cores number of cores for generating random variables from a
 #'   multivariate normal distribution. Passed to [mvnfast::rmvn()].
 #'   Parallelization will take place only if OpenMP is supported (but appears
 #'   to work on Windows with current `R`).
+#' @param method character; which method should be used to draw samples from
+#'   the posterior distribution. `"gaussian"` uses a Gaussian (Laplace)
+#'   approximation to the posterior. `"mh"` uses a Metropolis Hastings sample
+#'   that alternates t proposals with proposals based on a shrunken version of
+#'   the posterior covariance matrix. `"inla"` uses a variant of Integrated
+#'   Nested Laplace Approximation due to Wood (2019), (currently not
+#'   implemented). `"user"` allows for user-supplied posterior draws
+#'   (currently not implemented).
+#' @param burnin numeric; number of samples to discard as the burnin draws.
+#'   Only used with `method = "mh"`.
+#' @param thin numeric; the number of samples to skip when taking `n` draws.
+#'   Results in `thin * n` draws from the posterior being taken. Only used with
+#'   `method = "mh"`.
+#' @param t_df numeric; degrees of freedome for t distribution proposals. Only
+#'   used with `method = "mh"`.
+#' @param rw_scale numeric; Factor by which to scale posterior covariance
+#'   matrix when generating random walk proposals. Negative or non finite to
+#'   skip the random walk step. Only used with `method = "mh"`.
 #' @param ... arguments passed to other methods. For `fitted_samples()`, these
-#'   are passed on to `predict.gam()`.
[…]

### 2023-02-18 R/smooth-estimates.R

@@ -1675,6 +1675,14 @@
                               ...) {
     # handle splines on the sphere
 
+    # this currently needs the mapproj pkg for coord_map()
+    if (!requireNamespace("mapproj", quietly = TRUE)) {
+        message("\nPlotting SOS smooths uses `ggplot2::coord_map()`.\n",
+            "This requires that the {mapproj} package be installed.\n",
+            "Run: `install.packages(\"mapproj\")`\n")
+        stop("Package {mapproj} is not available.")
+    }
+
     if (is.null(variables)) {
         variables <- vars_from_label(unique(object[["smooth"]]))
     }

### 2023-02-18 tests/testthat/setup.R

@@ -25,7 +25,7 @@ su_m_quick_eg1 <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3),
                       data = quick_eg1,
                       method = "REML")
 
-su_m_quick_eg1_shrink <- gam(y ~ s(x0, bs = "ts") + s(x1, bs = "ts") +
+su_m_quick_eg1_shrink <- gam(y ~ s(x0, bs = "ts") + s(x1, bs = "cs") +
                                s(x2, bs = "ts") + s(x3, bs = "ts"),
                              data = quick_eg1,
                              method = "REML")
@@ -39,15 +39,19 @@ su_m_penalty <- gam(y ~ s(x0, bs = 'cr') + s(x1, bs = 'bs') +
                     data = su_eg1,
                     method = "REML")
 
-su_m_bivar <- gam(y ~ s(x, z, k = 40),
-                  data = su_eg2,
-                  method = "REML")
+su_m_bivar <- gam(y ~ s(x, z, k = 40), data = su_eg2, method = "REML")
+
+su_m_bivar_ds <- gam(y ~ s(x, z, k = 20, bs = "ds"), data = su_eg2,
+  method = "REML")
 
 su_m_trivar <- gam(y ~ s(x0, x1, x2), data = su_eg1, method = "REML")
 
 su_m_quadvar <- gam(y ~ s(x0, x1, x2, x3), data = su_eg1, method = "REML")
 
 su_m_bivar_te <- gam(y ~ te(x, z, k = c(5, 5)), data = su_eg2, method = "REML")
+
+su_m_bivar_ti <- gam(y ~ s(x, k = 5) + s(z, k = 5) + ti(x, z, k = c(5, 5)),
+  data = su_eg2, method = "REML")
 
 su_m_bivar_t2 <- gam(y ~ t2(x, z, k = c(5, 5)), data = su_eg2, method = "REML")
 
@@ -66,8 +70,7 @@ su_m_quadvar_t2 <- gam(y ~ t2(x0, x1, x2, x3, k = c(3, 3, 3, 3)),
 su_m_cont_by <- gam(y ~ s(x2, by = x1), data = su_eg3, method = "REML")
 
 su_m_factor_by <- gam(y ~ fac + s(x2, by = fac) + s(x0),
-                      data = su_eg4,
-                      method = "REML")
+                      data = su_eg4, method = "REML")
 
[…]

### 2023-02-18 tests/testthat/test-basis.R

@@ -37,3 +37,35 @@ test_that("basis() works with a scam", {
     plt <- draw(bs)
     expect_doppelganger("draw basis works with a scam single smooth", plt)
 })
+
+test_that("basis() works with bivariate tprs smooths", {
+    expect_silent(ds <- data_slice(su_eg2, x = evenly(x, n = 20),
+        z = evenly(z, n = 20)))
+    expect_silent(bs <- basis(s(x, z, k = 25), data = su_eg2, at = ds,
+        constraints = TRUE))
+    expect_s3_class(bs, "basis")
+
+    plt <- draw(bs)
+    expect_doppelganger("draw basis works with a bivariate tprs", plt)
+
+    skip_on_cran()
+    skip_on_os("mac")
+    plt <- draw(bs, contour = TRUE)
+    expect_doppelganger("draw basis works with a bivariate tprs contour", plt)
+})
+
+test_that("basis() works with bivariate te smooths", {
+    expect_silent(ds <- data_slice(su_eg2, x = evenly(x, n = 20),
+        z = evenly(z, n = 20)))
+    expect_silent(bs <- basis(te(x, z, k = c(5, 5)), data = su_eg2, at = ds,
+        constraints = TRUE))
+    expect_s3_class(bs, "basis")
+
+    plt <- draw(bs)
+    expect_doppelganger("draw basis works with a bivariate te", plt)
+
+    skip_on_cran()
+    skip_on_os("mac")
+    plt <- draw(bs, contour = TRUE)
+    expect_doppelganger("draw basis works with a bivariate te contour", plt)
+})

### 2023-02-18 tests/testthat/test-draw-methods.R

@@ -6,12 +6,6 @@ library("gratia")
 library("mgcv")
 library("ggplot2")
 
-## Need a local wrapper to allow conditional use of vdiffr
-# `expect_doppelganger` <- function(title, fig, ...) {
-#   testthat::skip_if_not_installed("vdiffr")
-#   vdiffr::expect_doppelganger(title, fig, ...)
-# }
-
 test_that("draw.evaluated_1d_smooth() plots the smooth", {
     withr::local_options(lifecycle_verbosity = "quiet")
     sm <- evaluate_smooth(su_m_univar_4, "s(x2)")
@@ -347,6 +341,31 @@ test_that("draw.derivates() plots derivatives for a GAM rotated labels", {
         plt)
 })
 
+test_that("draw plots partial derivatives for a GAM", {
+    d1 <- partial_derivatives(su_m_bivar_te, term = "te(x,z)", focal = "z",
+        type = "central")
+    plt <- draw(d1)
+    expect_doppelganger("draw partial derivatives for a GAM", plt)
+
+    plt <- draw(d1, scales = "fixed")
+    ## skip_on_ci()
+    expect_doppelganger("draw partial derivatives for a GAM with fixed scales",
+        plt)
+})
+
+test_that("draw plots partial derivs for GAM rotated labels", {
+    skip_on_cran()
+    d1 <- partial_derivatives(su_m_bivar_te, term = "te(x,z)", focal = "z",
+        type = "central")
+    plt <- draw(d1, angle = 45)
+    expect_doppelganger("draw partial derivatives for GAM rotated labels",
+        plt)
+
+    plt <- draw(d1, scales = "fixed", angle = 45)
+    expect_doppelganger("draw partial derivatives for GAM fixed scales rotated",
+        plt)
+})
+
 ## test that issue 39 stays fixed
 test_that("draw.gam doesn't create empty plots with multiple parametric terms", {
     plt <- draw(m_2_fac, rug = FALSE)
@@ -553,4 +572,12 @@ test_that("draw.penalty_df works with normalization", {
     plt <- draw(penalty(su_m_univar_4, "s(x1)"), normalize = TRUE)
     expect_doppelganger("draw penalty_df with single smooths normalized",
                         plt)
+})
+
+test_that("plotting sos smooths works", {
+    skip_on_cran()
+    skip_if_not_installed("mapproj")
+    skip_on_os("mac")
+    expect_silent(plt <- draw(m_sos, n = 25))
+    expect_doppelganger("draw works for sos smooths", plt)
 })

### 2023-02-18 tests/testthat/test-evenly.R

@@ -9,3 +9,27 @@ test_that("evenly with by works", {
     expect_identical(length(x), 7L)
     expect_identical(seq(1987, 1999, by = 2), x)
 })
+
+test_that("seq min max with by works", {
+    expect_silent(x <- seq_min_max(1987:1999, by = 2))
+    expect_identical(length(x), 7L)
+    expect_identical(seq(1987, 1999, by = 2), x)
+})
+
+test_that("ref_level works", {
+    f <- factor(sample(letters[1:5], 100, replace = TRUE),
+        levels = letters[1:5])
+    expect_silent(rl <- ref_level(f))
+    expect_identical(factor("a", levels = letters[1:5]), rl)
+
+    expect_error(ref_level(1:10), "'fct' must be a factor")
+})
+
+test_that("ref_level works", {
+    f <- factor(sample(letters[1:5], 100, replace = TRUE),
+        levels = letters[1:5])
+    expect_silent(rl <- level(f, "b"))
+    expect_identical(factor("b", levels = letters[1:5]), rl)
+
+    expect_error(level(1:10), "'fct' must be a factor")
+})

### 2023-02-18 tests/testthat/test-family-utils.R

@@ -223,6 +223,13 @@ test_that("link() works for cox.ph() family objects", {
     expect_identical(f, cox.ph()$linkfun)
 })
 
+test_that("link() works for cnorm() family objects", {
+    f <- link(cnorm())
+    expect_type(f, "closure")
+    expect_identical(f(val), val)
+    expect_identical(f, cnorm()$linkfun)
+})
+
 ## inv_link
 test_that("inv_link() works for gaussian() family objects", {
     f <- inv_link(gaussian())
@@ -344,6 +351,13 @@ test_that("inv_link() works for cox.ph() family objects", {
     expect_type(f, "closure")
     expect_identical(f(val), cox.ph()$linkinv(val))
     expect_identical(f, cox.ph()$linkinv)
+})
+
+test_that("inv_link() works for cnorm() family objects", {
+    f <- inv_link(cnorm())
+    expect_type(f, "closure")
+    expect_identical(f(val), val)
+    expect_identical(f, cnorm()$linkinv)
 })
 
 test_that("extract_link() works on gaussian() family objects", {
@@ -997,6 +1011,20 @@ test_that("extract_link() works on shash() family objects", {
     expect_identical(f(val), fam$linfo[[4L]]$linkinv(val))
     expect_identical(f, fam$linfo[[4L]]$linkinv)
 })
+
+test_that("extract_link() works on cnorm() family objects", {
+    ## link
+    f <- extract_link(cnorm())
+    expect_type(f, "closure")
+    expect_identical(f(val), val)
+    expect_identical(f, cnorm()$linkfun)
+    ## inverse
+    f <- extract_link(cnorm(), inverse = TRUE)
+    expect_type(f, "closure")
+    expect_identical(f(val), val)
+    expect_identical(f, cnorm()$linkinv)
+})
+
 
 ## tests some specific extract functions
 test_that("twlss_link() can extract a link function", {

### 2023-02-18 tests/testthat/test-posterior-samples.R

@@ -8,17 +8,46 @@ library("gratia")
 test_that("smooth_samples works for a continuous by GAM", {
     expect_silent(sm <- smooth_samples(su_m_cont_by, n = 5, n_vals = 100,
                                        seed = 42))
-    expect_s3_class(sm, c("smooth_samples", "posterior_samples", "tbl_df",
+    expect_s3_class(sm, c("smooth_samples", "tbl_df",
                           "tbl", "data.frame"))
     ## 500 == 1 smooth * 5 * 100
     expect_identical(NROW(sm), 500L)
     expect_identical(NCOL(sm), 8L) # 8 cols, univatiate smooths
+    skip_on_cran()
+    skip_on_ci()
+    expect_snapshot(sm)
 })
 
 test_that("smooth_samples works for a simple GAM", {
     expect_silent(sm <- smooth_samples(m_1_smooth, n = 5, n_vals = 100,
                                        seed = 42))
-    expect_s3_class(sm, c("smooth_samples", "posterior_samples", "tbl_df",
+    expect_s3_class(sm, c("smooth_samples", "tbl_df",
+                          "tbl", "data.frame"))
+    ## 500 == 1 smooth * 5 * 100
+    expect_identical(NROW(sm), 500L)
+    expect_identical(NCOL(sm), 8L) # 8 cols, univatiate smooths
+    skip_on_cran()
+    skip_on_ci()
+    expect_snapshot(sm)
+})
+
+test_that("smooth_samples works for a simple GAM multi rng calls", {
+    expect_silent(sm <- smooth_samples(m_1_smooth, n = 5, n_vals = 100,
+                                       seed = 42, rng_per_smooth = TRUE))
+    expect_s3_class(sm, c("smooth_samples", "tbl_df",
+                          "tbl", "data.frame"))
+    ## 500 == 1 smooth * 5 * 100
+    expect_identical(NROW(sm), 500L)
+    expect_identical(NCOL(sm), 8L) # 8 cols, univatiate smooths
+    skip_on_cran()
+    skip_on_ci()
+    expect_snapshot(sm)
+})
+
+test_that("smooth_samples works for a simple GAM MH sampling", {
+    expect_silent(sm <- smooth_samples(m_1_smooth, n = 5, n_vals = 100,
+                                       method = "mh", seed = 42))
+    expect_s3_class(sm, c("smooth_samples", "tbl_df",
                           "tbl", "data.frame"))
[…]

### 2023-03-11 README.Rmd
update the readme as so smooths do work

@@
 
-    Note specialist smoothers (`bs %in% c("mrf","so")`) are not currently
-    supported, but univariate, *factor* and *continuous* `by`-variable smooths,
-    simple random effect smooths (`bs = 're'`), factor-smooth interaction
-    smooths (`bs = "fs"`), and bivariate tensor product smooths are supported,
+    Note that some specialist smoothers (`bs %in% c("mrf","sw", "sf")`) are not
+    currently supported, but univariate, *factor* and *continuous* `by`-variable
+    smooths, simple random effect smooths (`bs = 're'`), factor-smooth
+    interaction smooths (`bs = "fs"`), constrained factor smooths (`bs = "sz"`),
+    full soap film smooths (`bs = "so"`), and bivariate, trivariate, and
+    quadvariate TPRS and tensor product smooths are supported,
 

### 2023-03-13 README.Rmd
codecov is redirecting their URLs and check catches this

@@
 [![R build status](https://github.com/gavinsimpson/gratia/workflows/R-CMD-check/badge.svg)](https://github.com/gavinsimpson/gratia/actions)
-[![codecov](https://codecov.io/gh/gavinsimpson/gratia/branch/main/graph/badge.svg?token=GG5NQfgRFu)](https://codecov.io/gh/gavinsimpson/gratia)
+[![codecov](https://codecov.io/gh/gavinsimpson/gratia/branch/main/graph/badge.svg?token=GG5NQfgRFu)](https://app.codecov.io/gh/gavinsimpson/gratia)
 [![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/gratia)](https://cran.r-project.org/package=gratia)

### 2023-03-14 README.Rmd
Deploying to gh-pages from @ gavinsimpson/gratia@88e289c9186924b234f185dfb8820d512726c988 🚀

@@
 ```{r install-r-universe, eval = FALSE}
-# Enable repository from gavinsimpson
-options(repos = c(
-  gavinsimpson = "https://gavinsimpson.r-universe.dev",
-  CRAN = "https://cloud.r-project.org"))
-
-# Download and install gratia in R
-install.packages("gratia")
+# Install gratia in R
+install.packages("gratia", repos = c("https://gavinsimpson.r-universe.dev",
+  "https://cloud.r-project.org"))
 ```

### 2023-04-06 tests/testthat/test-response-derivatives.R
fix warning about incorrect usage in tidyselect; add tests for response derivatives

@@ -1,0 +1,25 @@
+# uses df_pois & m_nb
+
+# response derivatives
+test_that("response derivatives works", {
+    skip_on_cran()
+    N <- 50L
+    expect_silent(ds <- data_slice(m_nb, x2 = evenly(x2, n = N)))
+    expect_silent(yd <- response_derivatives(m_nb, data = ds,
+        type = "central", focal = "x2"))
+    expect_s3_class(yd, class = "response_derivatives")
+    expect_identical(nrow(yd), N)
+    expect_snapshot(print(yd))
+
+    expect_silent(yd <- response_derivatives(m_nb, data = ds,
+        type = "forward", focal = "x2"))
+    expect_s3_class(yd, class = "response_derivatives")
+    expect_identical(nrow(yd), N)
+    expect_snapshot(print(yd))
+
+    expect_silent(yd <- response_derivatives(m_nb, data = ds,
+        type = "backward", focal = "x2"))
+    expect_s3_class(yd, class = "response_derivatives")
+    expect_identical(nrow(yd), N)
+    expect_snapshot(print(yd))
+})

### 2023-04-06 R/derivatives.R
Deploying to gh-pages from @ gavinsimpson/gratia@e61499c3b651731a06d5d8b39c1ec5569c98f190 🚀

@@ -987,7 +987,7 @@
 #'
 #' # response derivatives
 #' y_d <- response_derivatives(m, data = ds, type = "central",
-#'                             focal = "x2", eps = 0.01)
+#'                             focal = "x2", eps = 0.01, seed = 21)
 #'
 #' # draw fitted values along x2
 #' p1 <- fv |>

### 2023-09-25 tests/test-all.R
all the fricking snapshots failed again and all the errors aren't my fault

@@ -2,6 +2,7 @@
 
 ## Setup
 library("testthat")
+library("gratia")
 
 ## Runs the tests in tests/testthat
 test_check("gratia")

### 2023-09-25 tests/testthat/test-diagnostic-plots.R

@@ -65,9 +65,9 @@ test_that("worm_plot works for a GLM", {
 })
 
 test_that("worm_plot works for a LM", {
-    expect_silent(plt2 <- withr::with_seed(1, worm_plot(m_lm)))
+    expect_silent(plt1 <- withr::with_seed(1, worm_plot(m_lm)))
 
-    expect_silent(pl2t <- withr::with_seed(1, worm_plot(m_lm,
+    expect_silent(plt2 <- withr::with_seed(1, worm_plot(m_lm,
         method = "simulate")))
 
     expect_silent(plt3 <- withr::with_seed(1, worm_plot(m_lm,

### 2023-09-25 tests/testthat/test-difference-smooths.R

@@ -89,7 +89,7 @@ test_that("difference_smooths() works for a gamm4 model", {
 test_that("difference_smooths() works with user data", {
     df <- with(su_eg4,
                expand_grid(fac = factor(levels(fac), levels = levels(fac)),
-                           x2 = seq_min_max(x2, n = 20),
+                           x2 = evenly(x2, n = 100),
                            x0 = mean(x0)))
     expect_silent(ds <-
       difference_smooths(su_m_factor_by, smooth = "s(x2)", data = df))
@@ -99,7 +99,7 @@ test_that("difference_smooths() works with user data", {
     plt <- draw(ds)
 
     skip_on_ci()
-    expect_doppelganger("draw difference_smooths gam", plt)
+    expect_doppelganger("draw difference_smooths gam user data", plt)
 })
 
 test_that("difference_smooths() works for a bivariate gam", {

### 2023-09-25 tests/testthat/test-draw-methods.R

@@ -99,8 +99,8 @@ test_that("draw.evaluated_2d_smooth() plots the smooth", {
     skip_on_ci()
     expect_silent(plt1 <- draw(sm))
     expect_silent(plt2 <- draw(sm, contour_col = "red"))
-    expect_doppelganger("draw 2d smooth", plt)
-    expect_doppelganger("draw 2d smooth diff contour colour", plt)
+    expect_doppelganger("draw 2d smooth", plt1)
+    expect_doppelganger("draw 2d smooth diff contour colour", plt2)
 })
 
 test_that("draw.evaluated_2d_smooth() plots the smooth without contours", {
@@ -173,8 +173,8 @@ test_that("draw.gam() plots an AM with a single factor by-variable smooth", {
     plt2 <- draw(su_m_factor_by, scales = "fixed", rug = FALSE)
 
     skip_on_ci()
-    expect_doppelganger("draw factor by-variable smooth with fixed scales", plt1)
-    expect_doppelganger("draw AM with factor by-variable smooth", plt2)
+    expect_doppelganger("draw AM with factor by-variable smooth", plt1)
+    expect_doppelganger("draw factor by-variable smooth with fixed scales", plt2)
 })
 
 ## simulate date from y = f(x2)*x1 + error
@@ -676,8 +676,9 @@ test_that("draw.penalty_df accepts user-specified continuous_fill", {
 
     expect_silent(pen <- penalty(su_m_univar_4))
     plt1 <- draw(pen,
-                continuous_fill = scale_fill_distiller(palette = "Spectral",
-                                                       type = "div"))
+        continuous_fill = scale_fill_distiller(palette = "Spectral",
+            type = "div"))
+    
     plt2 <- draw(penalty(su_m_univar_4, "s(x1)"),
         continuous_fill = scale_fill_distiller(palette = "Spectral",
             type = "div"))
@@ -698,9 +699,9 @@ test_that("draw.penalty_df works with normalization", {
     plt2 <- draw(penalty(su_m_univar_4, "s(x1)"), normalize = TRUE)
 
     skip_on_ci()
-    expect_doppelganger("draw penalty_df with single smooths normalized",
-                        plt1)
     expect_doppelganger("draw penalty_df with multiple smooths normalized",
+                        plt1)
+    expect_doppelganger("draw penalty_df with single smooths normalized",
                         plt2)
 })
 

### 2023-09-26 tests/testthat/test-posterior-samples.R
update all tests as needed

@@ -118,7 +118,7 @@ test_that("fitted_samples works for a simple GAM", {
     ## 1000 == 5 * 200 (nrow(dat))
     expect_identical(NROW(sm), 1500L)
     expect_identical(NCOL(sm), 3L) # 3 cols
-    expect_named(sm, expected = c("row", "draw", "fitted"))
+    expect_named(sm, expected = c(".row", ".draw", ".fitted"))
 })
 
 test_that("fitted_samples works for a multi-smooth GAM", {
@@ -128,7 +128,7 @@ test_that("fitted_samples works for a multi-smooth GAM", {
     ## 5000 == 5 draws * 1000 observations in data
     expect_identical(NROW(sm), 5000L)
     expect_identical(NCOL(sm), 3L) # 3 cols
-    expect_named(sm, expected = c("row", "draw", "fitted"))
+    expect_named(sm, expected = c(".row", ".draw", ".fitted"))
 })
 
 test_that("fitted_samples works for a multi-smooth factor by GAM", {
@@ -138,7 +138,7 @@ test_that("fitted_samples works for a multi-smooth factor by GAM", {
     ## 2000 == 5 draws * 400 observations in data
     expect_identical(NROW(sm), 2000L)
     expect_identical(NCOL(sm), 3L) # 3 cols
-    expect_named(sm, expected = c("row", "draw", "fitted"))
+    expect_named(sm, expected = c(".row", ".draw", ".fitted"))
 })
 
 test_that("fitted_samples sets seed when seed not provided", {
@@ -158,7 +158,7 @@ test_that("predicted_samples works for a simple GAM", {
     ## 2000 == 5 * 100 (nrow(dat))
     expect_identical(NROW(sm), 1500L)
     expect_identical(NCOL(sm), 3L) # 3 cols
-    expect_named(sm, expected = c("row", "draw", "response"))
+    expect_named(sm, expected = c(".row", ".draw", ".response"))
 })
 
 test_that("predicted_samples works for a multi-smooth GAM", {
@@ -168,7 +168,7 @@ test_that("predicted_samples works for a multi-smooth GAM", {
     ## 5000 == 5 draws * 1000 observations in data
     expect_identical(NROW(sm), 5000L)
     expect_identical(NCOL(sm), 3L) # 3 cols
-    expect_named(sm, expected = c("row", "draw", "response"))
+    expect_named(sm, expected = c(".row", ".draw", ".response"))
 })
 
[…]

### 2023-09-26 R/derivatives.R
Deploying to gh-pages from @ gavinsimpson/gratia@65f7a909cabbb55e8162e31b1f98b5100ea08139 🚀

@@ -958,11 +958,13 @@
 #' @rdname response_derivatives
 #'
 #' @return A tibble, currently with the following variables:
-#' * `focal`: the name of the variable for which the partial derivative was
-#'     evaluated,
-#' * `derivative`: the estimated partial derivative,
-#' * `lower`: the lower bound of the confidence or simultaneous interval,
-#' * `upper`: the upper bound of the confidence or simultaneous interval,
+#' * `.row`: integer, indexing the row of `data` each row in the output
+#'   represents
+#' * `.focal`: the name of the variable for which the partial derivative was
+#'   evaluated,
+#' * `.derivative`: the estimated partial derivative,
+#' * `.lower_ci`: the lower bound of the confidence or simultaneous interval,
+#' * `.upper_ci`: the upper bound of the confidence or simultaneous interval,
 #' * additional columns containing the covariate values at which the derivative
 #'   was eveluated.
 #'
@@ -987,9 +989,9 @@
 #' # fitted values along x2
 #' fv <- fitted_values(m, smooth = "s(x2)", data = ds)
 #'
-#' # response derivatives
-#' y_d <- response_derivatives(m, data = ds, type = "central",
-#'                             focal = "x2", eps = 0.01, seed = 21)
+#' # response derivatives - ideally n_sim = >10000
+#' y_d <- response_derivatives(m, data = ds, type = "central", focal = "x2",
+#'     eps = 0.01, seed = 21, n_sim = 1000)
 #'
 #' # draw fitted values along x2
 #' p1 <- fv |>
@@ -1001,8 +1003,8 @@
 #'
 #' # draw response derivatives
 #' p2 <- y_d |>
-#'     ggplot(aes(x = x2, y = derivative)) +
-#'     geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.2) +
+#'     ggplot(aes(x = x2, y = .derivative)) +
+#'     geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
 #'     geom_line() +
 #'     labs(title = "Estimated 1st derivative of estimated count",
 #'          y = "First derivative")
@@ -1023,22 +1025,21 @@
     ...) {
     yd <- derivative_samples(object, focal = focal, data = data, order = order,
         type = type, scale = scale, method = method, n = n, eps = eps,
-        n_sim = n_sim)
-    
[…]

### 2023-09-26 R/posterior-samples.R

@@ -139,12 +139,13 @@
 
     # need to extend weights by number of draws
     weights <- rep(weights, times = n)
-    #scale <- rep(scale, times = n)
+    # scale <- rep(scale, times = n)
 
     # replace fitted with
     sim_eta <- sim_eta |>
-        mutate(response = rd_fun(mu = fitted, wt = weights, scale = scale)) |>
-        select(all_of(c("row", "draw", "response")))
+        mutate(.response = rd_fun(mu = .data$.fitted, wt = weights,
+            scale = scale)) |>
+        select(all_of(c(".row", ".draw", ".response")))
 
     ## add classes
     class(sim_eta) <- c("posterior_samples", class(sim_eta))
@@ -206,6 +207,9 @@
 #' @rdname fitted_samples
 #'
 #' @importFrom stats vcov coef predict
+#' @importFrom dplyr mutate
+#' @importFrom tidyr gather
+#' @importFrom tibble as_tibble
 #' @importFrom mvnfast rmvn
 #'
 #' @examples
@@ -278,9 +282,9 @@
     colnames(sims) <- paste0(".V", seq_len(NCOL(sims)))
     sims <- as_tibble(sims)
     names(sims) <- as.character(seq_len(ncol(sims)))
-    sims <- add_column(sims, row = seq_len(nrow(sims)))
-    sims <- gather(sims, key = "draw", value = "fitted", - row)
-    sims[["draw"]] <- as.integer(sims[["draw"]])
+    sims <- add_column(sims, .row = seq_len(nrow(sims)))
+    sims <- gather(sims, key = ".draw", value = ".fitted", - .row) |>
+        mutate(.draw = as.integer(.data$.draw))
     attr(sims, "seed") <- RNGstate
     ## add classes
     class(sims) <- c("fitted_samples", class(sims))
@@ -361,9 +365,9 @@
     colnames(sims) <- paste0(".V", seq_len(NCOL(sims)))
     sims <- as_tibble(sims)
     names(sims) <- as.character(seq_len(ncol(sims)))
-    sims <- add_column(sims, row = seq_len(nrow(sims)))
-    sims <- gather(sims, key = "draw", value = "response", - row)
-    sims[["draw"]] <- as.integer(sims[["draw"]])
+    sims <- add_column(sims, .row = seq_len(nrow(sims)))
+    sims <- gather(sims, key = ".draw", value = ".response", - .row) |>
+        mutate(.draw = as.integer(.data$.draw))
     attr(sims, "seed") <- RNGstate
     ## add classes
[…]

### 2023-09-26 tests/testthat/test-response-derivatives.R

@@ -12,17 +12,17 @@ test_that("response derivatives works", {
         type = "central", focal = "x2", seed = 2))
     expect_s3_class(yd, class = "response_derivatives")
     expect_identical(nrow(yd), N)
-    expect_snapshot(print(yd))
+    expect_snapshot(print(yd), variant = "central")
 
     expect_silent(yd <- response_derivatives(m_nb, data = ds,
         type = "forward", focal = "x2", seed = 2))
     expect_s3_class(yd, class = "response_derivatives")
     expect_identical(nrow(yd), N)
-    expect_snapshot(print(yd))
+    expect_snapshot(print(yd), variant = "forward")
 
     expect_silent(yd <- response_derivatives(m_nb, data = ds,
         type = "backward", focal = "x2", seed = 2))
     expect_s3_class(yd, class = "response_derivatives")
     expect_identical(nrow(yd), N)
-    expect_snapshot(print(yd))
+    expect_snapshot(print(yd), variant = "backward")
 })

### 2023-10-13 R/add-methods.R
Deploying to gh-pages from @ gavinsimpson/gratia@a60e224cd12a1eaa0f8bcced7d456c81772d1276 🚀

@@ -245,8 +245,9 @@
                  call. = FALSE)
         }
         object <- mutate(object,
-                         across(any_of(c("est", "lower", "upper")),
-                                .fns = ~ .x + constant))
+            across(any_of(c("est", "lower", "upper", ".estimate", ".lower_ci",
+            ".upper_ci")),
+            .fns = ~ .x + constant))
     }
 
     object
@@ -264,8 +265,9 @@
                  call. = FALSE)
         }
         object <- mutate(object,
-                         across(any_of(c("est", "lower_ci", "upper_ci")),
-                                .fns = ~ .x + constant))
+            across(any_of(c("est", "lower_ci", "upper_ci", ".estimate", ".lower_ci",
+                ".upper_ci")),
+            .fns = ~ .x + constant))
     }
 
     object
@@ -283,8 +285,9 @@
                  call. = FALSE)
         }
         object <- mutate(object,
-                         across(any_of(c("value", "lower", "upper")),
-                                .fns = ~ .x + constant))
+            across(any_of(c("value", "lower", "upper",
+                ".value", ".lower_ci", ".upper_ci")),
+            .fns = ~ .x + constant))
     }
 
     object
@@ -302,8 +305,9 @@
                  call. = FALSE)
         }
         object <- mutate(object,
-                         across(any_of(c("est", "lower_ci", "upper_ci")),
-                                .fns = ~ .x + constant))
+            across(any_of(c("est", "lower_ci", "upper_ci", ".estimate",
+            ".lower_ci", ".upper_ci")),
+                .fns = ~ .x + constant))
     }
 
     object
@@ -322,8 +326,9 @@
                  ">", call. = FALSE)
         }
         object <- mutate(object,
-                         across(any_of(c("partial", "lower", "upper")),
-                                .fns = ~ .x + constant))
+            across(any_of(c("partial", "lower", "upper", ".estimate",
+            ".lower_ci", ".upper_ci")),
+                .fns = ~ .x + constant))
     }
 
     object
@@ -394,9 +399,9 @@
     # check if this is nested
     nms <- names(object)
 
-    if (!all(c("est", "se") %in% nms)) {
+    if (!all(c(".estimate", ".se") %in% nms)) {
         stop("'object' does not contain one or both of ",
-             "'est', 'se'.",
+             "'.estimate', '.se'.",
[…]

### 2023-10-13 R/compare-smooths.R

@@ -124,18 +124,16 @@
     crit <- coverage_normal(coverage)
 
     ## add the frequentist confidence interval
-    object <- mutate(object,
-                     lower_ci = .data[["est"]] + (crit * .data[["se"]]),
-                     upper_ci = .data[["est"]] - (crit * .data[["se"]]))
+    object <- object |> add_confint(coverage = coverage)
 
     ## basic plot
     plt <- ggplot(object, aes(x = .data[[sm_vars[1L]]],
-                              y = .data[["est"]],
+                              y = .data[[".estimate"]],
                               group = .data[["model"]]))
 
     ## add uncertainty bands
-    plt <- plt + geom_ribbon(aes(ymin = .data[["lower_ci"]],
-                                 ymax = .data[["upper_ci"]],
+    plt <- plt + geom_ribbon(aes(ymin = .data[[".lower_ci"]],
+                                 ymax = .data[[".upper_ci"]],
                                  fill = .data[["model"]]),
                              alpha = 0.2)
 

### 2023-10-13 R/confint-methods.R

@@ -329,7 +329,7 @@
         for (i in seq_along(out)) {
             out[[i]] <- smooth_estimates(object, smooth = uS[i],
                 n = n, data = data, partial_match = partial_match)
-            out[[i]][["crit"]] <- coverage_normal(level)
+            out[[i]][[".crit"]] <- coverage_normal(level)
         }
     } else {
         ## function to do simultaneous intervals for a smooth
@@ -343,7 +343,7 @@
             para.seq <- start:end
             Cg <- PredictMat(smooth, data)
             simDev <- Cg %*% t(buDiff[, para.seq])
-            absDev <- abs(sweep(simDev, 1L, data[["se"]], FUN = "/"))
+            absDev <- abs(sweep(simDev, 1L, data[[".se"]], FUN = "/"))
             masd <- apply(absDev, 2L, max)
             unname(quantile(masd, probs = level, type = 8))
         }
@@ -366,12 +366,11 @@
             if (is.null(by_levs)) {        # not by variable smooth
                 smooth <- get_smooth(object, parm) # get the specific smooth
                 crit <- sim_interval(smooth, level = level, data = out[[i]])
-                out[[i]][["crit"]] <- crit # add on the critical value
             } else {                       # is a by variable smooth
                 smooth <- old_get_smooth(object, uS[i])
                 crit <- sim_interval(smooth, level = level, data = out[[i]])
-                out[[i]][["crit"]] <- crit # add on the critical value
             }
+            out[[i]][[".crit"]] <- crit # add on the critical value
         }
     }
 
@@ -396,18 +395,18 @@
 
     ## using se and crit, compute the lower and upper intervals
     out <- add_column(out,
-                      lower = out$est - (out$crit * out$se),
-                      upper = out$est + (out$crit * out$se))
+                      .lower_ci = out$.estimate - (out$.crit * out$.se),
+                      .upper_ci = out$.estimate + (out$.crit * out$.se))
 
     ## transform
-    out[["est"]]   <- ilink(out[["est"]] + const)
-    out[["lower"]] <- ilink(out[["lower"]] + const)
-    out[["upper"]] <- ilink(out[["upper"]] + const)
+    out <- out |>
+        mutate(across(all_of(c(".estimate", ".lower_ci", ".upper_ci")),
+            .fns = ilink))
 
     # smooth_estimates has columns in different places, relocate them to match
     # old output as much as possible
[…]

### 2023-10-13 R/fitted_values.R

@@ -59,8 +59,12 @@
                                     "linear predictor"),
                                 ci_level = 0.95, ...) {
     # Handle everything up to and including the extended families, but not more
+    fn <- family_type(object)
     if (inherits(family(object), "general.family")) {
-        stop("General likelihood GAMs not yet supported.")
+        allowed <- c("gaulss", "gammals", "gumbls", "gevlss", "shash", "ziplss")
+        if (!fn %in% allowed) {
+            stop("General likelihood GAMs not yet supported.")
+        }
     }
     scale <- match.arg(scale)
 
@@ -73,10 +77,18 @@
 
     # handle special distributions that return more than vector fit & std. err.
     # find the name of the function that produces fitted values for this family
-    fit_vals_fun <- get_fit_fun(object)
+    fit_vals_fun <- get_fit_fun(fn)
+    extra_fns <- switch(fn,
+        "gumbls"  = post_link_funs(location = exp, scale = exp),
+        "gammals" = post_link_funs(location = exp, scale = exp),
+        "gevlss"  = post_link_funs(scale = exp),
+        "shash"   = post_link_funs(scale = exp, kurtosis = exp),
+        "ziplss"  = post_link_funs(location = exp,
+            pi = inv_link(binomial("cloglog"))),
+        post_link_funs())
     # compute fitted values
     fit <- fit_vals_fun(object, data = data, ci_level = ci_level,
-        scale = scale, ...)
+        scale = scale, extra_fns = extra_fns, ...)
     fit
 }
 
@@ -94,23 +106,157 @@
     fit <- predict(object, newdata = data, ..., type = "link",
         se.fit = TRUE) |>
         as.data.frame() |>
-        rlang::set_names(c("fitted", "se")) |>
+        rlang::set_names(c(".fitted", ".se")) |>
         as_tibble()
     fit <- bind_cols(data, fit)
 
     # create the confidence interval
     crit <- coverage_normal(ci_level)
     fit <- mutate(fit,
-        lower = .data[["fitted"]] - (crit * .data[["se"]]),
-        upper = .data[["fitted"]] + (crit * .data[["se"]]))
+        ".lower_ci" = .data[[".fitted"]] - (crit * .data[[".se"]]),
+        ".upper_ci" = .data[[".fitted"]] + (crit * .data[[".se"]]))
 
     # convert to the response scale if requested
[…]

### 2023-10-13 R/parametric-effects.R

@@ -114,29 +114,29 @@
             eval(term_expr, data, enclos = envir)
         }
 
-        out <- bind_cols(level = x_data,
-                         partial = pred[["fit"]][, term],
-                         se = pred[["se.fit"]][, term]) |>
-          distinct(.data$level, .keep_all = TRUE)
+        out <- bind_cols(.level = x_data,
+                         .partial = pred[["fit"]][, term],
+                         .se = pred[["se.fit"]][, term]) |>
+          distinct(.data$.level, .keep_all = TRUE)
         nr <- nrow(out)
-        is_fac <- is.factor(out$level)
+        is_fac <- is.factor(out$.level)
         type <- "factor"
         if (!is_fac) {
-            type <- class(out$level) # numeric but I presume logical too
+            type <- class(out$.level) # numeric but I presume logical too
             out <- out |>
-                rename("value" = "level")
+                rename(".value" = ".level")
         } else {
-            if (is.ordered(out$level)) {
+            if (is.ordered(out$.level)) {
                 type <- "ordered"
             }
             out <- out |>
-            mutate(level = as.character(.data$level))
+            mutate(.level = as.character(.data$.level))
         }
         out <- out |>
           add_column(term = rep(term, times = nr),
                      type = rep(type, times = nr),
                      .before = 1L) %>%
-          nest(data = any_of(c("level", "value", "partial", "se")))
+          nest(data = any_of(c(".level", ".value", ".partial", ".se")))
         out
     }
 
@@ -145,7 +145,7 @@
 
     if (unnest) {
         effects <- unnest(effects, cols = "data") |>
-          relocate(c("partial", "se"), .after = last_col())
+          relocate(c(".partial", ".se"), .after = last_col())
     }
 
     ## add confidence interval -- be consistent and don't add this, but could?
@@ -190,8 +190,8 @@ draw.parametric_effects <- function(object,
     # Add CI
     crit <- coverage_normal(ci_level)
     object <- mutate(object,
-                     lower = .data$partial - (crit * .data$se),
-                     upper = .data$partial + (crit * .data$se))
+                     .lower_ci = .data$.partial - (crit * .data$.se),
+                     .upper_ci = .data$.partial + (crit * .data$.se))
 
     # fixed or free?
     scales <- match.arg(scales)
[…]

### 2023-10-13 R/posterior-samples.R

@@ -440,6 +440,146 @@
          class(model)[[1L]], ">", .call = FALSE)
 }
 
+#' @export
+`smooth_samples.scam` <- function(model, term = NULL, n = 1, data = newdata,
+    method = c("gaussian", "mh", "inla", "user"),
+    seed = NULL,
+    freq = FALSE,
+    unconditional = FALSE,
+    n_cores = 1L,
+    n_vals = 200,
+    burnin = 1000,
+    thin = 1,
+    t_df = 40,
+    rw_scale = 0.25,
+    rng_per_smooth = FALSE,
+    ...,
+    newdata = NULL,
+    ncores = NULL) {
+    # smooth_samples begins
+    if (!exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)) {
+        runif(1)
+    }
+    if (is.null(seed)) {
+        RNGstate <- get(".Random.seed", envir = .GlobalEnv)
+    } else {
+        R.seed <- get(".Random.seed", envir = .GlobalEnv)
+        set.seed(seed)
+        RNGstate <- structure(seed, kind = as.list(RNGkind()))
+        on.exit(assign(".Random.seed", R.seed, envir = .GlobalEnv))
+    }
+
+    S <- smooths(model)             # vector of smooth labels - "s(x)"
+    take <- seq_along(S)            # in default case 1,2,3,..,n smooths
+    if (!is.null(term)) {
+        take <- which_smooths(model, term)
+        S <- S[take]
+    }
+
+    # At least for now, don't work for random effect smooths
+    # do we have ranef smooths?
+    re_sms <- vapply(get_smooths_by_id(model, take), FUN = is_re_smooth,
+                     FUN.VALUE = logical(1L))
+    if (any(re_sms)) {
+        message("\nRandom effect smooths not currently supported.\nIgnoring:",
+                " <", paste(S[re_sms], collapse = ", "), ">\n")
+        S <- S[!re_sms]
+        take <- take[!re_sms]
+    }
+
+    # do we have any remaining terms?
+    if (length(S) < 1L) {
+        stop("No smooths left that can be sampled from.")
+    }
+
+    if (!is.null(newdata)) {
+        newdata_deprecated()
+    }
+
+    if (!is.null(ncores)) {
+        message("Argument `ncores` is deprecated. Use `n_cores` instead.")
+        n_cores <- ncores
+    }
+
+    need_data <- FALSE
+    if (is.null(data)) {
+        need_data <- TRUE
+    }
+
[…]

### 2023-10-13 R/smooth-type.R

@@ -173,14 +173,14 @@
 #' @export
 #' @rdname smooth_type
 `smooth_type.mpi.smooth` <- function(smooth) {
-    sm_type <- "Monotone incr."
+    sm_type <- "Mono. incr."
     sm_type
 }
 
 #' @export
 #' @rdname smooth_type
 `smooth_type.mpd.smooth` <- function(smooth) {
-    sm_type <- "Monotone decr."
+    sm_type <- "Mono. decr."
     sm_type
 }
 
@@ -201,27 +201,27 @@
 #' @export
 #' @rdname smooth_type
 `smooth_type.micx.smooth` <- function(smooth) {
-    sm_type <- "Monotone incr. convex"
+    sm_type <- "Mono. incr. conv."
     sm_type
 }
 
 #' @export
 #' @rdname smooth_type
 `smooth_type.micv.smooth` <- function(smooth) {
-    sm_type <- "Monotone incr. concave"
+    sm_type <- "Mono. incr. conc."
     sm_type
 }
 
 #' @export
 #' @rdname smooth_type
 `smooth_type.mdcx.smooth` <- function(smooth) {
-    sm_type <- "Monotone decr. convex"
+    sm_type <- "Mono. decr. conv."
     sm_type
 }
 
 #' @export
 #' @rdname smooth_type
 `smooth_type.mdcv.smooth` <- function(smooth) {
-    sm_type <- "Monotone decr. concave"
+    sm_type <- "Mono. decr. conc."
     sm_type
 }

### 2023-10-13 tests/testthat/test-confint-methods.R

@@ -1,5 +1,7 @@
 ## Test confint() methods
 
+var_nms <- c(".estimate", ".se", ".crit", ".lower_ci", ".upper_ci")
+
 ## first derivatives of all smooths...
 test_that("Point-wise confidence interval for a first derivatives of a GAM works", {
     skip_if_not_installed("withr")
@@ -32,8 +34,7 @@ test_that("Point-wise confidence interval for a GAM works", {
     ci <- confint(m_gam, parm = "s(x1)", type = "confidence")
     expect_s3_class(ci, "confint.gam")
     expect_s3_class(ci, "tbl_df")
-    expect_named(ci, expected = c("smooth", "type", "by", "x1", "est", "se",
-                                  "crit", "lower", "upper"))
+    expect_named(ci, expected = c("smooth", "type", "by", "x1", var_nms))
 })
 
 test_that("Simultaneous interval for a GAM works", {
@@ -43,8 +44,7 @@ test_that("Simultaneous interval for a GAM works", {
         confint(m_gam, parm = "s(x1)", type = "simultaneous", nsim = 100))
     expect_s3_class(ci, "confint.gam")
     expect_s3_class(ci, "tbl_df")
-    expect_named(ci, expected = c("smooth", "type", "by", "x1", "est", "se",
-                                  "crit", "lower", "upper"))
+    expect_named(ci, expected = c("smooth", "type", "by", "x1", var_nms))
 })
 
 ## 2d smooth
@@ -54,8 +54,7 @@ test_that("Point-wise confidence interval for a 2d smooth works", {
     ci <- confint(su_m_bivar_te, parm = "te(x,z)", type = "confidence")
     expect_s3_class(ci, "confint.gam")
     expect_s3_class(ci, "tbl_df")
-    expect_named(ci, expected = c("smooth", "type", "by", "x", "z", "est",
-                                  "se", "crit", "lower", "upper"))
+    expect_named(ci, expected = c("smooth", "type", "by", "x", "z", var_nms))
 })
 
 test_that("Simultaneous interval for a 2d smooth works", {
@@ -66,8 +65,7 @@ test_that("Simultaneous interval for a 2d smooth works", {
             nsim = 100))
     expect_s3_class(ci, "confint.gam")
     expect_s3_class(ci, "tbl_df")
-    expect_named(ci, expected = c("smooth", "type", "by", "x", "z", "est",
-                                  "se", "crit", "lower", "upper"))
[…]

### 2023-10-13 tests/testthat/test-family-utils.R

@@ -649,7 +649,7 @@ test_that("extract_link() works on gammals() family objects", {
     expect_type(f, "closure")
     expect_identical(f(val), fam$linfo[[2L]]$linkfun(val))
     expect_identical(f, fam$linfo[[2L]]$linkfun)
-    f <- extract_link(fam, parameter = "theta")
+    f <- extract_link(fam, parameter = "phi")
     expect_type(f, "closure")
     expect_identical(f(val), fam$linfo[[2L]]$linkfun(val))
     expect_identical(f, fam$linfo[[2L]]$linkfun)
@@ -658,7 +658,7 @@ test_that("extract_link() works on gammals() family objects", {
     expect_type(f, "closure")
     expect_identical(f(val), fam$linfo[[2L]]$linkinv(val))
     expect_identical(f, fam$linfo[[2L]]$linkinv)
-    f <- extract_link(fam, parameter = "theta", inverse = TRUE)
+    f <- extract_link(fam, parameter = "phi", inverse = TRUE)
     expect_type(f, "closure")
     expect_identical(f(val), fam$linfo[[2L]]$linkinv(val))
     expect_identical(f, fam$linfo[[2L]]$linkinv)

### 2023-10-13 tests/testthat/test-partial-derivatives.R

@@ -7,3 +7,11 @@ test_that("partial derivatives works with single data point", {
     expect_silent(pd_x <- partial_derivatives(su_m_bivar_te, data = ds,
         type = "central", focal = "x"))
 })
+
+# partial derivatives
+test_that("partial derivatives works with single data point sim interval", {
+    # data slice through te(x,z) holding z == 0.4
+    ds <- data_slice(su_m_bivar_te, x = 0.0204313075, z = 0.4)
+    expect_silent(pd_x <- partial_derivatives(su_m_bivar_te, data = ds,
+        type = "central", focal = "x", interval = "simultaneous", seed = 2))
+})

### 2023-10-13 tests/testthat/test-posterior-samples.R

@@ -1,4 +1,4 @@
-## Test posterior sampling functions
+# Test posterior sampling functions
 
 test_that("smooth_samples works for a continuous by GAM", {
     expect_silent(sm <- smooth_samples(su_m_cont_by, n = 5, n_vals = 100,

### 2023-10-13 tests/testthat/test-utilities.R

@@ -373,7 +373,7 @@ test_that("transform_fun works for parametric_effects", {
     envir = teardown_env()),
                    "Interaction terms are not currently supported.")
     expect_silent(pe <- transform_fun(pe, fun = abs))
-    expect_true(all(!pe$partial < 0L))
+    expect_true(all(!pe$.partial < 0L))
 })
 
 test_that("transform_fun works for evaluated_smooth", {

### 2024-01-03 R/add-methods.R
Deploying to gh-pages from @ gavinsimpson/gratia@9576be559e1c037bee022a463551ceace283ccb1 🚀

@@ -20,10 +20,6 @@
 #'
 #' @param type character; the type of predictions to return. See
 #'   [mgcv::predict.gam()] for options.
-#' @param prefix character; string to prepend to names of predicted values when
-#'   `type` is `"terms"`, `"iterms"`, `"lpmatrix"`. These prediction types
-#'   result in a matrix of values being returned. `prefix` will be prepended to
-#'   each of the names of columns returned by such prediction types.
 #' @param ... additional arguments passed to [mgcv::predict.gam()].
 #'
 #' @return A data frame (tibble) formed from `data` and predictions from
@@ -54,8 +50,8 @@
 #' ## with type = "terms" or "iterms"
 #' add_fitted(df, m, type = "terms")
 #' \dontshow{options(op)}
-`add_fitted.gam` <- function(data, model, value = ".value", type = "response",
-                             prefix = ".", ...) {
+`add_fitted.gam` <- function(data, model, value = ".fitted", type = "response",
+                             ...) {
     ## coerce to tibble
     data <- as_tibble(data)
 
@@ -70,11 +66,10 @@
     ## having pruned off any standard errors, process the result
     if (is.array(pred_vals) && length(dim(pred_vals)) > 1L) {
         pred_vals <- as_tibble(pred_vals)
-        pred_vals <- set_names(pred_vals, nm = paste0(prefix, names(pred_vals)))
         if (type %in% c("terms", "iterms")) {
             pred_vals <-
                 add_column(pred_vals,
-                           !!(paste0(prefix, "constant")) := coef(model)[1],
+                           !!(".constant") := coef(model)[1],
                            .before = 1L)
         }
         data <- bind_cols(data, pred_vals)
@@ -181,7 +176,7 @@
 #' }
 #' df <- data_sim("eg1", seed = 1)
 #' df <- df[, c("y","x0","x1","x2","x3")]
-#' m <-  gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, method = 'REML')
+#' m <-  gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, method = "REML")
 #'
 #' ## add partial residuals
 #' add_partial_residuals(df, m)
@@ -262,11 +257,11 @@
     if (!is.null(constant)) {
         if (!is.numeric(constant)) {
             stop("'constant' must be numeric: supplied <", constant, ">",
-                 call. = FALSE)
+                call. = FALSE)
         }
[…]

### 2024-01-03 R/compare-smooths.R

@@ -62,23 +62,23 @@
 
     ## loop over list of smooth estimates and add model column
     for (i in seq_along(sm_est)) {
-        sm_est[[i]] <- add_column(sm_est[[i]], model = model_names[i],
+        sm_est[[i]] <- add_column(sm_est[[i]], .model = model_names[i],
                                   .before = 1L)
     }
 
     `unnest_nest` <- function(x) {
         x %>%
-            group_by(.data$smooth) %>%
+            group_by(.data$.smooth) %>%
             group_split() %>%
             purrr::map(unnest, cols = all_of("data")) %>%
-            purrr::map(nest, data = !all_of(c("model", "smooth",
-                                              "type", "by"))) %>%
+            purrr::map(nest, data = !all_of(c(".model", ".smooth",
+                                              ".type", ".by"))) %>%
             bind_rows()
     }
     sm_est <- purrr::map(sm_est, unnest_nest)
 
     sm_est <- bind_rows(sm_est) %>%
-        arrange(.data$smooth)
+        arrange(.data$.smooth)
     class(sm_est) <- c("compare_smooths", class(sm_est))
     sm_est
 }
@@ -97,7 +97,7 @@
                                    ncol = NULL, nrow = NULL,
                                    guides = "collect",
                                    ...) {
-    l <- group_split(object, .data$smooth)
+    l <- group_split(object, .data$.smooth)
 
     plts <- map(l, plot_comparison_of_smooths)
 
@@ -116,7 +116,7 @@
 #' @importFrom ggplot2 ggplot geom_ribbon geom_line
 `plot_comparison_of_smooths` <- function(object, coverage = 0.95, ...) {
     ## get the covariate labels
-    sm_vars <- vars_from_label(unique(object[["smooth"]]))
+    sm_vars <- vars_from_label(unique(object[[".smooth"]]))
     ## unnest data cols
     object <- unnest(object, cols = all_of("data"))
 
@@ -129,20 +129,20 @@
     ## basic plot
     plt <- ggplot(object, aes(x = .data[[sm_vars[1L]]],
                               y = .data[[".estimate"]],
-                              group = .data[["model"]]))
+                              group = .data[[".model"]]))
 
     ## add uncertainty bands
     plt <- plt + geom_ribbon(aes(ymin = .data[[".lower_ci"]],
                                  ymax = .data[[".upper_ci"]],
-                                 fill = .data[["model"]]),
+                                 fill = .data[[".model"]]),
[…]

### 2024-01-03 R/draw-gam.R

@@ -303,7 +303,7 @@
                 }
                 # merge with the evaluated smooth
                 sm_eval <- suppress_matches_multiple_warning(
-                  left_join(sm_eval, p_resids, by = "smooth")
+                  left_join(sm_eval, p_resids, by = ".smooth")
                 )
             }
         }
@@ -315,7 +315,7 @@
 
             # merge with the evaluated smooth
             sm_eval <- suppress_matches_multiple_warning(
-              left_join(sm_eval, rug_data, by = "smooth")
+              left_join(sm_eval, rug_data, by = ".smooth")
             )
         }
 
@@ -326,16 +326,16 @@
 
         # draw smooths
         sm_l <- if (isTRUE(grouped_by)) {
-            levs <- unique(str_split_fixed(sm_eval$smooth, ":", n = 2)[, 1])
+            levs <- unique(str_split_fixed(sm_eval$.smooth, ":", n = 2)[, 1])
             sm_eval |>
-                mutate(smooth = factor(.data$smooth, levels = S[select]),
-                .term = str_split_fixed(.data$smooth, ":", n = 2)[, 1]) |>
-                arrange(.data$smooth) |>
+                mutate(smooth = factor(.data$.smooth, levels = S[select]),
+                .term = str_split_fixed(.data$.smooth, ":", n = 2)[, 1]) |>
+                arrange(.data$.smooth) |>
                 relocate(".term", .before = 1L) |>
-                group_split(factor(.data$.term, levels = levs), .data$by)
+                group_split(factor(.data$.term, levels = levs), .data$.by)
         } else {
             # the factor is to reorder to way the smooths entered the model
-            group_split(sm_eval, factor(.data$smooth, levels = S[select]))
+            group_split(sm_eval, factor(.data$.smooth, levels = S[select]))
         }
         sm_plts <- map(sm_l,
             draw_smooth_estimates,

### 2024-01-03 R/nested-functions.R

@@ -33,7 +33,7 @@
 
     ## tidy partial residuals to long format
     p_resids <- pivot_longer(p_resids, everything(),
-                             names_to = "smooth",
+                             names_to = ".smooth",
                              values_to = "partial_residual")
 
     # nest this long format so we 1 row per smooth, and partial_residual is now
@@ -56,7 +56,7 @@
     p_resids <- p_resids %>%
       rowwise() %>%
       mutate(partial_residual =
-        list(get_obs_data_for_smooth(.data$smooth,
+        list(get_obs_data_for_smooth(.data$.smooth,
                                      object, data,
                                      extra = .data$partial_residual))) %>%
       ungroup()
@@ -93,10 +93,10 @@
         terms <- smooths(object)
     }
 
-    rug_vals <- tibble(smooth = terms) %>%
-        rowwise() %>%
+    rug_vals <- tibble(.smooth = terms) |>
+        rowwise() |>
         mutate(rug_data =
-                 list(get_obs_data_for_smooth(.data$smooth, object, data))) %>%
+                 list(get_obs_data_for_smooth(.data$.smooth, object, data))) |>
         ungroup()
 
     rug_vals

### 2024-01-03 R/smooth-estimates.R

@@ -319,7 +319,7 @@
     data <- select(data, all_of(keep_vars))
 
     ## Return object
-    tbl <- tibble(smooth = rep(label, nrow(X)), .estimate = fit, .se = se.fit)
+    tbl <- tibble(.smooth = rep(label, nrow(X)), .estimate = fit, .se = se.fit)
     ## bind on the data
     tbl <- bind_cols(tbl, data)
     ## nest all columns with varying data
@@ -386,7 +386,7 @@
     data <- select(data, all_of(keep_vars))
 
     ## Return object
-    tbl <- tibble(smooth = rep(label, nrow(data)), .estimate = fit,
+    tbl <- tibble(.smooth = rep(label, nrow(data)), .estimate = fit,
         .se = se_fit)
     ## bind on the data
     tbl <- bind_cols(tbl, data)
@@ -430,36 +430,33 @@
 
     ## deal with data if supplied
     data <- process_user_data_for_eval(data = data, model = model,
-                                       n = n, n_3d = n_3d, n_4d = n_4d,
-                                       id = which_smooth(model,
-                                                         smooth_label(smooth)))
+        n = n, n_3d = n_3d, n_4d = n_4d,
+        id = which_smooth(model,
+            smooth_label(smooth)))
 
     ## values of spline at data
     eval_sm <- spline_values2(smooth, data = data,
-                              unconditional = unconditional,
-                              model = model,
-                              overall_uncertainty = overall_uncertainty)
+        unconditional = unconditional,
+        model = model,
+        overall_uncertainty = overall_uncertainty)
 
     ## add on info regarding by variable
-    nr <- nrow(eval_sm)
-    eval_sm <- add_column(eval_sm, by = rep(by_var, nr),
-                          .after = 1L)
+    eval_sm <- add_by_var_column(eval_sm, by_var = by_var)
     ## add on spline type info
-    sm_type <- smooth_type(smooth)
-    eval_sm <- add_column(eval_sm, type = rep(sm_type, nr),
-                          .after = 1L)
+    eval_sm <- add_smooth_type_column(eval_sm, sm_type = smooth_type(smooth))
 
     # set some values to NA if too far from the data
     if (smooth_dim(smooth) == 2L && (!is.null(dist) && dist > 0)) {
         eval_sm <- too_far_to_na(smooth,
-                                 input = eval_sm,
-                                 reference = model[["model"]],
[…]

### 2024-01-03 R/utililties.R

@@ -1592,3 +1592,13 @@ reclass_scam_smooth <- function(smooth) {
     se <- sqrt(pmax(0, rowSums((X %*% vc) * X)))
     list(betas = beta_z, se = se)
 }
+
+#' @importFrom tibble add_column
+`add_by_var_column` <- function(object, by_var) {
+    add_column(object, .by = rep(by_var, nrow(object)), .after = 1L)
+}
+
+#' @importFrom tibble add_column
+`add_smooth_type_column` <- function(object, sm_type) {
+    add_column(object, .type = rep(sm_type, nrow(object)), .after = 1L)
+}

### 2024-01-03 tests/testthat/test-add-functions.R

@@ -1,90 +1,81 @@
 ## Test basis() and related functions
 
-## take only some columns
-data <- su_eg1[, c("y", "x0", "x1", "x2", "x3")]
-## fit the model
-#m <-  gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = su_eg1, method = 'REML')
+## data object to use in tests
+add_data <- su_eg1[, c("y", "x0", "x1", "x2", "x3")]
 
 test_that("add_fitted works for a GAM", {
-    expect_silent(df <- add_fitted(data, m_gam))
+    expect_silent(df <- add_fitted(add_data, m_gam))
     expect_s3_class(df, "tbl_df")
-    expect_named(df, c("y", "x0", "x1", "x2", "x3", ".value"))
+    expect_named(df, c("y", "x0", "x1", "x2", "x3", ".fitted"))
 })
 
 test_that("add_fitted works for a GAM with type = 'terms'", {
-    expect_silent(df <- add_fitted(data, m_gam, type = 'terms'))
+    expect_silent(df <- add_fitted(add_data, m_gam, type = 'terms'))
     expect_s3_class(df, "tbl_df")
     expect_named(df, c("y", "x0", "x1", "x2", "x3", ".constant",
-                       ".s(x0)", ".s(x1)", ".s(x2)", ".s(x3)"))
+                       "s(x0)", "s(x1)", "s(x2)", "s(x3)"))
 })
 
 test_that("add_fitted works for a GAM with se.fit = TRUE", {
-    expect_silent(df <- add_fitted(data, m_gam, se.fit = TRUE))
+    expect_silent(df <- add_fitted(add_data, m_gam, se.fit = TRUE))
     expect_s3_class(df, "tbl_df")
-    expect_named(df, c("y", "x0", "x1", "x2", "x3", ".value"))
-})
-
-test_that("prefix works for a GAM with type = 'terms'", {
-    expect_silent(df <- add_fitted(data, m_gam, type = 'terms', prefix = ".."))
-    expect_s3_class(df, "tbl_df")
-    expect_named(df, c("y", "x0", "x1", "x2", "x3", "..constant",
-                       "..s(x0)", "..s(x1)", "..s(x2)", "..s(x3)"))
+    expect_named(df, c("y", "x0", "x1", "x2", "x3", ".fitted"))
 })
 
 test_that("add_residuals works for a GAM", {
[…]

### 2024-01-03 tests/testthat/test-compare-smooths.R

@@ -8,7 +8,7 @@
 test_that("compare_smooths() works for a pair of GAMs", {
     expect_silent(cs <- compare_smooths(su_m_quick_eg1,
                                         su_m_quick_eg1_shrink))
-    expect_named(cs, expected = c("model", "smooth", "type", "by", "data"))
+    expect_named(cs, expected = c(".model", ".smooth", ".type", ".by", "data"))
     expect_s3_class(cs, c("tbl_df", "tbl", "data.frame"))
     expect_identical(nrow(cs), 8L) # 4 smooths x 2 models
 })
@@ -35,12 +35,12 @@ test_that("draw.compare_smooths() can plot a comparison of smooths", {
 
 test_that("compare_smooths() works when a model is not a simple object", {
     expect_silent(cs <- compare_smooths(m_gam, m_gamm$gam))
-    expect_named(cs, expected = c("model", "smooth", "type", "by", "data"))
+    expect_named(cs, expected = c(".model", ".smooth", ".type", ".by", "data"))
     expect_s3_class(cs, c("tbl_df", "tbl", "data.frame"))
     expect_identical(nrow(cs), 8L) # 4 smooths x 2 models
 
     expect_silent(cs <- compare_smooths(m_gam, m_gamm[["gam"]]))
-    expect_named(cs, expected = c("model", "smooth", "type", "by", "data"))
+    expect_named(cs, expected = c(".model", ".smooth", ".type", ".by", "data"))
     expect_s3_class(cs, c("tbl_df", "tbl", "data.frame"))
     expect_identical(nrow(cs), 8L) # 4 smooths x 2 models
 })
@@ -50,17 +50,17 @@ test_that("compare_smooths() works when a model is not a simple object", {
     l2 <- list(model2 = m_gamm$gam)
 
     expect_silent(cs <- compare_smooths(l1[[1]], l2[[1]]))
-    expect_named(cs, expected = c("model", "smooth", "type", "by", "data"))
+    expect_named(cs, expected = c(".model", ".smooth", ".type", ".by", "data"))
     expect_s3_class(cs, c("tbl_df", "tbl", "data.frame"))
     expect_identical(nrow(cs), 8L) # 4 smooths x 2 models
 
     expect_silent(cs <- compare_smooths(m_gam, l2[["model2"]]))
-    expect_named(cs, expected = c("model", "smooth", "type", "by", "data"))
[…]

### 2024-01-03 tests/testthat/test-confint-methods.R

@@ -34,7 +34,7 @@ test_that("Point-wise confidence interval for a GAM works", {
     ci <- confint(m_gam, parm = "s(x1)", type = "confidence")
     expect_s3_class(ci, "confint.gam")
     expect_s3_class(ci, "tbl_df")
-    expect_named(ci, expected = c("smooth", "type", "by", "x1", var_nms))
+    expect_named(ci, expected = c(".smooth", ".type", ".by", "x1", var_nms))
 })
 
 test_that("Simultaneous interval for a GAM works", {
@@ -44,7 +44,7 @@ test_that("Simultaneous interval for a GAM works", {
         confint(m_gam, parm = "s(x1)", type = "simultaneous", nsim = 100))
     expect_s3_class(ci, "confint.gam")
     expect_s3_class(ci, "tbl_df")
-    expect_named(ci, expected = c("smooth", "type", "by", "x1", var_nms))
+    expect_named(ci, expected = c(".smooth", ".type", ".by", "x1", var_nms))
 })
 
 ## 2d smooth
@@ -54,7 +54,7 @@ test_that("Point-wise confidence interval for a 2d smooth works", {
     ci <- confint(su_m_bivar_te, parm = "te(x,z)", type = "confidence")
     expect_s3_class(ci, "confint.gam")
     expect_s3_class(ci, "tbl_df")
-    expect_named(ci, expected = c("smooth", "type", "by", "x", "z", var_nms))
+    expect_named(ci, expected = c(".smooth", ".type", ".by", "x", "z", var_nms))
 })
 
 test_that("Simultaneous interval for a 2d smooth works", {
@@ -65,7 +65,7 @@ test_that("Simultaneous interval for a 2d smooth works", {
             nsim = 100))
     expect_s3_class(ci, "confint.gam")
     expect_s3_class(ci, "tbl_df")
-    expect_named(ci, expected = c("smooth", "type", "by", "x", "z", var_nms))
+    expect_named(ci, expected = c(".smooth", ".type", ".by", "x", "z", var_nms))
 })
 
 test_that("Point-wise confidence interval for a GAMM works", {
@@ -74,7 +74,7 @@ test_that("Point-wise confidence interval for a GAMM works", {
     ci <- confint(m_gamm, parm = "s(x1)", type = "confidence")
     expect_s3_class(ci, "confint.gam")
     expect_s3_class(ci, "tbl_df")
-    expect_named(ci, expected = c("smooth", "type", "by", "x1", var_nms))
[…]

### 2024-01-03 tests/testthat/test-derivatives.R

@@ -373,26 +373,27 @@ test_that("derivatives() works for factor by smooths issue 47", {
                        "derivative","se","crit","lower","upper"))
     plt1 <- draw(d)
 
-    m <- gam(y ~ x1 + s(x2) + fac + s(x0, by = fac), data = su_eg4,
+    m_47 <- gam(y ~ x1 + s(x2) + fac + s(x0, by = fac), data = su_eg4,
              method = "REML")
-    expect_silent(d <- derivatives(m))
+    expect_silent(d <- derivatives(m_47))
     expect_s3_class(d, "derivatives")
     expect_s3_class(d, "tbl_df")
     expect_named(d, c("smooth", "var", "by_var", "fs_var", "fac", "data",
                       "derivative", "se", "crit", "lower", "upper"))
     plt2 <- draw(d)
 
-    dat <- transform(su_eg4, ofac = ordered(fac))
-    m <- gam(y ~ x1 + s(x2) + ofac + s(x0) + s(x0, by = ofac), data = dat,
+    dat_47 <- transform(su_eg4, ofac = ordered(fac))
+    m_47 <- gam(y ~ x1 + s(x2) + ofac + s(x0) + s(x0, by = ofac), data = dat_47,
              method = "REML")
-    expect_silent(d <- derivatives(m))
+    expect_silent(d <- derivatives(m_47))
     expect_s3_class(d, "derivatives")
     expect_s3_class(d, "tbl_df")
     expect_named(d, c("smooth", "var", "by_var", "fs_var", "ofac", "data",
                       "derivative", "se", "crit", "lower", "upper"))
     plt3 <- draw(d)
-    m <- gamm(y ~ x1 + s(x2) + fac + s(x0, by = fac), data = su_eg4)
-    expect_silent(d <- derivatives(m))
+    
+    m_47 <- gamm(y ~ x1 + s(x2) + fac + s(x0, by = fac), data = su_eg4)
+    expect_silent(d <- derivatives(m_47))
     expect_s3_class(d, "derivatives")
     expect_s3_class(d, "tbl_df")
     expect_named(d, c("smooth", "var", "by_var", "fs_var", "fac", "data",

### 2024-01-03 tests/testthat/test-residuals.R

@@ -1,60 +1,67 @@
 ## Test partial_residuals() and related residuals functions
 
-N <- 400L
-df <- data_sim("eg1", n = N, seed = 42)
+N <- 1000L # 400L
+# df <- data_sim("eg1", n = N, seed = 42)
 ## fit the model
-m       <-  gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, method = 'REML')
-m_bam   <-  bam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, method = 'fREML')
-m_gamm  <-  gamm(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df)
-m_gamm4 <-  gamm4(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df)
+# m       <-  gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, method = 'REML')
+# m_bam   <-  bam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, method = 'fREML')
+# m_gamm  <-  gamm(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df)
+# m_gamm4 <-  gamm4(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df)
 
 test_that("partial_residuals returns a tibble", {
-    expect_silent(p_res <- partial_residuals(m))
-    expect_s3_class(p_res, class = c("tbl_df", "tbl", "data.frame"), exact = TRUE)
+    expect_silent(p_res <- partial_residuals(m_gam))
+    expect_s3_class(p_res, class = c("tbl_df", "tbl", "data.frame"),
+        exact = TRUE)
     expect_named(p_res, c("s(x0)", "s(x1)", "s(x2)", "s(x3)"))
     expect_identical(nrow(p_res), N)
 })
 
 test_that("partial_residuals returns a tibble", {
     expect_silent(p_res <- partial_residuals(m_bam))
-    expect_s3_class(p_res, class = c("tbl_df", "tbl", "data.frame"), exact = TRUE)
+    expect_s3_class(p_res, class = c("tbl_df", "tbl", "data.frame"),
+        exact = TRUE)
     expect_named(p_res, c("s(x0)", "s(x1)", "s(x2)", "s(x3)"))
     expect_identical(nrow(p_res), N)
 })
 
 test_that("partial_residuals returns a tibble", {
[…]

### 2024-02-03 R/by-variable-utilities.R
Deploying to gh-pages from @ gavinsimpson/gratia@7b350c299927a14c0d3e707ff9ead324faa1cc01 🚀

@@ -55,7 +55,7 @@
         .after = 1L)
 }
 
-## !! Only affects evaluate_smooth()
+## !! Only affects evaluate_smooth() which was removed in 0.9.0
 #' @importFrom tibble add_column
 `add_by_var_info_to_smooth` <- function(smooth, by_name, by_data, n) {
     nc <- NCOL(smooth)

### 2024-02-03 R/datagen-methods.R

@@ -1,7 +1,12 @@
 #' Generate data over the range of variables used in smooths
 #'
+#' @description
+#'
+#' `r lifecycle::badge('deprecated')`
+#'
 #' For each smooth in a GAM, generate new data over the range of the variables
-#' involved in a smooth.
+#' involved in a smooth. This function is deprecated as it is only useful for a
+#' very narrow use-case. Use [data_slice()] instead.
 #'
 #' @param x an object for which new data is required. Currently objects of
 #'   classes `"gam"`, and `"gamm"` are supported, as are smooths from **mgcv**
@@ -20,100 +25,96 @@
 #'
 #' @keywords internal
 `datagen` <- function(x, ...) {
-    UseMethod("datagen")
+  UseMethod("datagen")
 }
 
 #' @export
 #' @rdname datagen
 `datagen.mgcv.smooth` <- function(x, n = 100, data, ...) {
-    d <- smooth_dim(x)                 # how many dimensions in smooth
-    term <- smooth_terms(x)            # what term are we dealing with
+  lifecycle::deprecate_warn("0.9.0", "datagen()", "data_slice()")
+  d <- smooth_dim(x)                 # how many dimensions in smooth
+  term <- smooth_terms(x)            # what term are we dealing with
+  ## some smooths can't be plotted, esp n-d ones where n > 2
+  if (!x$plot.me || d > 2L) {
+    out <- data.frame()  # FIXME: or should we throw error/message
+  }
+  if (d == 1L) {                      # 1-d smooths
+    xvals <- data[[term]]
+    newvals <- seq_min_max(xvals, n = n)
+    out <- data.frame(smooth = rep(smooth_label(x), n), x = newvals)
+  } else if (d == 2L) {                            # 2-d smooths
+    xvals <- data[[term[1]]]
+    zvals <- data[[term[2]]]
+    newx <- seq_min_max(xvals, n = n)
+    newz <- seq_min_max(zvals, n = n)
+    out <- expand.grid(x1 = newx, x2 = newz)
+    out <- cbind(smooth = rep(smooth_label(x), n^2), out)
+  } else {
+    stop("Cannot handle smooths of three (3) or more terms.")
+  }
 
-    ## some smooths can't be plotted, esp n-d ones where n > 2
-    if (!x$plot.me || d > 2L) {
-        out <- data.frame()  # FIXME: or should we throw error/message
-    }
-
[…]

### 2024-02-03 R/draw-evaluate-smooth.R

@@ -1,11 +1,13 @@
-#' Plot estimated smooths
+#' Plot estimated parametric effects
+#'
+#' @description
+#' `r lifecycle::badge("deprecated")`
 #'
 #' Plots estimated univariate and bivariate smooths using ggplot2.
 #'
-#' @param object an object, the result of a call to [evaluate_smooth()].
-#' @param rug For `evaluate_smooth()`, a numeric vector of values for the
-#'   location of data on the x axis. The default of `NULL` results in no
-#'   rug plot being drawn. For `evaluate_parametric_terms()`, a logical to
+#' @param object an object, the result of a call to
+#'   [evaluate_parametric_term()].
+#' @param rug For `evaluate_parametric_terms()`, a logical to
 #'   indicate if a rug plot should be drawn.
 #' @param ci_level numeric between 0 and 1; the coverage of credible interval.
 #' @param constant numeric; a constant to add to the estimated values of the
@@ -15,9 +17,6 @@
 #'   and confidence interval before plotting. Can be a function or the name of a
 #'   function. Function `fun` will be applied after adding any `constant`, if
 #'   provided.
-#' @param partial_residuals data frame; partial residuals and data values if
-#'   partial residuals are drawn. Should have names `..p_resid` and `..orig_x`
-#'   if supplied.
 #' @param xlab character or expression; the label for the x axis. If not
 #'   supplied, a suitable label will be generated from `object`.
 #' @param ylab character or expression; the label for the y axis. If not
@@ -31,383 +30,21 @@
 #' @param response_range numeric; a vector of two values giving the range of
 #'   response data for the guide. Used to fix plots to a common scale/range.
 #'   Ignored if `show` is set to `"se"`.
+#' @param position Position adjustment, either as a string, or the result of a
+#'   call to a position adjustment function.
 #' @param ... arguments passed to other methods.
 #'
 #' @return A [ggplot2::ggplot()] object.
 #'
 #' @author Gavin L. Simpson
 #'
-#' @importFrom ggplot2 ggplot aes labs geom_line geom_ribbon
-#'   expand_limits
+#' @importFrom ggplot2 ggplot aes labs geom_line geom_ribbon geom_pointrange
+#'   geom_rug expand_limits
 #' @importFrom rlang .data
 #' @importFrom grid unit
 #'
 #' @export
-#' @name draw.evaluated_smooth
-#' @aliases draw.evaluated_1d_smooth draw.evaluated_2d_smooth geom_rug
-#'
-#' @examples
[…]

### 2024-02-03 R/evaluate_smooth.R

@@ -17,29 +17,8 @@
 #' function call to [smooth_estimates()]. However, there are some differences:
 #' 
 #' * the `newdata` argument becomes `data`
-#' 
-#' Consider `evaluate_smooth()` to be *soft*-deprecated; its use is discouraged
-#' and it may be removed at a later date if it becomes difficult to maintain
-#' the current functionality, but there are no intentions of removing it from
-#' gratia unless that situation arises.
 #'
 #' @param object an object of class `"gam"` or `"gamm"`.
-#' @param smooth character; a single smooth to evaluate.
-#' @param n numeric; the number of points over the range of the covariate at
-#'   which to evaluate the smooth.
-#' @param newdata a vector or data frame of points at which to evaluate the
-#'   smooth.
-#' @param unconditional logical; should confidence intervals include the
-#'   uncertainty due to smoothness selection? If `TRUE`, the corrected Bayesian
-#'   covariance matrix will be used.
-#' @param overall_uncertainty logical; should the uncertainty in the model
-#'  constant term be included in the standard error of the evaluate values of
-#'  the smooth?
-#' @param dist numeric; if greater than 0, this is used to determine when
-#'   a location is too far from data to be plotted when plotting 2-D smooths.
-#'   The data are scaled into the unit square before deciding what to exclude,
-#'   and `dist` is a distance within the unit square. See
-#'   [mgcv::exclude.too.far()] for further details.
 #' @param ... arguments passed to other methods.
 #'
 #' @return A data frame, which is of class `"evaluated_1d_smooth"` or
@@ -50,519 +29,14 @@
 #' @importFrom stats setNames
 #'
 #' @export
-#'
-#' @examples
-#' load_mgcv()
-#' \dontshow{
-#' op <- options(cli.unicode = FALSE, pillar.sigfig = 2)
-#' }
-#' dat <- data_sim("eg1", n = 500, dist = "normal", scale = 1, seed = 2)
-#' m1 <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = dat, method = "REML")
-#'
-#' evaluate_smooth(m1, "s(x1)")
-#'
-#' ## 2d example
-#' dat <- data_sim("eg2", n = 1000, dist = "normal", scale = 1, seed = 2)
-#' m2 <- gam(y ~ s(x, z, k = 30), data = dat, method = "REML")
-#'
[…]

### 2024-02-03 R/gratia-package.R

@@ -1,2 +1,6 @@
-#' @keywords internal 
+#' @keywords internal
 "_PACKAGE"
+
+## usethis namespace: start
+#' @importFrom lifecycle deprecated
+## usethis namespace: end

### 2024-02-03 R/other-scam-methods.R

@@ -1,39 +1,45 @@
 #' @export
-`vcov.scam` <- function (object, freq = FALSE, dispersion = NULL,
-                         parametrized = TRUE, ...)  {
-    if (freq) {
-        vc <- if (parametrized) {
-            object$Ve.t
-        } else {
-            object$Ve
-        }
+`vcov.scam` <- function(object, freq = FALSE, dispersion = NULL,
+  parametrized = TRUE, ...)  {
+  if (freq) {
+    vc <- if (parametrized) {
+      object$Ve.t
     } else {
-        vc <- if (parametrized) {
-            object$Vp.t
-        } else {
-            object$Vp
-        }
+      object$Ve
     }
-    if (!is.null(dispersion)) {
-        vc <- dispersion * vc/object$sig2
+  } else {
+    vc <- if (parametrized) {
+      object$Vp.t
+    } else {
+      object$Vp
     }
-    name <- names(object$edf)
-    dimnames(vc) <- list(name, name)
-    vc
+  }
+  if (!is.null(dispersion)) {
+    vc <- dispersion * vc / object$sig2
+  }
+  name <- names(object$edf)
+  dimnames(vc) <- list(name, name)
+  vc
 }
 
 #' Extract coefficients from a fitted `scam` model.
 #'
 #' @param object a model object fitted by `scam()`
-#' @param parametrized logical; extract parametrized coefficients, which respect the linear inequality constraints of the model.
+#' @param parametrized logical; extract parametrized coefficients, which
+#'   respect the linear inequality constraints of the model.
 #' @param ... other arguments.
 #'
 #' @export
 `coef.scam` <- function(object, parametrized = TRUE, ...) {
-    coefs <- if (parametrized) {
-        object$coefficients.t
-    } else {
-        object$coefficients
-    }
-    coefs
+  coefs <- if (parametrized) {
+    object$coefficients.t
+  } else {
+    object$coefficients
+  }
+  if (!is.null(nms <- dimnames(coefs))) {
+    dimnames(coefs) <- NULL
+    coefs <- as.numeric(coefs)
+    names(coefs) <- unlist(nms)
+  }
+  coefs
 }

### 2024-02-03 R/posterior-samples.R

@@ -451,182 +451,136 @@
 #' draw(sms)
 #' \dontshow{options(op)}
 `smooth_samples` <- function(model, ...) {
-    UseMethod("smooth_samples")
+  UseMethod("smooth_samples")
 }
 
 #' @export
 `smooth_samples.default` <- function(model, ...) {
-    stop("Don't know how to sample from the posterior of <",
-         class(model)[[1L]], ">", .call = FALSE)
+  stop("Don't know how to sample from the posterior of <",
+    class(model)[[1L]], ">", .call = FALSE)
 }
 
 #' @export
 `smooth_samples.scam` <- function(model, term = NULL, n = 1, data = newdata,
-    method = c("gaussian", "mh", "inla", "user"),
-    seed = NULL,
-    freq = FALSE,
-    unconditional = FALSE,
-    n_cores = 1L,
-    n_vals = 200,
-    burnin = 1000,
-    thin = 1,
-    t_df = 40,
-    rw_scale = 0.25,
-    rng_per_smooth = FALSE,
-    draws = NULL,
-    ...,
-    newdata = NULL,
-    ncores = NULL) {
-    # smooth_samples begins
-    if (!exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)) {
-        runif(1)
+  method = c("gaussian", "mh", "inla", "user"), seed = NULL,
+  freq = FALSE, unconditional = FALSE, n_cores = 1L, n_vals = 200,
+  burnin = 1000, thin = 1, t_df = 40, rw_scale = 0.25, rng_per_smooth = FALSE,
+  draws = NULL, mvn_method = c("mvnfast", "mgcv"), ...,
+  newdata = NULL, ncores = NULL) {
+  # smooth_samples begins
+  if (!exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)) {
+    runif(1)
+  }
+  if (is.null(seed)) {
+    RNGstate <- get(".Random.seed", envir = .GlobalEnv)
+  } else {
+    R.seed <- get(".Random.seed", envir = .GlobalEnv)
+    set.seed(seed)
+    RNGstate <- structure(seed, kind = as.list(RNGkind()))
+    on.exit(assign(".Random.seed", R.seed, envir = .GlobalEnv))
+  }
+  S <- smooths(model)             # vector of smooth labels - "s(x)"
+  take <- seq_along(S)            # in default case 1,2,3,..,n smooths
+  if (!is.null(term)) {
[…]

### 2024-02-03 R/smooth-estimates.R

@@ -3,8 +3,8 @@
 #' Evaluate a smooth at a grid of evenly spaced value over the range of the
 #' covariate associated with the smooth. Alternatively, a set of points at which
 #' the smooth should be evaluated can be supplied. `smooth_estimates()` is a new
-#' implementation of [evaluate_smooth()], and should be used instead of that
-#' other function.
+#' implementation of `evaluate_smooth()`, and replaces that function, which has
+#' been removed from the package.
 #'
 #' @param object an object of class `"gam"` or `"gamm"`.
 #' @param smooth character; a single smooth to evaluate.

### 2024-02-03 R/utililties.R

@@ -1,22 +1,36 @@
-## smooth_terms should be removed
+# smooth_terms should be removed
+#' List the variables involved in smooths
+#'
+#' @description
+#' `r lifecycle::badge("experimental")`
+#'
+#' @param object an R object the result of a call to [mgcv::gam()],
+#' [mgcv::bam()], or [mgcv::gamm()], or that inherits from classes `"gam"` or
+#' `"mgcv.smooth"`, or `"fs.interaction"`.
+#' @param ... arguments passed to other methods. Currently unused.
+#' @export
 `smooth_terms` <- function(object, ...) {
-    UseMethod("smooth_terms")
+  UseMethod("smooth_terms")
 }
 
+#' @export
 `smooth_terms.gam` <- function(object, ...) {
-    lapply(object[["smooth"]], `[[`, "term")
+  lapply(object[["smooth"]], `[[`, "term")
 }
 
+#' @export
 `smooth_terms.gamm` <- function(object, ...) {
-    smooth_terms(object[["gam"]], ...)
+  smooth_terms(object[["gam"]], ...)
 }
 
+#' @export
 `smooth_terms.mgcv.smooth` <- function(object, ...) {
-    object[["term"]]
+  object[["term"]]
 }
 
+#' @export
 `smooth_terms.fs.interaction` <- function(object, ...) {
-    object[["term"]]
+  object[["term"]]
 }
 
 #' Dimension of a smooth
@@ -1079,24 +1093,6 @@ vars_from_label <- function(label) {
 #' @author Gavin L. Simpson
 `transform_fun` <- function(object, fun = NULL , ...) {
     UseMethod("transform_fun")
-}
-
-#' @rdname transform_fun
-#' @export
-`transform_fun.evaluated_smooth` <- function(object, fun = NULL, ...) {
-    ## If fun supplied, use it to transform est and the upper and lower interval
-    if (!is.null(fun)) {
-        fun <- match.fun(fun)
-        object[["est"]] <- fun(object[["est"]])
-        if (!is.null(object[["upper"]])) {
-            object[["upper"]] <- fun(object[["upper"]])
-        }
-        if (!is.null(object[["lower"]])) {
-            object[["lower"]] <- fun(object[["lower"]])
-        }
-    }
-
-    object
 }
 
 #' @rdname transform_fun

### 2024-02-03 tests/testthat/test-by-variables.R

@@ -14,13 +14,6 @@ test_that("draw() works with continuous by and fixed scales", {
     expect_doppelganger("continuous by-variable smmoth fixed scales", plt)
 })
 
-test_that("evaluate_smooth() works with continuous by", {
-    skip_if_not_installed("withr")
-    withr::local_options(lifecycle_verbosity = "quiet")
-    sm  <- evaluate_smooth(su_m_cont_by, "s(x2)")
-    expect_s3_class(sm, "evaluated_1d_smooth")
-})
-
 test_that("is_by_smooth() is TRUE with continuous by", {
     expect_true(is_by_smooth(su_m_cont_by[["smooth"]][[1]]))
 })

### 2024-02-03 tests/testthat/test-datagen.R

@@ -1,6 +1,15 @@
 ## Test datagen() methods
 
+test_that("datagen is deprecated", {
+  skip_if_not_installed("withr")
+  withr::local_options(digits = 3)
+  expect_snapshot(datagen(get_smooths_by_id(m_gam, 1)[[1L]],
+    data = su_eg1, n = 10))
+})
+
 test_that("datagen works for a simple 1d smooth", {
+    skip_if_not_installed("withr")
+    withr::local_options(lifecycle_verbosity = "quiet")
     expect_silent(dg <- datagen(get_smooths_by_id(m_gam, 1)[[1L]],
                                 data = su_eg1, n = 50))
     expect_s3_class(dg, "data.frame")
@@ -10,6 +19,8 @@ test_that("datagen works for a simple 1d smooth", {
 })
 
 test_that("datagen works for a simple 2d smooth", {
+    skip_if_not_installed("withr")
+    withr::local_options(lifecycle_verbosity = "quiet")
     expect_silent(dg <- datagen(get_smooths_by_id(su_m_bivar, 1)[[1L]],
                                 data = su_eg2, n = 10))
     expect_s3_class(dg, "data.frame")
@@ -19,13 +30,16 @@ test_that("datagen works for a simple 2d smooth", {
 })
 
 test_that("datagen.mgcv.smooth fails for a 3d smooth", {
-    # m <- gam(y ~ s(x0, x1, x2), data = su_eg1, method = "REML")
+    skip_if_not_installed("withr")
+    withr::local_options(lifecycle_verbosity = "quiet")
     expect_error(datagen(su_m_trivar, smooth = "s(x0,x1,x2)"),
                  "Cannot handle smooths of three (3) or more terms.",
                  fixed = TRUE)
 })
 
 test_that("datagen gam method works for a simple 1d smooth", {
+    skip_if_not_installed("withr")
+    withr::local_options(lifecycle_verbosity = "quiet")
     expect_silent(dg <- datagen(m_gam, smooth = "s(x0)", n = 50))
     expect_s3_class(dg, "data.frame")
     expect_named(dg, c("smooth", "x"))
@@ -34,18 +48,24 @@ test_that("datagen gam method works for a simple 1d smooth", {
 })
 
 test_that("datagen gam method fails if smooth not specified", {
+    skip_if_not_installed("withr")
+    withr::local_options(lifecycle_verbosity = "quiet")
     expect_error(datagen(m_gam),
                  "Argument 'smooth' must be specified and not 'NULL'.",
[…]

### 2024-02-03 tests/testthat/test-posterior-samples.R

@@ -51,12 +51,12 @@ test_that("smooth_samples works for a simple GAM MH sampling", {
 })
 
 test_that("smooth_samples works for a multi-smooth GAM", {
-    expect_silent(sm <- smooth_samples(m_gam, n = 5, n_vals = 100, seed = 42))
-    expect_s3_class(sm, c("smooth_samples", "tbl_df",
-                          "tbl", "data.frame"))
-    ## 2000 == 4 smooths * 5 * 100
-    expect_identical(NROW(sm), 2000L)
-    expect_identical(NCOL(sm), 11L) # 11 cols, 4 univatiate smooths
+  expect_silent(sm <- smooth_samples(m_gam, n = 5, n_vals = 100, seed = 42))
+  expect_s3_class(sm, c("smooth_samples", "tbl_df",
+    "tbl", "data.frame"))
+  ## 2000 == 4 smooths * 5 * 100
+  expect_identical(NROW(sm), 2000L)
+  expect_identical(NCOL(sm), 11L) # 11 cols, 4 univatiate smooths
 })
 
 test_that("smooth_samples works for a multi-smooth factor by GAM", {
@@ -70,18 +70,17 @@ test_that("smooth_samples works for a multi-smooth factor by GAM", {
 })
 
 test_that("smooth_samples works for a multi-smooth SCAM", {
-    expect_silent(sm <- smooth_samples(m_scam, n = 5, n_vals = 50, eed = 42))
-    expect_s3_class(sm, c("smooth_samples", "tbl_df",
-                          "tbl", "data.frame"))
-    ## 500 == 2 smooths * 5 * 50
-    expect_identical(NROW(sm), 500L)
-    expect_identical(NCOL(sm), 9L)
+  expect_silent(sm <- smooth_samples(m_scam, n = 5, n_vals = 50, seed = 42))
+  expect_s3_class(sm, c("smooth_samples", "tbl_df", "tbl", "data.frame"))
+  ## 500 == 2 smooths * 5 * 50
+  expect_identical(NROW(sm), 500L)
+  expect_identical(NCOL(sm), 9L)
 })
 
 test_that("smooth_samples() fails if not suitable method available", {
-    expect_error(smooth_samples(1:10),
-                 "Don't know how to sample from the posterior of <integer>",
-                 fixed = TRUE)
+  expect_error(smooth_samples(1:10),
+    "Don't know how to sample from the posterior of <integer>",
+    fixed = TRUE)
 })
 
 test_that("smooth_samples sets seed when seed not provided", {

### 2024-02-03 tests/testthat/test-utilities.R

@@ -1,4 +1,4 @@
-## Test Utilties
+## Test Utilities
 
 test_that("smooth_terms() methods work", {
     st <- smooth_terms(m_gam)
@@ -383,16 +383,13 @@ test_that("term_variables works for a terms", {
 })
 
 test_that("transform_fun works for parametric_effects", {
+    skip_if_not_installed("withr")
+    withr::local_options(lifecycle_verbosity = "quiet")
     expect_message(pe <- parametric_effects(m_para_sm, data = df_2_fac,
     envir = teardown_env()),
                    "Interaction terms are not currently supported.")
     expect_silent(pe <- transform_fun(pe, fun = abs))
     expect_true(all(!pe$.partial < 0L))
-})
-
-test_that("transform_fun works for evaluated_smooth", {
-    expect_warning(sm <- evaluate_smooth(m_gam, smooth = "s(x1)"))
-    expect_silent(sm <- transform_fun(sm, fun = exp))
 })
 
 test_that("transform_fun works for smooth_estimates", {

### 2024-03-06 README.Rmd
style pkg files

@@
 options(tibble.print_min = 5, tibble.print_max = 5)
-library('mgcv')
-library('gratia')
+library("mgcv")
+library("gratia")
 ```
@@
 
-* Estimatation of derivatives of fitted smoothers: `derivatives()`,
+* Estimation of derivatives of fitted smoothers: `derivatives()`,
 
@@
 ```{r install, eval = FALSE}
-install.packages('gratia')
+install.packages("gratia")
 ```
@@
 # Install gratia in R
-install.packages("gratia", repos = c("https://gavinsimpson.r-universe.dev",
-  "https://cloud.r-project.org"))
+install.packages("gratia", repos = c(
+  "https://gavinsimpson.r-universe.dev",
+  "https://cloud.r-project.org"
+))
 ```

### 2024-03-06 tests/testthat/setup.R

@@ -17,33 +17,41 @@ library("ggplot2")
 n_quick <- 300
 quick_eg1 <- data_sim("eg1", n = n_quick, seed = 21)
 quick_eg1_off <- quick_eg1 |> mutate(off = 2)
-su_eg1 <- data_sim("eg1", n = 1000,  dist = "normal", scale = 2, seed = 1)
+su_eg1 <- data_sim("eg1", n = 1000, dist = "normal", scale = 2, seed = 1)
 su_eg2 <- data_sim("eg2", n = 2000, dist = "normal", scale = 0.5, seed = 42)
 su_eg3 <- data_sim("eg3", n = 400, seed = 32)
-su_eg4 <- data_sim("eg4", n = 400,  dist = "normal", scale = 2, seed = 1)
+su_eg4 <- data_sim("eg4", n = 400, dist = "normal", scale = 2, seed = 1)
 
 su_m_quick_eg1 <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3),
-                      data = quick_eg1,
-                      method = "REML")
+  data = quick_eg1,
+  method = "REML"
+)
 
-su_m_quick_eg1_shrink <- gam(y ~ s(x0, bs = "ts") + s(x1, bs = "cs") +
-                               s(x2, bs = "ts") + s(x3, bs = "ts"),
-                             data = quick_eg1,
-                             method = "REML")
+su_m_quick_eg1_shrink <- gam(
+  y ~ s(x0, bs = "ts") + s(x1, bs = "cs") +
+    s(x2, bs = "ts") + s(x3, bs = "ts"),
+  data = quick_eg1,
+  method = "REML"
+)
 
 su_m_univar_4 <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3),
-                     data = su_eg1,
-                     method = "REML")
+  data = su_eg1,
+  method = "REML"
+)
 
-su_m_penalty <- gam(y ~ s(x0, bs = 'cr') + s(x1, bs = 'bs') +
-                      s(x2, k = 15) + s(x3, bs = 'ps'),
-                    data = su_eg1,
-                    method = "REML")
+su_m_penalty <- gam(
+  y ~ s(x0, bs = "cr") + s(x1, bs = "bs") +
+    s(x2, k = 15) + s(x3, bs = "ps"),
+  data = su_eg1,
[…]

### 2024-05-19 tests/testthat/setup.R
update tests and snapshots; add test for #284; update tests because I missed adding a period to the names of objects returned by parametric_effects

@@ -87,6 +87,12 @@ su_m_factor_by <- gam(y ~ fac + s(x2, by = fac) + s(x0),
   data = su_eg4, method = "REML"
 )
 
+# for issue #285
+su_m_factor_by_re <- gam(y ~ s(fac, bs = "re") + s(x2) +
+    s(x2, by = fac, m = 1) + s(x0),
+  data = su_eg4, method = "REML"
+)
+
 su_m_factor_by_gamm <- gamm(y ~ fac + s(x2, by = fac) + s(x0),
   data = su_eg4, REML = TRUE
 )
@@ -576,3 +582,18 @@ soap_data <- soap_fs_data(bnd = soap_fsb)
 m_soap <- gam(y ~
     s(v, w, k = 30, bs = "so", xt = list(bnd = soap_fsb, nmax = 100)),
   data = soap_data, method = "REML", knots = soap_knots)
+
+# Issue 284
+df_284 <- data_sim("eg1", seed = 42)
+df_284 <- df_284 |>
+  mutate(
+    month = factor(
+      rep(
+        month.abb[1:10],
+        times = 40),
+      levels = month.abb[1:10],
+      ordered = TRUE
+    )
+  )
+m_284 <- gam(y ~ month + s(x0) + s(x1) + s(x2) + s(x3),
+  data = df_284, method = "REML")

### 2024-05-19 tests/testthat/test-draw-parametric-effects.R

@@ -204,3 +204,13 @@ test_that("parametric effects works with messing data in model fit", {
   # skip_on_ci() # testing without as moved to mac os x
   expect_doppelganger("issue 219 parametric effects", plt)
 })
+
+test_that("factor parametric effects preserve levels when plotting #284", {
+  expect_silent(pe <- parametric_effects(m_284, data = df_284,
+    envir = teardown_env()))
+
+  skip_on_cran()
+  expect_silent(plt <- draw(pe))
+
+  expect_doppelganger("draw peff preserves factor levels", plt)
+})

### 2024-05-19 tests/testthat/test-parametric-effects.R

@@ -6,7 +6,8 @@ both_nms <- c(".level", ".value", ".partial", ".se")
 
 test_that("parametric_effects works for m_2_fac", {
   expect_message(
-    peff <- parametric_effects(m_2_fac, envir = teardown_env(), data = df_2_fac),
+    peff <- parametric_effects(m_2_fac, envir = teardown_env(),
+    data = df_2_fac),
     "Interaction terms are not currently supported."
   )
   expect_s3_class(peff, class = c(
@@ -15,12 +16,13 @@ test_that("parametric_effects works for m_2_fac", {
   ))
   expect_identical(ncol(peff), 5L)
   expect_identical(nrow(peff), 7L)
-  expect_named(peff, c("term", "type", lev_nms))
+  expect_named(peff, c(".term", ".type", lev_nms))
 })
 
 test_that("parametric_effects works for m_para_sm", {
   expect_message(
-    peff <- parametric_effects(m_para_sm, data = df_2_fac, envir = teardown_env()),
+    peff <- parametric_effects(m_para_sm, data = df_2_fac,
+    envir = teardown_env()),
     "Interaction terms are not currently supported."
   )
   expect_s3_class(peff, class = c(
@@ -29,7 +31,7 @@ test_that("parametric_effects works for m_para_sm", {
   ))
   expect_identical(ncol(peff), 6L)
   expect_identical(nrow(peff), 407L)
-  expect_named(peff, c("term", "type", both_nms))
+  expect_named(peff, c(".term", ".type", both_nms))
 })
 
 test_that("parametric_effects works for m_2_fac select term", {
@@ -43,7 +45,7 @@ test_that("parametric_effects works for m_2_fac select term", {
   ))
   expect_identical(ncol(peff), 5L)
   expect_identical(nrow(peff), 3L)
-  expect_named(peff, c("term", "type", lev_nms))
+  expect_named(peff, c(".term", ".type", lev_nms))
 })
 
 test_that("parametric_effects works for m_para_sm select term", {
@@ -57,7 +59,7 @@ test_that("parametric_effects works for m_para_sm select term", {
   ))
   expect_identical(ncol(peff), 5L)
   expect_identical(nrow(peff), 3L)
-  expect_named(peff, c("term", "type", lev_nms))
+  expect_named(peff, c(".term", ".type", lev_nms))
 })
 
 test_that("parametric_effects works with only parametric terms", {
[…]

### 2024-06-27 paper.Rmd
add draft joss paper on gratia

---
title: 'gratia: An R package for working with generalized additive models'
tags:
  - R
  - splines
  - generalized additive models
  - posterior simulation
authors:
  - name: Gavin L. Simpson
    orcid: 0000-0002-9084-8413
    affiliation: 1
affiliations:
 - name: Department of Animal and Veterinary Science, Aarhus University, Denmark
   index: 1
date: 26 June 2024
citation_author: Simpson
year: 2024
bibliography: paper.bib
output:
  rticles::joss_article:
    keep_md: true
csl: apa.csl
journal: JOSS
---

# Summary

Generalized additive models [GAMs, @Hastie1990-bx; @Wood2017-qi] are an extension of the generalized linear model that allows the effects of covariates to be modelled as smooth functions. GAMs are increasingly used in many areas of science [e.g. @Pedersen2019-ff; @Simpson2018-wc] because the smooth functions allow nonlinear relationships between covariates and the response to be learned from the data through the use of penalized splines. Within the R [@rcore2024] ecosystem, Simon Wood's  *mgcv* package [@Wood2017-qi] is widely used to fit GAMs and is a *Recommended* package that ships with R as part of the default install. A growing number of other R packages build upon *mgcv*, for example as an engine to fit specialised models not handled by *mgcv* itself [e.g. *GJMR*, @Marra2023-gjrm], or to make use of the wide range of splines available in *mgcv* [e.g. *brms*, @Burkner2017-ms].

The *gratia* package builds upon *mgcv* by providing functions that make working with GAMs easier. *gratia* takes a *tidy* approach [@Wickham2014-ev] providing *ggplot2* [@Wickham2016-dg] replacements for *mgcv*'s base graphics-based plots, functions for model diagnostics and exploration of fitted models, and a family of functions for drawing samples from the posterior distribution of a fitted GAM. Additional functionality is provided to facilitate the teaching and understanding of GAMs.

# Generalized additive models

A GAM has the form
\begin{align*}
y_i &\sim    \mathcal{D}(\mu_i, \phi) \\
g(\mu_i) &=  \mathbf{A}_i\boldsymbol{\gamma} + \sum_{j=1} f_j(x_{ji})
\end{align*}
where observations $y_i$ are assumed to be conditionally distributed $\mathcal{D}$ with expectation $\mathbb{E}(y_i) = \mu_i$ and dispersion $\phi$. The expectation of $y_i$ is given by a linear predictor of strictly parametric terms, whose model matrix is $\mathbf{A}_i$ with parameters $\boldsymbol{\gamma}$, plus a sum of smooth functions of $j = 1, \dots, J$ covariates $f_j()$. $g()$ is a link function mapping values on the linear predictor to the scale of the response.

The smooth functions $f_j$ are represented in the GAM using penalised splines, which are themselves formed as weighted sums of basis functions, $b_k()$, [@De-Boor2001-vg] e.g.
$$
f_j(x_{ij}) = \sum_{k=1}^{K} \beta_{jk} b_{jk}(x_{ij})
$$
for a univariate spline. The weights, $\beta_k$, are model coefficients to be estimated alongside $\boldsymbol{\gamma}$. To avoid overfitting, estimates $\hat{\beta}_{jk}$ and $\hat{\boldsymbol{\gamma}}$ are sought to minimise the penalised log-likelihood of the model
$$
\mathcal{L}(\boldsymbol{\beta}) = \ell(\boldsymbol{\beta}) - \frac{1}{2\phi} \sum_{j} \lambda_{j} \boldsymbol{\beta}^{\mathsf{T}}_j \mathbf{S}_j \boldsymbol{\beta}_j
$$
where $\ell$ is the log likelihood of the data at the parameter estimates, $\mathbf{S}_j$ are penalty matrices and $\lambda_{j}$ are smoothing parameters associated with each smooth. Note that $\boldsymbol{\beta}$ now contains the coefficients $\boldsymbol{\gamma}$ and $\beta_{jk}$. $\boldsymbol{\beta}^{\mathsf{T}}_j \mathbf{S}_j \boldsymbol{\beta}_j$ measures the wiggliness of $f_j$, which, with the default penalty, is the integrated squared second derivative of $f_j$. The smoothing parameters, $\boldsymbol{\lambda}$, control the trade-off between fit to the data and the complexity of the estimated functions.

The default spline created by *mgcv*'s `s()` is a low rank, thin plate regression spline [@Wood2003-qy]. Figure \ref{fig:basis-funs}, shows the basis functions for such a spline fitted to data simulated from the function
$$
f = 0.2x^{11}\{10(1 - x)\}^6 + 10(10x)^3(1 - x)^{10} \label{gwf2}
$$
with additive Gaussian noise ($\mu = 0, \sigma = 1$), and the associated penalty matrix, prepared using functions from *gratia*.

```{r penalised-spline-basis-and-penalty, echo = FALSE, fig.height = 4, fig.width = 10, fig.cap = "\\label{fig:basis-funs}Basis functions (a) and associated penalty matrix (b) for a penalised, low rank, thin plate regression spline. a) shows the individual basis functions (thin coloured lines), as well as the data (black points) to which the GAM was fitted. The estimated smooth is shown as the thick grey line. b) shows the penalty matrix for the basis shown in a). Note the 9th basis function (labelled 'F9', which is the linear function at the lower left to upper right in a), is not affected by the penalty as it has 0 second derivative everywhere, and hence the resulting penalty for this function is 0."}
pkgs <- c("mgcv", "gratia", "ggplot2", "dplyr", "tibble", "patchwork")
res <- vapply(pkgs, library, logical(1), character.only = TRUE,
  logical.return = TRUE)
# simulate data
sim_d <- tibble(x = withr::with_seed(1234, runif(100))) |>
  mutate(
    f = gw_f2(x),
    y = withr::with_seed(1234, rnorm(100, mean = f, sd = 1))
  )
# fit model
sim_m <- gam(y ~ s(x, bs = "tp"), data = sim_d, method = "REML",
  drop.intercept = FALSE)
# basis functions for s(x)
bfs <- basis(sim_m, select = "s(x)")
# penalty S
f_x <- bfs |>
  group_by(x) |>
  summarise(.spline = sum(.value), .groups = "drop")
# recentre on the data
bfs <- bfs |>
  mutate(.value = .value + model_constant(sim_m))
f_x <- f_x |>
  mutate(.spline = .spline + model_constant(sim_m))
# plot basis functions and fitted spline
p_bfs <- bfs |>
  ggplot(aes(x = x, y = .value)) +
  geom_line(aes(colour = .bf)) +
  geom_point(data = sim_d, aes(x = x, y = y)) +
  geom_line(data = f_x, aes(x = x, y = .spline),
    linewidth = 1.25, alpha = 0.5) +
  guides(colour = "none") +
  labs(x = "x", y = "y", title = "s(x)")
# plot the penalty
S <- penalty(sim_m, select = "s(x)")
p_S <- S |>
  draw()
p_bfs + p_S + plot_layout(ncol = 2, nrow = 1) +
  plot_annotation(tag_levels = "a", tag_suffix = ")")
```

# Statement of need

*mgcv* is state-of-the-art software for fitting GAMs and their extensions to data sets on the order of millions of observations [e.g. @Li2020-ch; @Wood2011-kn; @Wood2016-fx]. *mgcv* provides functions for plotting estimated smooth functions, as well as for producing model diagnostic plots. These functions produce plots using base graphics, the original plotting system for R. Additionally, *mgcv* returns fitted GAMs as complex list objects (see `?mgcv::gamObject`), the contents of which are not easily used for downstream analysis without careful study of *mgcv* and its help pages, plus a good understanding of GAMs themselves. One of the motivations driving the development of *gratia* was to provide equivalent plotting capabilities using the *ggplot2* package [@Wickham2016-dg]. To facilitate this, *gratia* provides functions for representing the model components as objects using *tidy* principles, which are suitable for plotting with *ggplot2* or manipulation by packages in the *tidyverse* [e.g. @Wickham2023-uj]. This functionality allows for high-level plotting using the `draw()` method, as well as easily customisable plot creation using lower-level functionality.

Taking a Bayesian approach to smoothing with penalized splines [@Kimeldorf1970-cn; @Wahba1983-mi; @Silverman1985-kw; @Wahba1985-bw; see @Miller2019-nf for a summary], it can be shown that GAMs fitted by *mgcv* are an empirical Bayesian model with an improper multivariate normal prior on the basis function coeficients. Samples from the posterior distribution of these models can be used to estimate the uncertainty in quantities derived from a GAM. This can be invaluable in applied research, where, for example, a quantity of interest may arise as an operation on predictions from the model. *gratia* provides functions for sampling from the posterior distribution of estimated smooths and from the model as a whole, where sampling can include the uncertainty in the estimated coefficients (`fitted_samples()`), the sampling uncertainty of the response (`predicted_samples()`), or both (`posterior_samples()`). By default, a Gaussian approximation to the posterior distribution is used, but a simple Metropolis Hasting sampler can be substituted (using `mgcv::gam.mh()`), which has better performance when the posterior is not well approximated by a Gaussian approximation.

The teaching of GAMs can benefit from visualisation of the spline basis functions and associated penalty matrices. *gratia* provides this functionality via `basis()` and `penalty()`, which can be applied either to a smooth specification (e.g. `s(x, z, bs = "ds")`) or to a fitted GAM (see Figure \ref{fig:basis-funs}). These functions expose functionality already available in *mgcv*, but supply outputs in a tidy format, which makes access to these features more intuitive than the original implementations in *mgcv*. Additional utility functions are provided, for example: `model_constant()`, `edf()`, `model_edf()`, `overview()`, and `inv_link()`, which extract the model intercept term (or terms), the effective degrees of freedom of individual smooths and the overall model, shows a summary of the fitted GAM, and extracts the inverse of the link function(s) used, respectively.

The overall aim of *gratia* is to abstract away some of the complexity of working with GAMs fitted using *mgcv* to allow researchers to focus on using and interrogating their model rather than the technical R programming needed to achieve this. As a result, *gratia* is increasingly being used by researchers in many fields, and has, at the time of writing, been cited over 200 times (data from Google Scholar).

# Example usage

In this short example, I illustrate a few of the features of *gratia* using a data set of sea surface chlorophyll *a* measurements at a number of locations in the Atlantic Ocean, whose spatial locations are given as geographical coordinates (`lat` and `lon`), plus two additional covariates; `bathy`, the depth of the ocean, in metres, at the sampling location, and `jul.day`, the day of the year in which the observation was made. These data are in the `chl` dataset provided by the *gamair* package accompanying @Wood2017-qi.

The packages required for this example are loaded, as is the data set `chl` with
```{r load-packages, results = "hide", cache = FALSE}
pkgs <- c("mgcv", "gamair", "gratia", "ggplot2", "dplyr", "ggdist")
loaded <- vapply(pkgs, library, logical(1L), logical.return = TRUE,
  character.only = TRUE)
data(chl, package = "gamair")
```

A simple GAM for these data is to model the response (`chl`) with a spatial smooth of latitude (`lat`) and longitude (`lon`) as covariates. Here, I use a spline on the sphere (SOS) smoother built using a Duchon spline with second order derivative penalty [@Duchon1977-jr]. Additional terms included in the linear predictor are a smooth of the day of year of sample collection (`jul.day`) and a smooth of ocean depth (`bath`). The response is assumed to be conditionally distributed Tweedie, with the power parameter ($p$) of the distribution estimated during fitting. Model coefficients and smoothing parameters are estimated using restricted maximum likelihood [@Wood2011-kn]
```{r fit-chl-model, cache = TRUE}
ctrl <- gam.control(nthreads = 10)
m1 <- gam(
  chl ~ s(lat, lon, bs = "sos", m = -1, k = 150) +
    s(jul.day, bs = "cr", k = 20) +
    s(bath, k = 10),
  data = chl, method = "REML", control = ctrl, family = tw()
)
```

Model diagnostic plots can be produced using `appraise()`, which by default produces four plots: i) a QQ plot of model residuals, with theoretical quantiles and reference bands generated following @Augustin2012-sc, ii) a plot of residuals (deviance residuals are the default) against linear predictor values, iii) a histogram of residuals, and iv) a plot of observed versus fitted values. Model diagnostic plots for the model, with simulated residuals-based reference bands on the QQ plot, are produced with
```{r m1-appraise, fig.show = "hide", dependson = "fit-chl-model", cache = FALSE, fig.height = 7, fig.width = 10}
appraise(m1, method = "simulate")
```
which show significant heteroscedasticity and departure from the condtional distribution of the response given the model (Figure \ref{fig:m1-appraise}).

```{r m1-appraise, echo = FALSE, fig.height = 7, fig.width = 10, fig.cap = "\\label{fig:m1-appraise}Model diagnostic plots for the GAM fitted to the ocean chlorophyll *a* data produced by the `appraise()` function. The four plots produced are: i) a QQ plot of model residuals, with theoretical quantiles and reference bands generated following @Augustin2012-sc (upper left), ii) a plot of residuals (deviance residuals are default) against linear predictor values (upper right), iii) a histogram of deviance residuals (lower left), and iv) a plot of observed versus fitted values (lower right)"}
```

The problems with the model aparent in the diagnostics plots are probably due to important controls on chlorophyll *a* missing from the covariates available in the example data. However, the original model assumed constant values for the scale, $\varphi$, and the power parameter $p$, which may be too inflexible given the absence of important effects in the model. A distributional GAM, where linear predictors for all distributional parameters, may improve the model diagnostics.

A distributional GAM for $\mathcal{D}$ Tweedie, with linear predictors for $\mu$, $p$, and $\varphi$ is fitted below using *mgcv*'s `twlss()` family
```{r fit-chl-lss-model, cache = TRUE}
m2 <- gam(
  list(
    chl ~ s(lat, lon, bs = "sos", m = -1, k = 150) + # location
      s(jul.day, bs = "cr", k = 20) +
      s(bath, k = 10),
    ~ s(lat, lon, bs = "sos", m = -1, k = 100) +     # power
      s(jul.day, bs = "cr", k = 20) +
      s(bath, k = 10),
    ~ s(lat, lon, bs = "sos", m = -1, k = 100) +     # scale
      s(jul.day, bs = "cr", k = 20) +
      s(bath, k = 10)),
  data = chl, method = "REML", control = ctrl, family = twlss()
)
```

This model has much better model diagnostics although some large residuals remain (Figure \ref{fig:m2-appraise}). Note that the QQ plot uses theoretical quantiles from a standard normal distribution as the simulation-based values are not currently available in *mgcv* or *gratia* for some of the distributional families, including the `twlss()` family, and as such, the reference bands may not be appropriate.

```{r m2-appraise, echo = FALSE, fig.height = 7, fig.width = 10, fig.cap = "\\label{fig:m2-appraise}Model diagnostic plots for the distributional GAM fitted to the ocean chlorophyll *a* data produced by the `appraise()` function. Refer to the caption for Figure \\ref{fig:m1-appraise} for a description of the plots shown.", dependson = "fit-chl-lsslmodel", cache = FALSE}
appraise(m2)
```

*gratia* can handle distributional GAMs fitted with *mgcv* and also *GJRM*'s `gamlss()`. Below, the estimated smooths from `m2` are plotted using `draw()`
```{r m2-draw, fig.show = "hide", dependson = "fit-chl-lsslmodel", cache = TRUE, fig.pos = "t!", out.width = ".8\\linewidth"}
crs <- "+proj=ortho +lat_0=20 +lon_0=-40"
draw(m2, crs = crs, default_crs = 4326, dist = 0.05, rug = FALSE)
```
Here, we see a specialised plot drawn for spline-on-the-sphere smooths $f(\mathtt{lat}_i,\mathtt{lon}_i)$ (Figure \ref{fig:m2-draw}), which uses `ggplot2::coord_sf()` and functionality from the *sf* package [@Pebesma2018-ws; @Pebesma2023-fe] to visualise the smooth via an orthographic projection.

```{r m2-draw, echo = FALSE, fig.height = 12, fig.width = 15, fig.cap = "\\label{fig:m2-draw}Estimated smooth functions for the distributional GAM, `m2`, fitted to the ocean chlorophyll *a* data. The first row of plots is for the linear predictor of the conditional mean chlorophyll *a*, while the second and third rows are for the conditional power parameter and conditional scale, respectively. The shaded ribbons are 95% Bayesian credible intervals."}
```

If the provided plots are insufficient for users' needs, lower-level functionality is provided by *gratia* to facilitate bespoke plotting with *ggplot2*. For example, to evaluate the SOS smooth at a grid (50x50) of values over the range of the covariates, we use `smooth_estimates()` and add a Bayesian credible interval with `add_confint()`:
```{r sm-est-m1, results = "hide", dependson = "fit-chl-lsslmodel", cache = TRUE}
smooth_estimates(m2, select = "s(lat,lon)", n = 50) |>
  add_confint()
```
This returns a data frame of the requested values, which is easily plotted using `ggplot()`.

## Posterior sampling

Perhaps we are interested in the average expected chlorophyll *a* between 40--50 degrees N and 40--50 degrees W. It would be quite a simple matter to calculate this value from the fitted model: we first create a slice through the data for the spatial locations were are interested in using the `data_slice()` function, which ensures that `ds` contains everything we need to predict from the fitted model
```{r avg-chl-ds, dependson = "fit-chl-lsslmodel", cache = TRUE}
ds <- data_slice(m2,
  lat = evenly(lat, lower = 40, upper = 50, by = 0.5),
  lon = evenly(lon, lower = -50, upper = -40, by = 0.5)
)
```
Next, `fitted_values()` returns the predicted values at the specified locations. I only include the spatial effects, excluding the effects of ocean depth and day of year:
```{r avg-chl-fv, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
use <- c("(Intercept)", "s(lat,lon)") # , "s.1(lat,lon)", "s.2(lat,lon)")
fv <- fitted_values(m2, data = ds, terms = use) # predict
```
Finally, I summarise the predictions for the location parameter to yield the average of the predicted values
```{r avg-chl-sumamrise, dependson = "avg-chl-fv", cache = TRUE}
fv |>
  filter(.parameter == "location") |>
  summarise(chl_a = mean(.fitted))
```
While this is an acceptable answer to the question, it lacks an uncertainty estimate. This is where posterior sampling is useful. With a small modification of the above code and a little data wrangling, we can produce an uncertainty estimate ,using `fitted_samples()` to generate posterior draws of the expected chlorophyll *a*:
```{r post-sampling-1, warning = FALSE, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
fs <- fitted_samples(m2,     # model
  data = ds,                 # values of covariates to predict at
  terms = use,               # which terms to include in predictions
  n = 10000,                 # number of posterior draws
  method = "gaussian",       # Gaussian approximation to the posterior
  unconditional = TRUE,      # incl uncertainty for estimating lambda
  n_cores = 4,               # how many CPU cores to compute MVN samples
  seed = 342)                # set the random number seed, used internally
```
The posterior draws can then be summarised as before, except now the average chlorophyll *a* is calculated separately for each posterior draw (`.draw`)
```{r summarise-posterior, dependson = "post-sampling-1", cache = TRUE}
fs |>                                 # take the posterior draws
  group_by(.draw) |>                  # group them by `.draw`
  summarise(chl_a = mean(.fitted)) |> # compute mean of fitted chl a
  ggdist::median_qi()                 # summarise posterior
```
The posterior distribution of average chlorophyll *a* is summarised using `median_qi()` from the *ggdist* package [@Kay2024-rv; @Kay2024-uj]. While it would be a simple matter to compute the interval with base R commands, the use of `median_qi()` illustrates how *gratia* tries to interact with other packages.

# References

### 2024-06-27 paper.Rmd
build the paper in this branch of the repo; trigger rebuild of pdf

@@
 b) shows the penalty matrix for the basis shown in a).
-Note the 9th basis function (labelled 'F9', which is the linear function at the lower left to upper right in a), is not affected by the penalty as it has 0 second derivative everywhere, and hence the resulting penalty for this function is 0."}
+Note the 9th basis function (labelled 'F9', which is the linear function at the lower left to upper right in a), is not affected by the penalty as it has 0 second derivative everywhere, and hence the resulting penalty for this function is 0.", fig.pos = "t!"}
 pkgs <- c("mgcv", "gratia", "ggplot2", "dplyr", "tibble", "patchwork")
@@
 Model diagnostic plots for the model, with simulated residuals-based reference bands on the QQ plot, are produced with
-```{r m1-appraise, fig.show = "hide", dependson = "fit-chl-model", cache = FALSE, fig.height = 7, fig.width = 10}
+```{r m1-appraise, fig.show = "hide", dependson = "fit-chl-model", cache = FALSE, fig.height = 7, fig.width = 10, fig.pos = "t!"}
 appraise(m1, method = "simulate")
@@
 ```{r m1-appraise, echo = FALSE, fig.height = 7, fig.width = 10, fig.cap = "\\label{fig:m1-appraise}Model diagnostic plots for the GAM fitted to the ocean chlorophyll *a* data produced by the `appraise()` function.
-The four plots produced are: i) a QQ plot of model residuals, with theoretical quantiles and reference bands generated following @Augustin2012-sc (upper left), ii) a plot of residuals (deviance residuals are default) against linear predictor values (upper right), iii) a histogram of deviance residuals (lower left), and iv) a plot of observed versus fitted values (lower right)"}
+The four plots produced are: i) a QQ plot of model residuals, with theoretical quantiles and reference bands generated following @Augustin2012-sc (upper left), ii) a plot of residuals (deviance residuals are default) against linear predictor values (upper right), iii) a histogram of deviance residuals (lower left), and iv) a plot of observed versus fitted values (lower right)", fig.pos = "t!"}
 ```
@@
 ```{r m2-appraise, echo = FALSE, fig.height = 7, fig.width = 10, fig.cap = "\\label{fig:m2-appraise}Model diagnostic plots for the distributional GAM fitted to the ocean chlorophyll *a* data produced by the `appraise()` function.
-Refer to the caption for Figure \\ref{fig:m1-appraise} for a description of the plots shown.", dependson = "fit-chl-lsslmodel", cache = FALSE}
+Refer to the caption for Figure \\ref{fig:m1-appraise} for a description of the plots shown.", dependson = "fit-chl-lsslmodel", cache = FALSE, fig.pos = "t!"}
 appraise(m2)
@@
   lat = evenly(lat, lower = 40, upper = 50, by = 0.5),
-  lon = evenly(lon, lower = -50, upper = -40, by = 0.5)
-)
+  lon = evenly(lon, lower = -50, upper = -40, by = 0.5))
 ```

### 2024-06-27 paper.Rmd
change title

@@
 ---
-title: 'gratia: An R package for working with generalized additive models'
+title: 'gratia: An R package for exploring generalized additive models'
 tags:
@@
 Model diagnostic plots for the model, with simulated residuals-based reference bands on the QQ plot, are produced with
+
 ```{r m1-appraise, fig.show = "hide", dependson = "fit-chl-model", cache = FALSE, fig.height = 7, fig.width = 10, fig.pos = "t!"}
@@
 ```
+
 which show significant heteroscedasticity and departure from the condtional distribution of the response given the model (Figure \ref{fig:m1-appraise}).

### 2024-06-28 paper.Rmd
probably best to use all the terms even though the distributional etas don't affect the expectation - they would if this was posterior_samples()

@@
 ```{r avg-chl-fv, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
-use <- c("(Intercept)", "s(lat,lon)") # , "s.1(lat,lon)", "s.2(lat,lon)")
+use <- c("(Intercept)", "s(lat,lon)", "s.1(lat,lon)", "s.2(lat,lon)")
 fv <- fitted_values(m2, data = ds, terms = use) # predict

### 2024-06-28 paper.Rmd
in hindsight, simplifying the code is best; added a note in the text now to say I'm also excluding terms in the other linear predictors.

@@
 ```
-Next, `fitted_values()` returns the predicted values at the specified locations. I only include the spatial effects, excluding the effects of ocean depth and day of year:
+Next, `fitted_values()` returns the predicted values at the specified locations.
+I only include the spatial effects, excluding the effects of ocean depth and day of year, and ignore terms in the other linear predictors as they do not affect the expected chlorophyll *a*:
 ```{r avg-chl-fv, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
-use <- c("(Intercept)", "s(lat,lon)", "s.1(lat,lon)", "s.2(lat,lon)")
+use <- c("(Intercpt)", "s(lat,lon)")
 fv <- fitted_values(m2, data = ds, terms = use) # predict

### 2024-07-17 tests/testthat/test-draw-methods.R
update snaps and tests

@@ -525,7 +525,7 @@ test_that("draw() can handle a mixture of numeric and factor random effects", {
 test_that("draw.gam uses fixed scales if asked for them: #73", {
   skip_on_cran()
   # skip_on_ci() # testing without as moved to mac os x
-  df <- data_sim("eg1", n = 1000, seed = 1)
+  # df <- data_sim("eg1", n = 1000, seed = 1)
   m <- gam(y ~ s(x1) + s(x2) + ti(x1, x2), data = su_eg1, method = "REML")
   plt <- draw(m, scales = "fixed", rug = FALSE)
 

### 2024-07-17 tests/testthat/test-hgam-paper.R

@@ -138,7 +138,7 @@ test_that("draw() can plot bird_move model 1", {
 test_that("draw() can plot bird_move model 2", {
   skip_on_cran()
   skip_on_ci()
-  expect_warning(
+  # expect_warning(
     bird_mod2 <- bam(
       count ~ te(week, latitude,
         bs = c("cc", "tp"),
@@ -151,9 +151,9 @@ test_that("draw() can plot bird_move model 2", {
       data = bird_move, method = "fREML", family = poisson(),
       knots = list(week = c(0, 52)),
       control = ctrl, discrete = FALSE
-    ),
-    "fitted rates numerically 0 occurred"
-  )
+    )#,
+    #"fitted rates numerically 0 occurred"
+  #)
   plt <- draw(bird_mod2, rug = FALSE, n = 50)
   expect_doppelganger("hgam-paper-bird-move-model-2", plt)
 })

### 2024-08-05 README.Rmd
try to stop alt text being printed
indentation
indentation again
more indentation
even more indetation
fig_caption isn't an argument to github format
rebuild
The readme could benefit from a better statement of need in the overview section; #305

@@
 
-Graceful 'ggplot'-based graphics and utility functions for working with
-generalized additive models (GAMs) fitted using the 'mgcv' package. Provides a
+Working with GAMs within the 'tidyverse' can be tedious and even difficult
+without a good understanding of GAMs themselves and how the model is returned
+by 'mgcv' and what the model objects contain. 'gratia' is designed to help with
+this.
+
+'gratia' provides 'ggplot'-based graphics and utility functions for working with
+generalized additive models (GAMs) fitted using the 'mgcv' package, via a
 reimplementation of the `plot()` method for GAMs that 'mgcv' provides, as well
@@
 
-```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide", dependson = -2}
+```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide"}
 appraise(m1)

### 2024-08-05 paper.Rmd
fix accessibility point: #304

@@
 ```
-This returns a data frame of the requested values, which is easily plotted using `ggplot()`.
+This returns a data frame of the requested values that is ammenable for plotted with `ggplot()`.
 

### 2024-08-05 paper.Rmd
fix accessibility point 2: #304

@@
 Perhaps we are interested in the average expected chlorophyll *a* between 40--50 degrees N and 40--50 degrees W.
-It would be quite a simple matter to calculate this value from the fitted model: we first create a slice through the data for the spatial locations were are interested in using the `data_slice()` function, which ensures that `ds` contains everything we need to predict from the fitted model
+An estimate for this value can be computed directly from the fitted model as follows.
+First create a slice through the data for the spatial locations were are interested in using the `data_slice()` function, which ensures that `ds` contains everything we need to predict from the fitted model
 ```{r avg-chl-ds, dependson = "fit-chl-lsslmodel", cache = TRUE}

### 2024-08-05 paper.Rmd
fix accessibility point 3: #304

@@
 The posterior distribution of average chlorophyll *a* is summarised using `median_qi()` from the *ggdist* package [@Kay2024-rv; @Kay2024-uj].
-While it would be a simple matter to compute the interval with base R commands, the use of `median_qi()` illustrates how *gratia* tries to interact with other packages.
+While we could use the base R function `quantile()` to compute the interval, the use of `median_qi()` illustrates how *gratia* tries to interact with other packages.
 

### 2024-08-05 paper.Rmd
fix comma typo; #304

@@
 This is where posterior sampling is useful.
-With a small modification of the above code and a little data wrangling, we can produce an uncertainty estimate ,using `fitted_samples()` to generate posterior draws of the expected chlorophyll *a*:
+With a small modification of the above code and a little data wrangling, we can produce an uncertainty estimate, using `fitted_samples()` to generate posterior draws of the expected chlorophyll *a*:
 ```{r post-sampling-1, warning = FALSE, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}

### 2024-11-21 paper.Rmd
address @vankesteren review comment re urgency; move a statement to the summary #304
move overall motivation statement earleir in statement of need section #304

@@
 Additional functionality is provided to facilitate the teaching and understanding of GAMs.
+The overall aim of *gratia* is to abstract away some of the complexity of working with GAMs fitted using *mgcv* to allow researchers to focus on using and interrogating their model rather than the technical R programming needed to achieve this.
 
@@
 Additionally, *mgcv* returns fitted GAMs as complex list objects (see `?mgcv::gamObject`), the contents of which are not easily used for downstream analysis without careful study of *mgcv* and its help pages, plus a good understanding of GAMs themselves.
+The overall aim of *gratia* is to abstract away some of the complexity of working with GAMs fitted using *mgcv* to allow researchers to focus on using and interrogating their model rather than the technical R programming needed to achieve this.
+As a result, *gratia* is also increasingly being used by researchers in many fields, and has, at the time of writing, been cited over 250 times (data from Google Scholar).
+
 One of the motivations driving the development of *gratia* was to provide equivalent plotting capabilities using the *ggplot2* package [@Wickham2016-dg].
@@
 Additional utility functions are provided, for example: `model_constant()`, `edf()`, `model_edf()`, `overview()`, and `inv_link()`, which extract the model intercept term (or terms), the effective degrees of freedom of individual smooths and the overall model, shows a summary of the fitted GAM, and extracts the inverse of the link function(s) used, respectively.
-
-The overall aim of *gratia* is to abstract away some of the complexity of working with GAMs fitted using *mgcv* to allow researchers to focus on using and interrogating their model rather than the technical R programming needed to achieve this.
-As a result, *gratia* is increasingly being used by researchers in many fields, and has, at the time of writing, been cited over 200 times (data from Google Scholar).
 

### 2024-11-21 paper.Rmd
add comment regarding what the proj4 string is in the code as per #304

@@
 ```{r m2-draw, fig.show = "hide", dependson = "fit-chl-lsslmodel", cache = TRUE, fig.pos = "t!", out.width = ".8\\linewidth"}
+# Define the coordinate system to use for plotting on a map
 crs <- "+proj=ortho +lat_0=20 +lon_0=-40"

### 2024-11-21 paper.Rmd
add a paragraph describing what the plots produced by draw show; #304

--- 2024-11-21 paper.Rmd
@@
 
-*gratia* can handle distributional GAMs fitted with *mgcv* and also *GJRM*'s `gamlss()`. Below, the estimated smooths from `m2` are plotted using `draw()`
+*gratia* can handle distributional GAMs fitted with *mgcv*, and also those fitted using *GJRM*'s `gamlss()`. Below, the estimated smooths from `m2` are plotted using `draw()`
 ```{r m2-draw, fig.show = "hide", dependson = "fit-chl-lsslmodel", cache = TRUE, fig.pos = "t!", out.width = ".8\\linewidth"}
+# Define the coordinate system to use for plotting on a map
 crs <- "+proj=ortho +lat_0=20 +lon_0=-40"
@@
 ```
-Here, we see a specialised plot drawn for spline-on-the-sphere smooths $f(\mathtt{lat}_i,\mathtt{lon}_i)$ (Figure \ref{fig:m2-draw}), which uses `ggplot2::coord_sf()` and functionality from the *sf* package [@Pebesma2018-ws; @Pebesma2023-fe] to visualise the smooth via an orthographic projection.
+The plots produced by `draw()` from a fitted model are known as *partial effect plots* (Figure \ref{fig:m2-draw}), which show the component contributions, on the link scale, of each model term to the linear predictor.
+The y axis on these plots is typically centred around 0 due to most smooths having a sum-to-zero identifiability constraint applied to them.
+This constraint is what allows the model to include multiple smooths and remain identifiable.
+These plots allow you to read off the contributions of each smooth to the fitted response (on the link scale); they show link-scale *predictions* of the response for each smooth, conditional upon all other terms in the model, including any parametric effects and the intercept, having zero contribution.
+In the paralance of the *marginaleffects* package, these plots show *adjusted predictions*, just where the adjustment includes setting the contribution of *all* other model terms to the predicted value to zero.
+For *partial derivatives* (what *marginaleffects* would call a *marginal effect* or slope), *gratia* provides the `derivatives()`.
+
+Here, we see a specialised plot drawn for spline-on-the-sphere smooths $f(\mathtt{lat}_i,\mathtt{lon}_i)$ (Figure \ref{fig:m2-draw}, left-hand column), which uses `ggplot2::coord_sf()` and functionality from the *sf* package [@Pebesma2018-ws; @Pebesma2023-fe] to visualise the smooth via an orthographic projection.
 

### 2024-11-21 paper.Rmd
add a state of the field section as per #304

@@
 
+# State of the field
+
+Several R packages provide similar functionality to that of *gratia*.
+Notably, the *mgcViz* package [@Fasiolo2020-nz] provides sophisticated capabilities for plotting estimated smooths for models fitted by *mgcv* and *qgam* [@], and model diagnostics plots.
+A particular feature of *mgcViz* is that it was designed to be scalable, easily handling models fitted to data with millions of observations, a use case not currently well handled by *gratia*.
+*tidymv* [@tidymv-2023] and its succesor, *tidygam* [@tidygam-2023], provide plots of estimated smooths and model predictions respectively, as well as differences of smooths, but the focus is on univariate smooths, inconstrast to *gratia*'s ability to plot multivariate and specialist smooths (e.g.
+soap film smooths).
+Like *gratia*, the *itsadug* package [@itsadug-2022] is motivated to make working with estimated GAMs and related models easier.
+*itsadug* uses base graphics for plotting, but in place of partial effects plots, produces plots of adjusted predictions, and while useful for a broad range of models, is especially focused on GAMs fitted to longitudinal data with random effects.
+
+Areas where *gratia* stands out from these other packages are i) the consistent functionality for range of posterior simulations (*mgcViz* has some support for simulating new response values) as illustrated below, ii) support for computing derivatives of smooths, partial derivatives, and derivatives on the response scale (i.e.
+conditional derivatives), iii) tools for learning or teaching how GAMs work, iv) the range of utility functions that make working with GAMs easier, and v) the simple and consistent *tidy* interface to all functionality.
+
 # Example usage
@@
 These plots allow you to read off the contributions of each smooth to the fitted response (on the link scale); they show link-scale *predictions* of the response for each smooth, conditional upon all other terms in the model, including any parametric effects and the intercept, having zero contribution.
-In the paralance of the *marginaleffects* package, these plots show *adjusted predictions*, just where the adjustment includes setting the contribution of *all* other model terms to the predicted value to zero.
+In the paralance of the *marginaleffects* package [@Arel-BundockGreiferHeiss:2024], these plots show *adjusted predictions*, just where the adjustment includes setting the contribution of *all* other model terms to the predicted value to zero.
 For *partial derivatives* (what *marginaleffects* would call a *marginal effect* or slope), *gratia* provides the `derivatives()`.

### 2024-11-21 paper.Rmd
add a conclusion as per #304 and an acknowledgments section

@@
 
+# Conclusion
+
+*gratia* provides a range of functionality to make working with estimated GAMs easier for users.
+It is designed to take some of the pain out of working with models, simplifying plotting of smooths and related features (differences, derivatives) and exposing the powerful machinary of the *mgcv* package without the need for a deep understanding of GAMs, splines, and the inner structure of *mgcv*'s model objects.
+As such, it provides a useful addition for anyone wanting to use GAMs in their data analyses without requiring them to be a GAM expert.
+
+# Acknowledgements
+
+Development of *gratia* was supported by a Natural Sciences and Engineering Research Council of Canada (NSERC) Discovery Grant to the author (RGPIN-2014-04032).
+
 # References

### 2024-11-21 paper.Rmd
fix the phi varphi thing and better describe the twlss model #296

@@
 \end{align*}
-where observations $y_i$ are assumed to be conditionally distributed $\mathcal{D}$ with expectation $\mathbb{E}(y_i) = \mu_i$ and dispersion $\phi$.
+where observations $y_i$ are assumed to be conditionally distributed $\mathcal{D}$ with expectation $\mathbb{E}(y_i) = \mu_i$ and dispersion parameter $\phi$.
 The expectation of $y_i$ is given by a linear predictor of strictly parametric terms, whose model matrix is $\mathbf{A}_i$ with parameters $\boldsymbol{\gamma}$, plus a sum of smooth functions of $j = 1, \dots, J$ covariates $f_j()$.
@@
 $$
-\mathcal{L}(\boldsymbol{\beta}) = \ell(\boldsymbol{\beta}) - \frac{1}{2\phi} \sum_{j} \lambda_{j} \boldsymbol{\beta}^{\mathsf{T}}_j \mathbf{S}_j \boldsymbol{\beta}_j
+\ell_p(\boldsymbol{\beta}) = \ell(\boldsymbol{\beta}) - \frac{1}{2\phi} \sum_{j} \lambda_{j} \boldsymbol{\beta}^{\mathsf{T}}_j \mathbf{S}_j \boldsymbol{\beta}_j
 $$
-where $\ell$ is the log likelihood of the data at the parameter estimates, $\mathbf{S}_j$ are penalty matrices and $\lambda_{j}$ are smoothing parameters associated with each smooth.
+where $\ell$ and $\ell_p$ are the log likelihood and penalized log likelihood, respectively, of the data at the parameter estimates, $\mathbf{S}_j$ are penalty matrices and $\lambda_{j}$ are smoothing parameters associated with each smooth.
 Note that $\boldsymbol{\beta}$ now contains the coefficients $\boldsymbol{\gamma}$ and $\beta_{jk}$.
@@
 The problems with the model aparent in the diagnostics plots are probably due to important controls on chlorophyll *a* missing from the covariates available in the example data.
-However, the original model assumed constant values for the scale, $\varphi$, and the power parameter $p$, which may be too inflexible given the absence of important effects in the model.
-A distributional GAM, where linear predictors for all distributional parameters, may improve the model diagnostics.
+However, the original model assumed constant values for the dispersion, $\phi$, and the power parameter $p$, which may be too inflexible given the absence of important effects in the model.
+A distributional GAM, containing linear predictors for all distributional parameters,
+\begin{align*}
+y_i &\sim \mathcal{D}(\mu_i, p_i, \varphi_i) \\
+g(\mu_i) &= \beta_1 + f_1(\text{lat}_i, \text{lon}_i) + f_2(\text{jul.day}_i) + f_3(\text{bath}_i) \\
+g(p_i) &= \beta_2 + f_4(\text{lat}_i, \text{lon}_i) + f_5(\text{jul.day}_i) + f_6(\text{bath}_i) \\
+g(\varphi_i) &= \beta_3 + f_7(\text{lat}_i, \text{lon}_i) + f_8(\text{jul.day}_i) + f_9(\text{bath}_i)
+\end{align*}
+may improve the model diagnostics.
 
-A distributional GAM for $\mathcal{D}$ Tweedie, with linear predictors for $\mu$, $p$, and $\varphi$ is fitted below using *mgcv*'s `twlss()` family
+A distributional GAM for $\mathcal{D}$ Tweedie with linear predictors for $\mu$, $p$, and $\varphi$ is fitted below using *mgcv*'s `twlss()` family
 ```{r fit-chl-lss-model, cache = TRUE}

### 2024-11-21 paper.Rmd
more tweaks to rendering and modifications for #296 re phi or varphi

@@
 A particular feature of *mgcViz* is that it was designed to be scalable, easily handling models fitted to data with millions of observations, a use case not currently well handled by *gratia*.
-*tidymv* [@tidymv-2023] and its succesor, *tidygam* [@tidygam-2023], provide plots of estimated smooths and model predictions respectively, as well as differences of smooths, but the focus is on univariate smooths, inconstrast to *gratia*'s ability to plot multivariate and specialist smooths (e.g.
+*tidymv* [@tidymv-2023] and its succesor, *tidygam* [@tidygam-2023], provide plots of estimated smooths and model predictions respectively, as well as differences of smooths, but the focus is on univariate smooths, in constrast to *gratia*'s ability to plot multivariate and specialist smooths (e.g.
 soap film smooths).
@@
 
-```{r m1-appraise, fig.show = "hide", dependson = "fit-chl-model", cache = FALSE, fig.height = 7, fig.width = 10, fig.pos = "t!"}
+```{r m1-appraise, fig.show = "hide", dependson = "fit-chl-model", cache = FALSE, fig.height = 6.5, fig.width = 10, fig.pos = "t!"}
 appraise(m1, method = "simulate")
@@
 
-```{r m1-appraise, echo = FALSE, fig.height = 7, fig.width = 10, fig.cap = "\\label{fig:m1-appraise}Model diagnostic plots for the GAM fitted to the ocean chlorophyll *a* data produced by the `appraise()` function.
+```{r m1-appraise, echo = FALSE, fig.height = 6, fig.width = 10, fig.cap = "\\label{fig:m1-appraise}Model diagnostic plots for the GAM fitted to the ocean chlorophyll *a* data produced by the `appraise()` function.
 The four plots produced are: i) a QQ plot of model residuals, with theoretical quantiles and reference bands generated following @Augustin2012-sc (upper left), ii) a plot of residuals (deviance residuals are default) against linear predictor values (upper right), iii) a histogram of deviance residuals (lower left), and iv) a plot of observed versus fitted values (lower right)", fig.pos = "t!"}
@@
 However, the original model assumed constant values for the dispersion, $\phi$, and the power parameter $p$, which may be too inflexible given the absence of important effects in the model.
+
 A distributional GAM, containing linear predictors for all distributional parameters,
@@
 
-```{r m2-appraise, echo = FALSE, fig.height = 7, fig.width = 10, fig.cap = "\\label{fig:m2-appraise}Model diagnostic plots for the distributional GAM fitted to the ocean chlorophyll *a* data produced by the `appraise()` function.
+```{r m2-appraise, echo = FALSE, fig.height = 6, fig.width = 10, fig.cap = "\\label{fig:m2-appraise}Model diagnostic plots for the distributional GAM fitted to the ocean chlorophyll *a* data produced by the `appraise()` function.
 Refer to the caption for Figure \\ref{fig:m1-appraise} for a description of the plots shown.", dependson = "fit-chl-lsslmodel", cache = FALSE, fig.pos = "t!"}
@@
 
-```{r m2-draw, echo = FALSE, fig.height = 12, fig.width = 15, fig.cap = "\\label{fig:m2-draw}Estimated smooth functions for the distributional GAM, `m2`, fitted to the ocean chlorophyll *a* data.
+```{r m2-draw, echo = FALSE, fig.height = 10, fig.width = 15, fig.cap = "\\label{fig:m2-draw}Estimated smooth functions for the distributional GAM, `m2`, fitted to the ocean chlorophyll *a* data.
 The first row of plots is for the linear predictor of the conditional mean chlorophyll *a*, while the second and third rows are for the conditional power parameter and conditional scale, respectively.

### 2024-11-21 paper.Rmd
address @dill's comments from #296

@@
 y_i &\sim    \mathcal{D}(\mu_i, \phi) \\
-g(\mu_i) &=  \mathbf{A}_i\boldsymbol{\gamma} + \sum_{j=1} f_j(x_{ji})
+g(\mu_i) &=  \mathbf{A}_i\boldsymbol{\gamma} + f_1(x_{1i}) + f_2(x_{2i}) + f_3(x_{3i}, x_{4i}) + \cdots
 \end{align*}
 where observations $y_i$ are assumed to be conditionally distributed $\mathcal{D}$ with expectation $\mathbb{E}(y_i) = \mu_i$ and dispersion parameter $\phi$.
-The expectation of $y_i$ is given by a linear predictor of strictly parametric terms, whose model matrix is $\mathbf{A}_i$ with parameters $\boldsymbol{\gamma}$, plus a sum of smooth functions of $j = 1, \dots, J$ covariates $f_j()$.
+The expectation of $y_i$ is given by a linear predictor of strictly parametric terms, whose model matrix is $\mathbf{A}_i$ with parameters $\boldsymbol{\gamma}$, plus a sum of $j = 1, \dots, J$ smooth functions of covariates $f_j()$.
 $g()$ is a link function mapping values on the linear predictor to the scale of the response.
@@
 Note that $\boldsymbol{\beta}$ now contains the coefficients $\boldsymbol{\gamma}$ and $\beta_{jk}$.
-$\boldsymbol{\beta}^{\mathsf{T}}_j \mathbf{S}_j \boldsymbol{\beta}_j$ measures the wiggliness of $f_j$, which, with the default penalty, is the integrated squared second derivative of $f_j$.
+$\boldsymbol{\beta}^{\mathsf{T}}_j \mathbf{S}_j \boldsymbol{\beta}_j$ measures the wiggliness of $f_j$, which, with the default basis-penalty combination, is the integrated squared second derivative of $f_j$.
 The smoothing parameters, $\boldsymbol{\lambda}$, control the trade-off between fit to the data and the complexity of the estimated functions.
@@
 ```{r penalised-spline-basis-and-penalty, echo = FALSE, fig.height = 4, fig.width = 10, fig.cap = "\\label{fig:basis-funs}Basis functions (a) and associated penalty matrix (b) for a penalised, low rank, thin plate regression spline.
-a) shows the individual basis functions (thin coloured lines), as well as the data (black points) to which the GAM was fitted.
+a) shows the individual basis functions (thin coloured lines) multiplied by their respective model coefficients, as well as the data (black points) to which the GAM was fitted.
 The estimated smooth is shown as the thick grey line.
@@
 
-Taking a Bayesian approach to smoothing with penalized splines [@Kimeldorf1970-cn; @Wahba1983-mi; @Silverman1985-kw; @Wahba1985-bw; see @Miller2019-nf for a summary], it can be shown that GAMs fitted by *mgcv* are an empirical Bayesian model with an improper multivariate normal prior on the basis function coeficients.
+Taking a Bayesian approach to smoothing with penalized splines [@Kimeldorf1970-cn; @Wahba1983-mi; @Silverman1985-kw; @Wahba1985-bw; see @Miller2019-nf for a summary], it can be shown that GAMs fitted by *mgcv* are an empirical Bayesian model with a (usually improper) multivariate normal prior on the basis function coefficients.
 Samples from the posterior distribution of these models can be used to estimate the uncertainty in quantities derived from a GAM.
@@
 *gratia* provides functions for sampling from the posterior distribution of estimated smooths and from the model as a whole, where sampling can include the uncertainty in the estimated coefficients (`fitted_samples()`), the sampling uncertainty of the response (`predicted_samples()`), or both (`posterior_samples()`).
-By default, a Gaussian approximation to the posterior distribution is used, but a simple Metropolis Hasting sampler can be substituted (using `mgcv::gam.mh()`), which has better performance when the posterior is not well approximated by a Gaussian approximation.
+By default, a Gaussian approximation to the posterior distribution is used, but a simple Metropolis Hastings sampler can be substituted (using `mgcv::gam.mh()`), which has better performance when the posterior is not well approximated by a Gaussian approximation.
 
@@
 A particular feature of *mgcViz* is that it was designed to be scalable, easily handling models fitted to data with millions of observations, a use case not currently well handled by *gratia*.
-*tidymv* [@tidymv-2023] and its succesor, *tidygam* [@tidygam-2023], provide plots of estimated smooths and model predictions respectively, as well as differences of smooths, but the focus is on univariate smooths, in constrast to *gratia*'s ability to plot multivariate and specialist smooths (e.g.
+*tidymv* [@tidymv-2023] and its successor, *tidygam* [@tidygam-2023], provide plots of estimated smooths and model predictions respectively, as well as differences of smooths, but the focus is on univariate smooths, in contrast to *gratia*'s ability to plot multivariate and specialist smooths (e.g.
 soap film smooths).
@@
 Additional terms included in the linear predictor are a smooth of the day of year of sample collection (`jul.day`) and a smooth of ocean depth (`bath`).
-The response is assumed to be conditionally distributed Tweedie, with the power parameter ($p$) of the distribution estimated during fitting.
-Model coefficients and smoothing parameters are estimated using restricted maximum likelihood [@Wood2011-kn]
+The response is assumed to be conditionally Tweedie distributed, with the power parameter ($p$) of the distribution estimated during fitting.
+Model coefficients and smoothing parameters are estimated using restricted maximum likelihood [@Wood2011-kn].
+To use parallel processing in some parts of the model fitting algorithm, the `nthreads` control parameter is set to 10 (set this lower if your machine doesn't have this many physical CPU cores).
 ```{r fit-chl-model, cache = TRUE}
@@
 
-which show significant heteroscedasticity and departure from the condtional distribution of the response given the model (Figure \ref{fig:m1-appraise}).
+which show significant heteroscedasticity and departure from the conditional distribution of the response given the model, as seen in the increasing spread of the deviance residuals at larger values of the linear predictor in the upper right panel of Figure \ref{fig:m1-appraise}.
 
@@
 
-The problems with the model aparent in the diagnostics plots are probably due to important controls on chlorophyll *a* missing from the covariates available in the example data.
-However, the original model assumed constant values for the dispersion, $\phi$, and the power parameter $p$, which may be too inflexible given the absence of important effects in the model.
+The problems with the model apparent in the diagnostics plots are probably due to important controls on chlorophyll *a* missing from the covariates available in the example data.
+However, the original model assumed constant values for the dispersion, $\phi$, and the power parameter $p$, which may be too inflexible if missing covariates mean that important effects are not included in the model.
 
@@
 These plots allow you to read off the contributions of each smooth to the fitted response (on the link scale); they show link-scale *predictions* of the response for each smooth, conditional upon all other terms in the model, including any parametric effects and the intercept, having zero contribution.
-In the paralance of the *marginaleffects* package [@Arel-BundockGreiferHeiss:2024], these plots show *adjusted predictions*, just where the adjustment includes setting the contribution of *all* other model terms to the predicted value to zero.
+In the parlance of the *marginaleffects* package [@Arel-BundockGreiferHeiss:2024], these plots show *adjusted predictions*, just where the adjustment includes setting the contribution of *all* other model terms to the predicted value to zero.
 For *partial derivatives* (what *marginaleffects* would call a *marginal effect* or slope), *gratia* provides the `derivatives()`.
@@
 ```
-This returns a data frame of the requested values that is ammenable for plotted with `ggplot()`.
+This returns a data frame of the requested values that is amenable for plotted with `ggplot()`.
 
@@
 ```{r avg-chl-fv, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
-use <- c("(Intercpt)", "s(lat,lon)")
+use <- c("(Intercept)", "s(lat,lon)")
 fv <- fitted_values(m2, data = ds, terms = use) # predict
@@
 Finally, I summarise the predictions for the location parameter to yield the average of the predicted values
-```{r avg-chl-sumamrise, dependson = "avg-chl-fv", cache = TRUE}
+```{r avg-chl-summarise, dependson = "avg-chl-fv", cache = TRUE}
 fv |>
@@
 *gratia* provides a range of functionality to make working with estimated GAMs easier for users.
-It is designed to take some of the pain out of working with models, simplifying plotting of smooths and related features (differences, derivatives) and exposing the powerful machinary of the *mgcv* package without the need for a deep understanding of GAMs, splines, and the inner structure of *mgcv*'s model objects.
+It is designed to take some of the pain out of working with models, simplifying plotting of smooths and related features (differences, derivatives) and exposing the powerful machinery of the *mgcv* package without the need for a deep understanding of GAMs, splines, and the inner structure of *mgcv*'s model objects.
 As such, it provides a useful addition for anyone wanting to use GAMs in their data analyses without requiring them to be a GAM expert.

### 2024-11-21 paper.Rmd
implement the last of @dill's suggestions; labelling the BFs in figure 1

@@
 ```{r penalised-spline-basis-and-penalty, echo = FALSE, fig.height = 4, fig.width = 10, fig.cap = "\\label{fig:basis-funs}Basis functions (a) and associated penalty matrix (b) for a penalised, low rank, thin plate regression spline.
-a) shows the individual basis functions (thin coloured lines) multiplied by their respective model coefficients, as well as the data (black points) to which the GAM was fitted.
+a) shows the individual basis functions (thin coloured lines; numbers indicate which basis function each line represents) multiplied by their respective model coefficients, as well as the data (black points) to which the GAM was fitted.
 The estimated smooth is shown as the thick grey line.
@@
 Note the 9th basis function (labelled 'F9', which is the linear function at the lower left to upper right in a), is not affected by the penalty as it has 0 second derivative everywhere, and hence the resulting penalty for this function is 0.", fig.pos = "t!"}
-pkgs <- c("mgcv", "gratia", "ggplot2", "dplyr", "tibble", "patchwork")
+pkgs <- c("mgcv", "gratia", "ggplot2", "geomtextpath", "dplyr", "tibble", "patchwork")
 res <- vapply(pkgs, library, logical(1), character.only = TRUE,
@@
   ggplot(aes(x = x, y = .value)) +
-  geom_line(aes(colour = .bf)) +
+  geom_textline(
+    aes(x = x, y = .value, colour = .bf, label = .bf),
+    hjust = 0.9,
+    size = 5
+  ) +
   geom_point(data = sim_d, aes(x = x, y = y)) +

### 2024-11-21 paper.Rmd
fix the labels on figure 1a

@@
 
-```{r penalised-spline-basis-and-penalty, echo = FALSE, fig.height = 4, fig.width = 10, fig.cap = "\\label{fig:basis-funs}Basis functions (a) and associated penalty matrix (b) for a penalised, low rank, thin plate regression spline.
+```{r penalised-spline-basis-and-penalty, echo = FALSE, fig.height = 5, fig.width = 10, fig.cap = "\\label{fig:basis-funs}Basis functions (a) and associated penalty matrix (b) for a penalised, low rank, thin plate regression spline.
 a) shows the individual basis functions (thin coloured lines; numbers indicate which basis function each line represents) multiplied by their respective model coefficients, as well as the data (black points) to which the GAM was fitted.
@@
   ggplot(aes(x = x, y = .value)) +
-  geom_textline(
-    aes(x = x, y = .value, colour = .bf, label = .bf),
-    hjust = 0.9,
-    size = 5
-  ) +
   geom_point(data = sim_d, aes(x = x, y = y)) +
@@
     linewidth = 1.25, alpha = 0.5) +
+  geom_labelline(
+    aes(x = x, y = .value, colour = .bf, label = .bf),
+    hjust = "auto",
+    label.padding = unit(0.15, "lines"),
+    size = 3
+  ) +
   guides(colour = "none") +

### 2024-11-21 paper.Rmd
tweak summary

@@
 
-Generalized additive models [GAMs, @Hastie1990-bx; @Wood2017-qi] are an extension of the generalized linear model that allows the effects of covariates to be modelled as smooth functions.
+Generalized additive models [GAMs, @Hastie1990-bx; @Wood2017-qi] are an extension of generalized linear models that allows the effects of covariates to be modelled as smooth functions.
 GAMs are increasingly used in many areas of science [e.g.

### 2024-11-21 paper.Rmd
tweak a code block

@@
     chl ~ s(lat, lon, bs = "sos", m = -1, k = 150) + # location
-      s(jul.day, bs = "cr", k = 20) +
-      s(bath, k = 10),
+      s(jul.day, bs = "cr", k = 20) + s(bath, k = 10),
     ~ s(lat, lon, bs = "sos", m = -1, k = 100) +     # power
-      s(jul.day, bs = "cr", k = 20) +
-      s(bath, k = 10),
+      s(jul.day, bs = "cr", k = 20) + s(bath, k = 10),
     ~ s(lat, lon, bs = "sos", m = -1, k = 100) +     # scale
-      s(jul.day, bs = "cr", k = 20) +
-      s(bath, k = 10)),
+      s(jul.day, bs = "cr", k = 20) + s(bath, k = 10)
+  ),
   data = chl, method = "REML", control = ctrl, family = twlss()

### 2024-11-21 paper.Rmd
few tweaks to fix typos; add a missing reference

@@
 Several R packages provide similar functionality to that of *gratia*.
-Notably, the *mgcViz* package [@Fasiolo2020-nz] provides sophisticated capabilities for plotting estimated smooths for models fitted by *mgcv* and *qgam* [@], and model diagnostics plots.
+Notably, the *mgcViz* package [@Fasiolo2020-nz] provides sophisticated capabilities for plotting estimated smooths for models fitted by *mgcv* and *qgam* [@Fasiolo2020-pr], and model diagnostics plots.
 A particular feature of *mgcViz* is that it was designed to be scalable, easily handling models fitted to data with millions of observations, a use case not currently well handled by *gratia*.
@@
 
-Areas where *gratia* stands out from these other packages are i) the consistent functionality for range of posterior simulations (*mgcViz* has some support for simulating new response values) as illustrated below, ii) support for computing derivatives of smooths, partial derivatives, and derivatives on the response scale (i.e.
+Areas where *gratia* stands out from these other packages are i) the consistent functionality for a range of posterior simulations (*mgcViz* has some support for simulating new response values) as illustrated below, ii) support for computing derivatives of smooths, partial derivatives, and derivatives on the response scale (i.e.
 conditional derivatives), iii) tools for learning or teaching how GAMs work, iv) the range of utility functions that make working with GAMs easier, and v) the simple and consistent *tidy* interface to all functionality.
@@
 In the parlance of the *marginaleffects* package [@Arel-BundockGreiferHeiss:2024], these plots show *adjusted predictions*, just where the adjustment includes setting the contribution of *all* other model terms to the predicted value to zero.
-For *partial derivatives* (what *marginaleffects* would call a *marginal effect* or slope), *gratia* provides the `derivatives()`.
+For *partial derivatives* (what *marginaleffects* would call a *marginal effect* or slope), *gratia* provides `derivatives()`.
 

### 2024-12-20 paper.Rmd
fix a couple of typos

@@
 affiliations:
- - name: Department of Animal and Veterinary Science, Aarhus University, Denmark
+ - name: Department of Animal and Veterinary Sciences, Aarhus University, Denmark
    index: 1
@@
 # Summary
+
+```{r, setup, echo = FALSE, cache = FALSE, results = "hide"}
+pkgs <- c("mgcv", "gratia", "ggplot2", "geomtextpath", "dplyr", "tibble", "patchwork")
+res <- vapply(pkgs, library, logical(1), character.only = TRUE,
+  logical.return = TRUE)
+```
 
@@
 Note the 9th basis function (labelled 'F9', which is the linear function at the lower left to upper right in a), is not affected by the penalty as it has 0 second derivative everywhere, and hence the resulting penalty for this function is 0.", fig.pos = "t!"}
-pkgs <- c("mgcv", "gratia", "ggplot2", "geomtextpath", "dplyr", "tibble", "patchwork")
-res <- vapply(pkgs, library, logical(1), character.only = TRUE,
-  logical.return = TRUE)
 # simulate data
@@
 *gratia* provides a range of functionality to make working with estimated GAMs easier for users.
-It is designed to take some of the pain out of working with models, simplifying plotting of smooths and related features (differences, derivatives) and exposing the powerful machinery of the *mgcv* package without the need for a deep understanding of GAMs, splines, and the inner structure of *mgcv*'s model objects.
+It is designed to take some of the pain out of working with models, simplifying plotting of smooths and related features (differences, derivatives) and exposes the powerful machinery of the *mgcv* package without the need for a deep understanding of GAMs, splines, and the inner structure of *mgcv*'s model objects.
 As such, it provides a useful addition for anyone wanting to use GAMs in their data analyses without requiring them to be a GAM expert.

### 2024-12-21 paper.Rmd
fix two references

@@
 ```{r, setup, echo = FALSE, cache = FALSE, results = "hide"}
-pkgs <- c("mgcv", "gratia", "ggplot2", "geomtextpath", "dplyr", "tibble", "patchwork")
+pkgs <- c(
+  "mgcv", "gratia", "ggplot2", "geomtextpath", "dplyr", "tibble", "patchwork"
+)
 res <- vapply(pkgs, library, logical(1), character.only = TRUE,
@@
 Model coefficients and smoothing parameters are estimated using restricted maximum likelihood [@Wood2011-kn].
-To use parallel processing in some parts of the model fitting algorithm, the `nthreads` control parameter is set to 10 (set this lower if your machine doesn't have this many physical CPU cores).
+To use parallel processing in some parts of the model fitting algorithm, the `nthreads` control parameter is set to 8 (set this lower if your machine doesn't have this many physical CPU cores).
 ```{r fit-chl-model, cache = TRUE}
-ctrl <- gam.control(nthreads = 10)
+ctrl <- gam.control(nthreads = 8)
 m1 <- gam(

### 2024-12-23 README.Rmd
add JOSS badge

@@
 [![CRAN Downloads](https://cranlogs.r-pkg.org/badges/grand-total/gratia)](https://cran.r-project.org/package=gratia)
+[![DOI](https://joss.theoj.org/papers/10.21105/joss.06962/status.svg)](https://doi.org/10.21105/joss.06962)
 <!-- badges: end -->

### 2025-06-21 README.Rmd
add r-universe badges

--- 2024-08-05 README.Rmd
@@
 [![CRAN Downloads](https://cranlogs.r-pkg.org/badges/grand-total/gratia)](https://cran.r-project.org/package=gratia)
+[![DOI](https://joss.theoj.org/papers/10.21105/joss.06962/status.svg)](https://doi.org/10.21105/joss.06962)
+
+[![r-universe version](https://r-lib.r-universe.dev/gratia/badges/version)](https://r-lib.r-universe.dev/gratia)
+[![r-universe status](https://r-lib.r-universe.dev/gratia/badges/checks)](https://r-lib.r-universe.dev/gratia)
 <!-- badges: end -->

### 2025-06-21 README.Rmd
wrong links on r-universe badges

@@
 
-[![r-universe version](https://r-lib.r-universe.dev/gratia/badges/version)](https://r-lib.r-universe.dev/gratia)
-[![r-universe status](https://r-lib.r-universe.dev/gratia/badges/checks)](https://r-lib.r-universe.dev/gratia)
+[![r-universe version](https://gavinsimpson.r-universe.dev/gratia/badges/version)](https://r-lib.r-universe.dev/gratia)
+[![r-universe status](https://gavinsimpson.r-universe.dev/gratia/badges/checks)](https://r-lib.r-universe.dev/gratia)
 <!-- badges: end -->

### 2025-06-21 README.Rmd
wrong links on r-universe badges

--- 2025-06-21 README.Rmd
@@
 
-[![r-universe version](https://r-lib.r-universe.dev/gratia/badges/version)](https://r-lib.r-universe.dev/gratia)
-[![r-universe status](https://r-lib.r-universe.dev/gratia/badges/checks)](https://r-lib.r-universe.dev/gratia)
+[![r-universe version](https://gavinsimpson.r-universe.dev/gratia/badges/version)](https://gavinsimpson.r-universe.dev/gratia)
+[![r-universe status](https://gavinsimpson.r-universe.dev/gratia/badges/checks)](https://gavinsimpson.r-universe.dev/gratia)
 <!-- badges: end -->

### 2025-06-21 README.Rmd
pull all badges together

--- 2025-06-21 README.Rmd
@@
 [![DOI](https://joss.theoj.org/papers/10.21105/joss.06962/status.svg)](https://doi.org/10.21105/joss.06962)
-
-[![r-universe version](https://r-lib.r-universe.dev/gratia/badges/version)](https://r-lib.r-universe.dev/gratia)
-[![r-universe status](https://r-lib.r-universe.dev/gratia/badges/checks)](https://r-lib.r-universe.dev/gratia)
+[![r-universe version](https://gavinsimpson.r-universe.dev/gratia/badges/version)](https://gavinsimpson.r-universe.dev/gratia)
+[![r-universe status](https://gavinsimpson.r-universe.dev/gratia/badges/checks)](https://gavinsimpson.r-universe.dev/gratia)
 <!-- badges: end -->

### 2025-08-07 README.Rmd
add a section on citing gratia
clean ups for release

@@
   fig.path = "man/figures/README-",
-  out.width = "100%"
+  out.width = "100%",
+  dev = "ragg_png",
+  dpi = 96
 )
@@
 
+## Citation
+
+If you use *gratia* in your work, please cite the package using:
+
+Simpson, G. L., (2024). gratia: An R package for exploring generalized additive
+models. *Journal of Open Source Software*, **9**(104), 6962,
+<https://doi.org/10.21105/joss.06962>
+
 ## Features
 
-```{r draw-gam-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide", message = FALSE}
+```{r draw-gam-figure, fig.height = 6, fig.width = 9, echo = FALSE, fig.show = "hide", message = FALSE}
 df1 <- data_sim("eg1", n = 400, seed = 42)
 m1 <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df1, method = "REML")
-draw(m1, residuals = TRUE)
+draw(m1, residuals = TRUE, rug = FALSE)
 ```
 
-```{r draw-gam-figure-2d, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide", message = FALSE}
+```{r draw-gam-figure-2d, fig.height = 6, fig.width = 9, echo = FALSE, fig.show = "hide", message = FALSE}
 df2 <- data_sim("eg2", n = 1000, seed = 42)
@@
 
-```{r appraise-figure, dev = "png", fig.height = 6, fig.width = 9, dpi = 120, echo = FALSE, fig.show = "hide"}
+```{r appraise-figure, fig.height = 6, fig.width = 9, echo = FALSE, fig.show = "hide"}
 appraise(m1)

### 2025-08-07 tests/testthat/test-draw-methods.R

@@ -114,11 +114,9 @@ test_that("draw.gam() plots a simple multi-smooth AM", {
 })
 
 test_that("draw.gam() can draw partial residuals", {
-  plt1 <- draw(m_tiny_eg1, residuals = TRUE, rug = FALSE)
   plt2 <- draw(m_tiny_eg1, residuals = TRUE, scales = "fixed", rug = FALSE)
 
   # skip_on_ci() # testing without as moved to mac os x
-  expect_doppelganger("draw simple partial residuals", plt1)
   expect_doppelganger("draw simple partial residuals with fixed scales", plt2)
 })
 

### 2025-08-07 README.Rmd
spell check

@@
 
-The English word "grace" derives from the Latin *gratia*, meaning "favor, charm,
+The English word "grace" derives from the Latin *gratia*, meaning "favour, charm,
 thanks" ([according to Merriam Webster](https://www.merriam-webster.com/dictionary/grace)).
 
-The chair that Grace Wabha currently holds is named after 
+The chair that Grace Wahba currently holds is named after 
 [Isaac J Schoenberg](https://en.wikipedia.org/wiki/Isaac_Jacob_Schoenberg), a

### 2025-08-07 R/add-methods.R

@@ -511,7 +511,7 @@
 #'   `"derivatives"`.
 #' @param type character; `"change"` adds a single variable to `object`
 #'   indicating where the credible interval on the derivative excludes 0.
-#'   `"sizer"` adds two variables indicating whether the derivative is postive
+#'   `"sizer"` adds two variables indicating whether the derivative is positive
 #'   or negative.
 #' @param ... arguments passed to other methods
 #'
@@ -646,7 +646,7 @@
 #' @param n integer; the number of posterior draws to add.
 #' @param seed numeric; a value to seed the random number generator.
 #' @param select character; select which smooth's posterior to draw from. The
-#'   default, `NULL`, means the posteriors of all smooths in model wil be
+#'   default, `NULL`, means the posteriors of all smooths in model will be
 #'   sampled from individually. If supplied, a character vector of requested
 #'   smooth terms.
 #' @param ... arguments are passed to the posterior draw function, currently one

### 2025-08-07 R/basis-funs.R

@@ -28,7 +28,7 @@
 #'   the smooth basis. See argument `absorb.cons` in [mgcv::smoothCon()].
 #' @param at a data frame containing values of the smooth covariate(s) at which
 #'   the basis should be evaluated.
-#' @param diagonalize logical; if `TRUE`, reparameterizes the smooth such that
+#' @param diagonalize logical; if `TRUE`, reparameterises the smooth such that
 #'   the associated penalty is an identity matrix. This has the effect of
 #'   turning the last diagonal elements of the penalty to zero, which highlights
 #'   the penalty null space.

### 2025-08-07 R/by-variable-utilities.R

@@ -1,7 +1,7 @@
 #' Tests for by variable smooths
 #'
 #' Functions to check if a smooth is a by-variable one and to test of the type
-#' of by-variable smooth is a factor-smooth or a continous-smooth interaction.
+#' of by-variable smooth is a factor-smooth or a continuous-smooth interaction.
 #'
 #' @param smooth an object of class `"mgcv.smooth"`
 #'

### 2025-08-07 R/concurvity.R

@@ -5,7 +5,7 @@
 #' @param terms currently ignored
 #' @param type character;
 #' @param pairwise logical; extract pairwise concurvity of model terms?
-#' @param ... arguents passed to other methods.
+#' @param ... arguments passed to other methods.
 #'
 #' @importFrom tibble as_tibble add_column rownames_to_column
 #' @importFrom mgcv concurvity

### 2025-08-07 R/confint-methods.R

@@ -203,7 +203,7 @@
 #' @param level numeric, `0 < level < 1`; the confidence level of the point-wise
 #'   or simultaneous interval. The default is `0.95` for a 95% interval.
 #' @param data data frame; new values of the covariates used in the model fit.
-#'   The selected smooth(s) wil be evaluated at the supplied values.
+#'   The selected smooth(s) will be evaluated at the supplied values.
 #' @param n numeric; the number of points to evaluate smooths at.
 #' @param type character; the type of interval to compute. One of `"confidence"`
 #'   for point-wise intervals, or `"simultaneous"` for simultaneous intervals.
@@ -226,7 +226,7 @@
 #'   an exact match? Can only be used if `length(parm)` is `1`.
 #' @param ... additional arguments for methods
 #' @param newdata DEPRECATED! data frame; containing new values of the
-#'   covariates used in the model fit. The selected smooth(s) wil be evaluated
+#'   covariates used in the model fit. The selected smooth(s) will be evaluated
 #'   at the supplied values.
 #'
 #' @return a tibble with components:

### 2025-08-07 R/data-sim.R

@@ -18,16 +18,16 @@
 #' (1 - x))^6 + 10 * (10 * x)^3 * (1 - x)^10}.
 #'   * `"eg4"`: a factor by smooth true model. The true model contains a factor
 #'     with 3 levels, where the response for the *n*th level follows the *n*th
-#'     Gu & Wabha function (for \eqn{n \in {1, 2, 3}}{n in {1,2,3}}).
+#'     Gu & Wahba function (for \eqn{n \in {1, 2, 3}}{n in {1,2,3}}).
 #'   * `"eg5"`: an additive plus factor true model. The response is a linear
-#'     combination of the Gu & Wabha functions 2, 3, 4 (the latter is a null
+#'     combination of the Gu & Wahba functions 2, 3, 4 (the latter is a null
 #'     function) plus a factor term with four levels.
 #'   * `"eg6"`: an additive plus random effect term true model.
 #'   * ´"eg7"`: a version of the model in `"eg1"`, but where the covariates are
 #'     correlated.
-#'   * `"gwf2"`: a model where the response is Gu & Wabha's
+#'   * `"gwf2"`: a model where the response is Gu & Wahba's
 #'     \eqn{f_2(x_i)}{f_2(x_i)} plus noise.
-#'   * `"lwf6"`: a model where the response is Luo & Wabha's "example 6"
+#'   * `"lwf6"`: a model where the response is Luo & Wahba's "example 6"
 #'     function \eqn{sin(2(4x-2)) + 2 exp(-256(x-0.5)^2)}{
 #'     sin(2 * ((4 * x) - 2)) + (2 * exp(-256 * (x - .5)^2))} plus noise.
 #'   * `"gfam"`: simulates data for use with GAMs with
@@ -152,8 +152,8 @@
     eg5 = additive_plus_factor_model,
     eg6 = four_term_plus_ranef_model,
     eg7 = correlated_four_term_additive_model,
-    gwf2 = gu_wabha_f2,
-    lwf6 = luo_wabha_f6,
+    gwf2 = gu_Wahba_f2,
+    lwf6 = luo_Wahba_f6,
     gfam = sim_gfam
   )
 
@@ -261,8 +261,8 @@
   x
 }
 
-## Gu Wabha functions
-#' Gu and Wabha test functions
+## Gu Wahba functions
+#' Gu and Wahba test functions
 #'
[…]

### 2025-08-07 R/datasets.R

@@ -69,8 +69,8 @@ NULL
 #'   generalized additive models: an introduction with mgcv. *PeerJ Preprints*
 #'   **6**:e27320v1 \doi{10.7287/peerj.preprints.27320v1}.
 #'
-#' @references Lathrop RC. (2000). Madison Wisonsin Lakes Zooplankton 1976--1994.
-#'   Environmental Data Initiative.
+#' @references Lathrop RC. (2000). Madison Wisconsin Lakes Zooplankton
+#'   1976--1994. Environmental Data Initiative.
 #'
 #' @format A data frame
 #' @name zooplankton

### 2025-08-07 R/derivatives.R

@@ -30,7 +30,7 @@
 
 #' @param select character; select which smooth's posterior to draw from.
 #'   The default (`NULL`) means the posteriors of all smooths in `model`
-#'   wil be sampled from. If supplied, a character vector of requested terms.
+#'   will be sampled from. If supplied, a character vector of requested terms.
 #'   Can be a partial match to a smooth term; see argument `partial_match`
 #'   below.
 #' @param term `r lifecycle::badge("deprecated")` Use `select` instead.

### 2025-08-07 R/diagnose.R

@@ -10,7 +10,7 @@
 #'   [stats::lm] are supported.
 #' @param method character; method used to generate theoretical quantiles.
 #'   The default is `"uniform"`, which generates reference quantiles using
-#'   random draws from a uniform distribution and the inverse cummulative
+#'   random draws from a uniform distribution and the inverse cumulative
 #'   distribution function (CDF) of the fitted values. The reference quantiles
 #'   are averaged over `n_uniform` draws. `"simulate"` generates reference
 #'   quantiles by simulating new response data from the model at the observed
@@ -52,7 +52,7 @@
 #' @param point_col colour of points on the QQ plot.
 #' @param point_alpha alpha transparency of points on the QQ plot.
 #' @param line_col colour used to draw the reference line.
-#' @param ... arguments passed ot other methods.
+#' @param ... arguments passed to other methods.
 #'
 #' @note The wording used in [mgcv::qq.gam()] uses *direct* in reference to the
 #'   simulated residuals method (`method = "simulated"`). To avoid confusion,
@@ -66,7 +66,7 @@
 #' is described in Augustin *et al* (2012):
 #'
 #' Augustin, N.H., Sauleau, E.-A., Wood, S.N., (2012) On quantile quantile plots
-#' for generalized linear models. *Computational Statatistics and Data Analysis*
+#' for generalized linear models. *Computational Statistics and Data Analysis*
 #' **56**, 2404-2409 \doi{doi:10.1016/j.csda.2012.01.026}.
 #'
 #'
@@ -772,7 +772,7 @@
 #'   [stats::lm] are supported.
 #' @param method character; method used to generate theoretical quantiles.
 #'   The default is `"uniform"`, which generates reference quantiles using
-#'   random draws from a uniform distribution and the inverse cummulative
+#'   random draws from a uniform distribution and the inverse cumulative
 #'   distribution function (CDF) of the fitted values. The reference quantiles
 #'   are averaged over `n_uniform` draws. `"simulate"` generates reference
 #'   quantiles by simulating new response data from the model at the observed

### 2025-08-07 R/edf.R

@@ -12,11 +12,11 @@
 #'   \deqn{\mathrm{EDF} = 2\mathrm{tr}(\mathbf{F}) -
 #'   \mathrm{tr}(\mathbf{FF}),}{EDF = 2 * tr(F) - tr(F),} where
 #'   \eqn{\mathrm{tr}} is the matrix trace and \eqn{\mathbf{F}}{F} is a matrix
-#'   mapping un-penalized coefficient estimates to the penalized coefficient
+#'   mapping unpenalized coefficient estimates to the penalized coefficient
 #'   estimates.  The trace of \eqn{\mathbf{F}}{F} is effectively the average
-#'   shrinkage of the coefficients multipled by the number of coefficients
+#'   shrinkage of the coefficients multiplied by the number of coefficients
 #'   (Wood, 2017). Smooth-specific EDFs then are obtained by summing up the
-#'   relevent elements of \eqn{\mathrm{diag}(2\mathbf{F} - \mathbf{FF})}.
+#'   relevant elements of \eqn{\mathrm{diag}(2\mathbf{F} - \mathbf{FF})}.
 #'
 #' @param object a fitted model from which to extract smooth-specific EDFs.
 #' @param smooth `r lifecycle::badge("deprecated")` Use `select` instead.

### 2025-08-07 R/evenly.R

@@ -84,7 +84,7 @@
 #'   forwards or backwards by an amount related to `eps`. This is particularly
 #'   useful in computing derivatives via finite differences where without this
 #'   adjustment we may be predicting for values outside the range of the data
-#'   and hence the conmstraints of the penalty.
+#'   and hence the constraints of the penalty.
 #'
 #' @param x numeric; vector over which evenly-spaced values are returned
 #' @param n numeric; the number of evenly-spaced values to return

### 2025-08-07 R/fderiv.R

@@ -5,7 +5,7 @@
 #'
 #' This function was deprecated because it was limited to first order forward
 #' finite differences for derivatives only, but couldn't be improved to offer
-#' the needed functionality without breaking backwards compatability with papers
+#' the needed functionality without breaking backwards compatibility with papers
 #' and blog posts that already used `fderiv()`. A replacement, [derivatives()],
 #' is now available and recommended for new analyses.
 #'

### 2025-08-07 R/posterior-samples.R

@@ -820,7 +820,7 @@
 #'   `data` not supplied
 #' @param select character; select which smooth's posterior to draw from.
 #'   The default (`NULL`) means the posteriors of all smooths in `model`
-#'   wil be sampled from. If supplied, a character vector of requested terms.
+#'   will be sampled from. If supplied, a character vector of requested terms.
 #' @param term `r lifecycle::badge("deprecated")` Use `select` instead.
 #' @param rng_per_smooth logical; if TRUE, the behaviour of gratia version
 #' 0.8.1 or earlier is used, whereby a separate call the the random number

### 2025-08-07 R/residuals.R

@@ -35,7 +35,7 @@
 #' ## load mgcv
 #' load_mgcv()
 #'
-#' ## example data - Gu & Wabha four term model
+#' ## example data - Gu & Wahba four term model
 #' df <- data_sim("eg1", n = 400, seed = 42)
 #' ## fit the model
 #' m <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, method = "REML")

### 2025-08-07 R/samplers.R

@@ -10,7 +10,7 @@
 #' @param method character; which algorithm to use to sample from the posterior.
 #'   Currently implemented methods are: `"gaussian"` and `"mh"`. `"gaussian"`
 #'   calls `gaussian_draws()` which uses a Gaussian approximation to the
-#'   posterior distribution. `"mh"` uses a simple Metropolis Hasting sampler
+#'   posterior distribution. `"mh"` uses a simple Metropolis Hastings sampler
 #'   which alternates static proposals based on a Gaussian approximation to the
 #'   posterior, with random walk proposals. Note, setting `t_df` to a low value
 #'   will result in heavier-tailed statistic proposals. See `mgcv::gam.mh()`

### 2025-08-07 R/simulate-methods.R

@@ -6,7 +6,7 @@
 #' distribution used when fitting the model.
 #'
 #' For `simulate.gam()` to function, the `family` component of the fitted
-#' model must contain, or be updateable to contain, the required random
+#' model must contain, or be updatable to contain, the required random
 #' number generator. See [mgcv::fix.family.rd()].
 #'
 #' @param object a fitted GAM, typically the result of a call to [mgcv::gam]`

### 2025-08-07 R/smooth-estimates.R

@@ -9,7 +9,7 @@
 #' @param object an object of class `"gam"` or `"gamm"`.
 #' @param select character; select which smooth's posterior to draw from.
 #'   The default (`NULL`) means the posteriors of all smooths in `model`
-#'   wil be sampled from. If supplied, a character vector of requested terms.
+#'   will be sampled from. If supplied, a character vector of requested terms.
 #' @param smooth `r lifecycle::badge("deprecated")` Use `select` instead.
 #' @param n numeric; the number of points over the range of the covariate at
 #'   which to evaluate the smooth.
@@ -486,7 +486,7 @@
 #'
 #' @param smooth currently an object that inherits from class `mgcv.smooth`.
 #' @param model a fitted model; currently only [mgcv::gam()] and [mgcv::bam()]
-#'   models are suported.
+#'   models are supported.
 #' @param data an optional data frame of values to evaluate `smooth` at.
 #' @param ... arguments passed to other methods
 #'

### 2025-08-07 R/utililties.R

@@ -323,7 +323,7 @@ stop_if_not_mgcv_smooth <- function(smooth) {
 #'
 #' @param object a fitted GAM model object.
 #' @param term character; the name of a smooth term to extract.
-#' @param level character; which level of the factor to exrtact the smooth
+#' @param level character; which level of the factor to extract the smooth
 #'   for.
 #'
 #' @return A single smooth object, or a list of smooths if several match the
@@ -774,7 +774,7 @@ stop_if_not_mgcv_smooth <- function(smooth) {
 #' @param h numeric; the amount to shift values in `df` by.
 #' @param i logical; a vector indexing columns of `df` that should not be
 #'   included in the shift.
-#' @param FUN function; a function to applut the shift. Typically `+` or `-`.
+#' @param FUN function; a function to apply the shift. Typically `+` or `-`.
 #' @param focal character; the focal variable when computing partial
 #'   derivatives. This allows shifting only the focal variable by `eps`.
 `shift_values` <- function(df, h, i, FUN = `+`, focal = NULL) {

### 2025-11-07 tests/testthat/test-derivatives.R
add tests for #362

@@ -533,3 +533,14 @@ test_that("derivatives works with models that include random effects", {
   expect_s3_class(d, "tbl_df")
   expect_named(d, c(deriv_nms, paste0("x", 0:2)))
 })
+
+test_that("derivatives works for sz smooths", {
+  expect_silent(d <- derivatives(m_sz))
+  expect_s3_class(d, "derivatives")
+  expect_s3_class(d, "tbl_df")
+  expect_named(d, c(deriv_nms, c("x2", "fac", "x0")))
+
+  skip_on_cran()
+  skip_on_ci()
+  expect_snapshot(print(d))
+})
